\documentclass[a4paper,oneside]{memoir}
\usepackage{dramatist}
\usepackage{dramatist-mods}
\sloppybottom
\raggedbottom
\raggedright
\setlrmarginsandblock{90pt}{90pt}{*}

\begin{document}

\Character[Subtle, the Alchemist]{Subtle}{subtle}
\Character[Face, the housekeeper]{Face}{face}
\Character[Dol Common, their colleague]{Dol}{dol}
\Character[Dapper, a lawyer's clerk]{Dapper}{dapper}
\Character[Drugger, a tobacco man]{Drugger}{drugger}
\Character[Lovewit, master of the house]{Lovewit}{lovewit}
\Character[Sir Epicure Mammon, a knight]{Mammon}{mammon}
\Character[Pertinax Surly, a gamester]{Surly}{surly}
\Character[Tribulation Wholesome, a pastor of Amsterdam]{Tribulation}{tribulation}
\Character[Ananias, a deacon there]{Ananias}{ananias}
\Character[Kastril, the angry boy]{Kastril}{kastril}
\Character[Dame Pliant, his sister, a widow]{Pliant}{pliant}
\Character[A Neighbour]{Neigh. 1}{neighone}
\Character[Another Neighbour]{Neigh. 2}{neightwo}
\Character[Another Neighbour]{Neigh. 3}{neighthree}
\Character[Another Neighbour]{Neigh. 4}{neighfour}
\Character[Another Neighbour]{Neigh. 5}{neighfive}
\Character[Another Neighbour]{Neigh. 6}{neighsix}
\Character[An officer]{Officer 1}{officerone}
\Character[Another officer]{Officer 2}{officertwo}

\settowidth{\speakswidth}{\speaksfont{Tribulation}\speaksdel\hspace{\Dlabelsep}}
\setlength\speaksindent{0in}

\DramPer


\begin{drama*}


\act

\scene

\stagedir{A ROOM IN LOVEWIT'S HOUSE. ENTER FACE, IN A CAPTAIN'S UNIFORM, WITH HIS SWORD DRAWN, AND SUBTLE WITH A VIAL, QUARRELLING, AND FOLLOWED BY DOL COMMON.}
\facespeaks Believe 't, I will.\\
\subtlespeaks \vinphantom{Believe 't, I will.} Thy worst. I fart at thee.\\
\dolspeaks Have you your wits? why, gentlemen! for love---\\
\facespeaks Sirrah, I'll strip you---\\
\subtlespeaks \vinphantom{Sirrah, I'll strip you---} What to do? lick figs\\
Out at my---\\
\facespeaks \vinphantom{Out at my---} Rogue, rogue!---out of all your sleights.\\
\dolspeaks Nay, look ye, sovereign, general, are you madmen?\\
\subtlespeaks O, let the wild sheep loose. I'll gum your silks\\
With good strong water, an you come.\\
\dolspeaks \vinphantom{With good strong water, an you come.} Will you have\\
The neighbours hear you? will you betray all?\\
Hark! I hear somebody.\\
\facespeaks \vinphantom{Hark! I hear somebody.} Sirrah---\\
\subtlespeaks \vinphantom{Hark! I hear somebody. Sirrah---} I shall mar\\
All that the tailor has made, if you approach.\\
\facespeaks You most notorious whelp, you insolent slave,\\
Dare you do this?\\
\subtlespeaks \vinphantom{Dare you do this?} Yes, faith; yes, faith.\\
\facespeaks \vinphantom{Dare you do this? Yes, faith; yes, faith.} Why, who\\
Am I, my mungrel? who am I?\\
\subtlespeaks \vinphantom{Am I, my mungrel? who am I?} I'll tell you.,\\
Since you know not yourself.\\
\facespeaks \vinphantom{Since you know not yourself.} Speak lower, rogue.\\
\subtlespeaks Yes, you were once (time's not long past) the good,\\
Honest, plain, livery-three-pound-thrum, that kept\\
Your master's worship's house here in the Friars,\\
For the vacations---\\
\facespeaks \vinphantom{For the vacations---} Will you be so loud?\\
\subtlespeaks Since, by my means, translated suburb-captain.\\
\facespeaks By your means, doctor dog!\\
\subtlespeaks \vinphantom{By your means, doctor dog!} Within man's memory,\\
All this I speak of.\\
\facespeaks \vinphantom{All this I speak of.} Why, I pray you, have I\\
Been countenanced by you, or you by me?\\
Do but collect, sir, where I met you first.\\
\subtlespeaks I do not hear well.\\
\facespeaks \vinphantom{I do not hear well.} Not of this, I think it.\\
But I shall put you in mind, sir;---at Pie-corner,\\
Taking your meal of steam in, from cooks' stalls,\\
Where, like the father of hunger, you did walk\\
Piteously costive, with your pinch'd-horn-nose,\\
And your complexion of the Roman wash,\\
Stuck full of black and melancholic worms,\\
Like powder corns shot at the artillery-yard.\\
\subtlespeaks I wish you could advance your voice a little.\\
\facespeaks When you went pinn'd up in the several rags\\
You had raked and pick'd from dunghills, before day;\\
Your feet in mouldy slippers, for your kibes;\\
A felt of rug, and a thin threaden cloke,\\
That scarce would cover your no buttocks---\\
\subtlespeaks \vinphantom{That scarce would cover your no buttocks---} So, sir!\\
\facespeaks When all your alchemy, and your algebra,\\
Your minerals, vegetals, and animals,\\
Your conjuring, cozening, and your dozen of trades,\\
Could not relieve your corps with so much linen\\
Would make you tinder, but to see a fire;\\
I gave you countenance, credit for your coals,\\
Your stills, your glasses, your materials;\\
Built you a furnace, drew you customers,\\
Advanced all your black arts; lent you, beside,\\
A house to practise in---\\
\subtlespeaks \vinphantom{A house to practise in---} Your master's house!\\
\facespeaks Where you have studied the more thriving skill\\
Of bawdry since.\\
\subtlespeaks \vinphantom{Of bawdry since.} Yes, in your master's house.\\
You and the rats here kept possession.\\
Make it not strange. I know you were one could keep\\
The buttery-hatch still lock'd, and save the chippings,\\
Sell the dole beer to aqua-vitae men,\\
The which, together with your Christmas vails\\
At post-and-pair, your letting out of counters,\\
Made you a pretty stock, some twenty marks,\\
And gave you credit to converse with cobwebs,\\
Here, since your mistress' death hath broke up house.\\
\facespeaks You might talk softlier, rascal.\\
\subtlespeaks \vinphantom{You might talk softlier, rascal.} No, you scarab,\\
I'll thunder you in pieces: I will teach you\\
How to beware to tempt a Fury again,\\
That carries tempest in his hand and voice.\\
\facespeaks The place has made you valiant.\\
\subtlespeaks \vinphantom{The place has made you valiant.} No, your clothes.---\\
Thou vermin, have I ta'en thee out of dung,\\
So poor, so wretched, when no living thing\\
Would keep thee company, but a spider, or worse?\\
Rais'd thee from brooms, and dust, and watering-pots,\\
Sublimed thee, and exalted thee, and fix'd thee\\
In the third region, call'd our state of grace?\\
Wrought thee to spirit, to quintessence, with pains\\
Would twice have won me the philosopher's work?\\
Put thee in words and fashion, made thee fit\\
For more than ordinary fellowships?\\
Giv'n thee thy oaths, thy quarrelling dimensions,\\
Thy rules to cheat at horse-race, cock-pit, cards,\\
Dice, or whatever gallant tincture else?\\
Made thee a second in mine own great art?\\
And have I this for thanks! Do you rebel,\\
Do you fly out in the projection?\\
Would you be gone now?\\
\dolspeaks \vinphantom{Would you be gone now?} Gentlemen, what mean you?\\
Will you mar all?\\
\subtlespeaks \vinphantom{Will you mar all?} Slave, thou hadst had no name---\\
\dolspeaks Will you undo yourselves with civil war?\\
\subtlespeaks Never been known, past equi clibanum,\\
The heat of horse-dung, under ground, in cellars,\\
Or an ale-house darker than deaf John's; been lost\\
To all mankind, but laundresses and tapsters,\\
Had not I been.\\
\dolspeaks \vinphantom{Had not I been.} Do you know who hears you, sovereign?\\
\facespeaks Sirrah---\\
\dolspeaks \vinphantom{Sirrah---} Nay, general, I thought you were civil.\\
\facespeaks I shall turn desperate, if you grow thus loud.\\
\subtlespeaks And hang thyself, I care not.\\
\facespeaks \vinphantom{And hang thyself, I care not.} Hang thee, collier,\\
And all thy pots, and pans, in picture, I will,\\
Since thou hast moved me---\\
\dolspeaks \vinphantom{Since thou hast moved me---} O, this will o'erthrow all.\\
\facespeaks Write thee up bawd in Paul's, have all thy tricks\\
Of cozening with a hollow cole, dust, scrapings,\\
Searching for things lost, with a sieve and sheers,\\
Erecting figures in your rows of houses,\\
And taking in of shadows with a glass,\\
Told in red letters; and a face cut for thee,\\
Worse than Gamaliel Ratsey's.\\
\dolspeaks \vinphantom{Worse than Gamaliel Ratsey's.} Are you sound?\\
Have you your senses, masters?\\
\facespeaks \vinphantom{Have you your senses, masters?} I will have\\
A book, but barely reckoning thy impostures,\\
Shall prove a true philosopher's stone to printers.\\
\subtlespeaks Away, you trencher-rascal!\\
\facespeaks \vinphantom{Away, you trencher-rascal!} Out, you dog-leech!\\
The vomit of all prisons---\\
\dolspeaks \vinphantom{The vomit of all prisons---} Will you be\\
Your own destructions, gentlemen?\\
\facespeaks \vinphantom{Your own destructions, gentlemen?} Still spew'd out\\
For lying too heavy on the basket.\\
\subtlespeaks  \vinphantom{For lying too heavy on the basket.} Cheater!\\
\facespeaks Bawd!\\
\subtlespeaks \vinphantom{Bawd!} Cow-herd!\\
\facespeaks \vinphantom{Bawd! Cow-herd!} Conjurer!\\
\subtlespeaks \vinphantom{Bawd! Cow-herd! Conjurer!} Cut-purse!\\
\facespeaks \vinphantom{Bawd! Cow-herd! Conjurer! Cut-purse!} Witch!\\
\dolspeaks \vinphantom{Bawd! Cow-herd! Conjurer! Cut-purse! Witch!} O me!\\
We are ruin'd, lost! have you no more regard\\
To your reputations? where's your judgment? 'slight,\\
Have yet some care of me, of your republic---\\
\facespeaks Away, this brach! I'll bring thee, rogue, within\\
The statute of sorcery, tricesimo tertio\\
Of Harry the Eighth: ay, and perhaps thy neck\\
Within a noose, for laundring gold and barbing it.\\
\dolspeaks \direct*{SNATCHES FACE'S SWORD} You'll bring your head within a cockscomb, will you?\\
And you, sir, with your menstrue---\\
\direct*{DASHES SUBTLE'S VIAL OUT OF HIS HAND.}
\vinphantom{And you, sir, with your menstrue--} Gather it up.---\\
'Sdeath, you abominable pair of stinkards,\\
Leave off your barking, and grow one again,\\
Or, by the light that shines, I'll cut your throats.\\
I'll not be made a prey unto the marshal,\\
For ne'er a snarling dog-bolt of you both.\\
Have you together cozen'd all this while,\\
And all the world, and shall it now be said,\\
You've made most courteous shift to cozen yourselves?\\
\direct*{TO FACE.}
You will accuse him! you will "bring him in\\
Within the statute!" Who shall take your word?\\
A whoreson, upstart, apocryphal captain,\\
Whom not a Puritan in Blackfriars will trust\\
So much as for a feather:\\
\direct*{TO SUBTLE.}
\vinphantom{So much as for a feather:} and you, too,\\
Will give the cause, forsooth! you will insult,\\
And claim a primacy in the divisions!\\
You must be chief! as if you only had\\
The powder to project with, and the work\\
Were not begun out of equality?\\
The venture tripartite? all things in common?\\
Without priority? 'Sdeath! you perpetual curs,\\
Fall to your couples again, and cozen kindly,\\
And heartily, and lovingly, as you should,\\
And lose not the beginning of a term,\\
Or, by this hand, I shall grow factious too,\\
And take my part, and quit you.\\
\facespeaks \vinphantom{And take my part, and quit you.} 'Tis his fault;\\
He ever murmurs, and objects his pains,\\
And says, the weight of all lies upon him.\\
\subtlespeaks Why, so it does.\\
\dolspeaks \vinphantom{Why, so it does.} How does it? do not we\\
Sustain our parts?\\
\subtlespeaks \vinphantom{Sustain our parts?} Yes, but they are not equal.\\
\dolspeaks Why, if your part exceed to-day, I hope\\
Ours may, to-morrow match it.\\
\subtlespeaks \vinphantom{Ours may, to-morrow match it.} Ay, they MAY.\\
\dolspeaks May, murmuring mastiff! ay, and do. Death on me!\\
Help me to throttle him.\\
\direct*{SEIZES SUB. BY THE THROAT.}
\subtlespeaks \vinphantom{Help me to throttle him.} Dorothy! mistress Dorothy!\\
'Ods precious, I'll do any thing. What do you mean?\\
\dolspeaks Because o' your fermentation and cibation?\\
\subtlespeaks Not I, by heaven---\\
\dolspeaks \vinphantom{Not I, by heaven---} Your Sol and Luna\\
\direct*{TO FACE.}
\vinphantom{Not I, by heaven--- Your Sol and Luna} ---help me.\\
\subtlespeaks Would I were hang'd then? I'll conform myself.\\
\dolspeaks Will you, sir? do so then, and quickly: swear.\\
\subtlespeaks What should I swear?\\
\dolspeaks \vinphantom{What should I swear?} To leave your faction, sir,\\
And labour kindly in the common work.\\
\subtlespeaks Let me not breathe if I meant aught beside.\\
I only used those speeches as a spur\\
To him.\\
\dolspeaks \vinphantom{To him.} I hope we need no spurs, sir. Do we?\\
\facespeaks 'Slid, prove to-day, who shall shark best.\\
\subtlespeaks \vinphantom{'Slid, prove to-day, who shall shark best.} Agreed.\\
\dolspeaks Yes, and work close and friendly.\\
\subtlespeaks \vinphantom{Yes, and work close and friendly.} 'Slight, the knot\\
Shall grow the stronger for this breach, with me.\\
\direct*{THEY SHAKE HANDS.}
\dolspeaks Why, so, my good baboons! Shall we go make\\
A sort of sober, scurvy, precise neighbours,\\
That scarce have smiled twice since the king came in,\\
A feast of laughter at our follies? Rascals,\\
Would run themselves from breath, to see me ride,\\
Or you t' have but a hole to thrust your heads in,\\
For which you should pay ear-rent? No, agree.\\
And may don Provost ride a feasting long,\\
In his old velvet jerkin and stain'd scarfs,\\
My noble sovereign, and worthy general,\\
Ere we contribute a new crewel garter\\
To his most worsted worship.\\
\subtlespeaks \vinphantom{To his most worsted worship.} Royal Dol!\\
Spoken like Claridiana, and thyself.\\
\facespeaks For which at supper, thou shalt sit in triumph,\\
And not be styled Dol Common, but Dol Proper,\\
Dol Singular: the longest cut at night,\\
Shall draw thee for his Doll Particular.\\
\direct*{BELL RINGS WITHOUT.}
\subtlespeaks Who's that? one rings. To the window, Dol:\\
\direct*{EXIT DOL.}
\vinphantom{Who's that? one rings. To the window, Dol:} ---pray heaven,\\
The master do not trouble us this quarter.\\
\facespeaks  O, fear not him. While there dies one a week\\
O' the plague, he's safe, from thinking toward London.\\
Beside, he's busy at his hop-yards now;\\
I had a letter from him. If he do,\\
He'll send such word, for airing of the house,\\
As you shall have sufficient time to quit it:\\
Though we break up a fortnight, 'tis no matter.\\
\direct*{RE-ENTER DOL.}
\subtlespeaks Who is it, Dol?\\
\dolspeaks \vinphantom{Who is it, Dol?} A fine young quodling.\\
\facespeaks \vinphantom{Who is it, Dol? A fine young quodling.} O,\\
My lawyer's clerk, I lighted on last night,\\
In Holborn, at the Dagger. He would have\\
(I told you of him) a familiar,\\
To rifle with at horses, and win cups.\\
\dolspeaks O, let him in.\\
\subtlespeaks \vinphantom{O, let him in.} Stay. Who shall do't?\\
\facespeaks \vinphantom{O, let him in. Stay. Who shall do't?} Get you\\
Your robes on: I will meet him as going out.\\
\dolspeaks And what shall I do?\\
\facespeaks \vinphantom{And what shall I do?} Not be seen; away!\\
\direct*{EXIT DOL.}
Seem you very reserv'd.\\
\subtlespeaks \vinphantom{Seem you very reserv'd.} Enough.\\
\direct*{EXIT.}
\facespeaks \direct*{ALOUD AND RETIRING} \vinphantom{Seem you very reserv'd. Enough.} God be wi' you, sir,\\
I pray you let him know that I was here:\\
His name is Dapper. I would gladly have staid, but---\\
\dapperspeaks \direct*{WITHIN} Captain, I am here.\\
\facespeaks \vinphantom{Captain, I am here.} Who's that?---He's come, I think, doctor.\\
\direct*{ENTER DAPPER.}
Good faith, sir, I was going away.\\
\dapperspeaks \vinphantom{Good faith, sir, I was going away.} In truth\\
I am very sorry, captain.\\
\facespeaks \vinphantom{I am very sorry, captain.} But I thought\\
Sure I should meet you.\\
\dapperspeaks \vinphantom{Sure I should meet you.} Ay, I am very glad.\\
I had a scurvy writ or two to make,\\
And I had lent my watch last night to one\\
That dines to-day at the sheriff's, and so was robb'd\\
Of my past-time.\\
\direct*{RE-ENTER SUBTLE IN HIS VELVET CAP AND GOWN.}
\vinphantom{Of my past-time.} Is this the cunning-man?\\
\facespeaks This is his worship.\\
\dapperspeaks \vinphantom{This is his worship.} Is he a doctor?\\
\facespeaks \vinphantom{This is his worship. Is he a doctor?} Yes.\\
\dapperspeaks And have you broke with him, captain?\\
\facespeaks \vinphantom{And have you broke with him, captain?} Ay.\\
\dapperspeaks \vinphantom{And have you broke with him, captain? Ay.} And how?\\
\facespeaks Faith, he does make the matter, sir, so dainty\\
I know not what to say.\\
\dapperspeaks \vinphantom{I know not what to say.} Not so, good captain.\\
\facespeaks Would I were fairly rid of it, believe me.\\
\dapperspeaks Nay, now you grieve me, sir. Why should you wish so?\\
I dare assure you, I'll not be ungrateful.\\
\facespeaks I cannot think you will, sir. But the law\\
Is such a thing---and then he says, Read's matter\\
Falling so lately.\\
\dapperspeaks \vinphantom{Falling so lately.} Read! he was an ass,\\
And dealt, sir, with a fool.\\
\facespeaks \vinphantom{And dealt, sir, with a fool.} It was a clerk, sir.\\
\dapperspeaks A clerk!\\
\facespeaks \vinphantom{A clerk!} Nay, hear me, sir. You know the law\\
Better, I think---\\
\dapperspeaks \vinphantom{Better, I think---} I should, sir, and the danger:\\
You know, I shewed the statute to you.\\
\facespeaks You did so.\\
\dapperspeaks \vinphantom{You did so.} And will I tell then! By this hand of flesh,\\
Would it might never write good court-hand more,\\
If I discover. What do you think of me,\\
That I am a chiaus?\\
\facespeaks \vinphantom{That I am a chiaus?} What's that?\\
\dapperspeaks \vinphantom{That I am a chiaus? What's that?} The Turk was here.\\
As one would say, do you think I am a Turk?\\
\facespeaks I'll tell the doctor so.\\
\dapperspeaks \vinphantom{I'll tell the doctor so.} Do, good sweet captain.\\
\facespeaks Come, noble doctor, pray thee let's prevail;\\
This is the gentleman, and he is no chiaus.\\
\subtlespeaks Captain, I have return'd you all my answer.\\
I would do much, sir, for your love---But this\\
I neither may, nor can.\\
\facespeaks \vinphantom{I neither may, nor can.} Tut, do not say so.\\
You deal now with a noble fellow, doctor,\\
One that will thank you richly; and he is no chiaus:\\
Let that, sir, move you.\\
\subtlespeaks \vinphantom{Let that, sir, move you.} Pray you, forbear---\\
\facespeaks \vinphantom{Let that, sir, move you. Pray you, forbear---} He has\\
Four angels here.\\
\subtlespeaks \vinphantom{Four angels here.} You do me wrong, good sir.\\
\facespeaks Doctor, wherein? to tempt you with these spirits?\\
\subtlespeaks To tempt my art and love, sir, to my peril.\\
Fore heaven, I scarce can think you are my friend,\\
That so would draw me to apparent danger.\\
\facespeaks I draw you! a horse draw you, and a halter,\\
You, and your flies together---\\
\dapperspeaks \vinphantom{You, and your flies together---} Nay, good captain.\\
\facespeaks That know no difference of men.\\
\subtlespeaks \vinphantom{That know no difference of men.} Good words, sir.\\
\facespeaks Good deeds, sir, doctor dogs-meat. 'Slight, I bring you\\
No cheating Clim o' the Cloughs or Claribels,\\
That look as big as five-and-fifty, and flush;\\
And spit out secrets like hot custard---\\
\dapperspeaks \vinphantom{And spit out secrets like hot custard---} Captain!\\
\facespeaks Nor any melancholic under-scribe,\\
Shall tell the vicar; but a special gentle,\\
That is the heir to forty marks a year,\\
Consorts with the small poets of the time,\\
Is the sole hope of his old grandmother;\\
That knows the law, and writes you six fair hands,\\
Is a fine clerk, and has his cyphering perfect.\\
Will take his oath o' the Greek Testament,\\
If need be, in his pocket; and can court\\
His mistress out of Ovid.\\
\dapperspeaks \vinphantom{His mistress out of Ovid.} Nay, dear captain---\\
\facespeaks Did you not tell me so?\\
\dapperspeaks \vinphantom{Did you not tell me so?} Yes; but I'd have you\\
Use master doctor with some more respect.\\
\facespeaks Hang him, proud stag, with his broad velvet head!---\\
But for your sake, I'd choak, ere I would change\\
An article of breath with such a puckfist:\\
Come, let's be gone.\\
\direct*{GOING.}
\subtlespeaks \vinphantom{Come, let's be gone.} Pray you let me speak with you.\\
\dapperspeaks His worship calls you, captain.\\
\facespeaks \vinphantom{His worship calls you, captain.} I am sorry\\
I e'er embark'd myself in such a business.\\
\dapperspeaks Nay, good sir; he did call you.\\
\facespeaks \vinphantom{Nay, good sir; he did call you.} Will he take then?\\
\subtlespeaks First, hear me---\\
\facespeaks \vinphantom{First, hear me---} Not a syllable, 'less you take.\\
\subtlespeaks Pray you, sir---\\
\facespeaks \vinphantom{Pray you, sir---} Upon no terms but an assumpsit.\\
\subtlespeaks Your humour must be law.\\
\direct*{HE TAKES THE FOUR ANGELS.}
\facespeaks \vinphantom{Your humour must be law.} Why now, sir, talk.\\
Now I dare hear you with mine honour. Speak.\\
So may this gentleman too.\\
\subtlespeaks \vinphantom{So may this gentleman too.} Why, sir---\\
\direct*{OFFERING TO WHISPER FACE.}
\facespeaks \vinphantom{So may this gentleman too. Why, sir---} No whispering.\\
\subtlespeaks Fore heaven, you do not apprehend the loss\\
You do yourself in this.\\
\facespeaks \vinphantom{You do yourself in this.} Wherein? for what?\\
\subtlespeaks Marry, to be so importunate for one,\\
That, when he has it, will undo you all:\\
He'll win up all the money in the town.\\
\facespeaks How!\\
\subtlespeaks \vinphantom{How!} Yes, and blow up gamester after gamester,\\
As they do crackers in a puppet-play.\\
If I do give him a familiar,\\
Give you him all you play for; never set him:\\
For he will have it.\\
\facespeaks \vinphantom{For he will have it.} You are mistaken, doctor.\\
Why he does ask one but for cups and horses,\\
A rifling fly; none of your great familiars.\\
\dapperspeaks Yes, captain, I would have it for all games.\\
\subtlespeaks I told you so.\\
\facespeaks \direct*{TAKING DAP. ASIDE} 'Slight, that is a new business!\\
I understood you, a tame bird, to fly\\
Twice in a term, or so, on Friday nights,\\
When you had left the office, for a nag\\
Of forty or fifty shillings.\\
\dapperspeaks \vinphantom{Of forty or fifty shillings.} Ay, 'tis true, sir;\\
But I do think now I shall leave the law,\\
And therefore---\\
\facespeaks \vinphantom{And therefore---} Why, this changes quite the case.\\
Do you think that I dare move him?\\
\dapperspeaks \vinphantom{Do you think that I dare move him?} If you please, sir;\\
All's one to him, I see.\\
\facespeaks \vinphantom{All's one to him, I see.} What! for that money?\\
I cannot with my conscience; nor should you\\
Make the request, methinks.\\
\dapperspeaks \vinphantom{Make the request, methinks.} No, sir, I mean\\
To add consideration.\\
\facespeaks \vinphantom{To add consideration.} Why then, sir,\\
I'll try.---\\
\direct*{GOES TO SUBTLE.}
Say that it were for all games, doctor.\\
\subtlespeaks \vinphantom{Say that it were for all games, doctor.} I say then, not a mouth shall eat for him\\
At any ordinary, but on the score,\\
That is a gaming mouth, conceive me.\\
\facespeaks \vinphantom{That is a gaming mouth, conceive me.} Indeed!\\
\subtlespeaks He'll draw you all the treasure of the realm,\\
If it be set him.\\
\facespeaks \vinphantom{If it be set him.} Speak you this from art?\\
\subtlespeaks Ay, sir, and reason too, the ground of art.\\
He is of the only best complexion,\\
The queen of Fairy loves.\\
\facespeaks \vinphantom{The queen of Fairy loves.} What! is he?\\
\subtlespeaks \vinphantom{The queen of Fairy loves. What! is he?} Peace.\\
He'll overhear you. Sir, should she but see him---\\
\facespeaks What?\\
\subtlespeaks \vinphantom{What?} Do not you tell him.\\
\facespeaks \vinphantom{What? Do not you tell him.} Will he win at cards too?\\
\subtlespeaks The spirits of dead Holland, living Isaac,\\
You'd swear, were in him; such a vigorous luck\\
As cannot be resisted. 'Slight, he'll put\\
Six of your gallants to a cloke, indeed.\\
\facespeaks A strange success, that some man shall be born to.\\
\subtlespeaks He hears you, man---\\
\dapperspeaks \vinphantom{He hears you, man---} Sir, I'll not be ingrateful.\\
\facespeaks Faith, I have confidence in his good nature:\\
You hear, he says he will not be ingrateful.\\
\subtlespeaks Why, as you please; my venture follows yours.\\
\facespeaks Troth, do it, doctor; think him trusty, and make him.\\
He may make us both happy in an hour;\\
Win some five thousand pound, and send us two on't.\\
\dapperspeaks Believe it, and I will, sir.\\
\facespeaks \vinphantom{Believe it, and I will, sir.} And you shall, sir.\\
\direct*{TAKES HIM ASIDE.}
You have heard all?\\
\dapperspeaks \vinphantom{You have heard all?} No, what was't? Nothing, I, sir.\\
\facespeaks Nothing!\\
\dapperspeaks \vinphantom{Nothing!} A little, sir.\\
\facespeaks \vinphantom{Nothing! A little, sir.} Well, a rare star\\
Reign'd at your birth.\\
\dapperspeaks \vinphantom{Reign'd at your birth.} At mine, sir! No.\\
\facespeaks \vinphantom{Reign'd at your birth. At mine, sir! No.} The doctor\\
Swears that you are---\\
\subtlespeaks \vinphantom{Swears that you are---} Nay, captain, you'll tell all now.\\
\facespeaks Allied to the queen of Fairy.\\
\dapperspeaks \vinphantom{Allied to the queen of Fairy.} Who! that I am?\\
Believe it, no such matter---\\
\facespeaks \vinphantom{Believe it, no such matter---} Yes, and that\\
You were born with a cawl on your head.\\
\dapperspeaks \vinphantom{You were born with a cawl on your head.} Who says so?\\
\facespeaks Come,\\
You know it well enough, though you dissemble it.\\
\dapperspeaks I'fac, I do not; you are mistaken.\\
\facespeaks How!\\
Swear by your fac, and in a thing so known\\
Unto the doctor? How shall we, sir, trust you\\
In the other matter? can we ever think,\\
When you have won five or six thousand pound,\\
You'll send us shares in't, by this rate?\\
\dapperspeaks \vinphantom{You'll send us shares in't, by this rate?} By Jove, sir,\\
I'll win ten thousand pound, and send you half.\\
I'fac's no oath.\\
\subtlespeaks \vinphantom{I'fac's no oath.} No, no, he did but jest.\\
\facespeaks Go to. Go thank the doctor: he's your friend,\\
To take it so.\\
\dapperspeaks \vinphantom{To take it so.} I thank his worship.\\
\facespeaks \vinphantom{To take it so. I thank his worship.} So!\\
Another angel.\\
\dapperspeaks \vinphantom{Another angel.} Must I?\\
\facespeaks \vinphantom{Another angel. Must I?} Must you! 'slight,\\
What else is thanks? will you be trivial?---Doctor,\\
\direct*{DAPPER GIVES HIM THE MONEY.}
When must he come for his familiar?\\
\dapperspeaks \vinphantom{When must he come for his familiar?} Shall I not have it with me?\\
\subtlespeaks O, good sir!\\
There must a world of ceremonies pass;\\
You must be bath'd and fumigated first:\\
Besides the queen of Fairy does not rise\\
Till it be noon.\\
\facespeaks \vinphantom{Till it be noon.} Not, if she danced, to-night.\\
\subtlespeaks And she must bless it.\\
\facespeaks \vinphantom{And she must bless it.} Did you never see\\
Her royal grace yet?\\
\dapperspeaks \vinphantom{Her royal grace yet?} Whom?\\
\facespeaks \vinphantom{Her royal grace yet? Whom?} Your aunt of Fairy?\\
\subtlespeaks Not since she kist him in the cradle, captain;\\
I can resolve you that.\\
\facespeaks \vinphantom{I can resolve you that.} Well, see her grace,\\
Whate'er it cost you, for a thing that I know.\\
It will be somewhat hard to compass; but\\
However, see her. You are made, believe it,\\
If you can see her. Her grace is a lone woman,\\
And very rich; and if she take a fancy,\\
She will do strange things. See her, at any hand.\\
'Slid, she may hap to leave you all she has:\\
It is the doctor's fear.\\
\dapperspeaks \vinphantom{It is the doctor's fear.} How will't be done, then?\\
\facespeaks Let me alone, take you no thought. Do you\\
But say to me, captain, I'll see her grace.\\
\dapperspeaks "Captain, I'll see her grace."\\
\facespeaks \vinphantom{"Captain, I'll see her grace."} Enough.\\
\direct*{KNOCKING WITHIN.}
\subtlespeaks \vinphantom{"Captain, I'll see her grace." Enough.} Who's there?\\
Anon.\\
\direct*{ASIDE TO FACE.}
---Conduct him forth by the back way.---\\
Sir, against one o'clock prepare yourself;\\
Till when you must be fasting; only take\\
Three drops of vinegar in at your nose,\\
Two at your mouth, and one at either ear;\\
Then bathe your fingers' ends and wash your eyes,\\
To sharpen your five senses, and cry "hum"\\
Thrice, and then "buz" as often; and then come.\\
\direct*{EXIT.}
\facespeaks Can you remember this?\\
\dapperspeaks \vinphantom{Can you remember this?} I warrant you.\\
\facespeaks Well then, away. It is but your bestowing\\
Some twenty nobles 'mong her grace's servants,\\
And put on a clean shirt: you do not know\\
What grace her grace may do you in clean linen.\\
\direct*{EXEUNT FACE AND DAPPER.}
\subtlespeaks \direct*{WITHIN} Come in! Good wives, I pray you forbear me now;\\
Troth I can do you no good till afternoon---\\
\direct*{RE-ENTERS, FOLLOWED BY DRUGGER.}
What is your name, say you? Abel Drugger?\\
\druggerspeaks Yes, sir.\\
\subtlespeaks \vinphantom{Yes, sir.} A seller of tobacco?\\
\druggerspeaks \vinphantom{Yes, sir. A seller of tobacco?} Yes, sir.\\
\subtlespeaks Umph!\\
Free of the grocers?\\
\druggerspeaks \vinphantom{Free of the grocers?} Ay, an't please you.\\
\subtlespeaks \vinphantom{Free of the grocers? Ay, an't please you.} Well---\\
Your business, Abel?\\
\druggerspeaks \vinphantom{Your business, Abel?} This, an't please your worship;\\
I am a young beginner, and am building\\
Of a new shop, an't like your worship, just\\
At corner of a street:---Here is the plot on't---\\
And I would know by art, sir, of your worship,\\
Which way I should make my door, by necromancy,\\
And where my shelves; and which should be for boxes,\\
And which for pots. I would be glad to thrive, sir:\\
And I was wish'd to your worship by a gentleman,\\
One captain Face, that says you know men's planets,\\
And their good angels, and their bad.\\
\subtlespeaks \vinphantom{And their good angels, and their bad.} I do,\\
If I do see them---\\
\direct*{RE-ENTER FACE.}
\facespeaks \vinphantom{If I do see them---} What! my honest Abel?\\
Though art well met here.\\
\druggerspeaks \vinphantom{Though art well met here.} Troth, sir, I was speaking,\\
Just as your worship came here, of your worship:\\
I pray you speak for me to master doctor.\\
\facespeaks He shall do any thing.---Doctor, do you hear?\\
This is my friend, Abel, an honest fellow;\\
He lets me have good tobacco, and he does not\\
Sophisticate it with sack-lees or oil,\\
Nor washes it in muscadel and grains,\\
Nor buries it in gravel, under ground,\\
Wrapp'd up in greasy leather, or piss'd clouts:\\
But keeps it in fine lily pots, that, open'd,\\
Smell like conserve of roses, or French beans.\\
He has his maple block, his silver tongs,\\
Winchester pipes, and fire of Juniper:\\
A neat, spruce, honest fellow, and no goldsmith.\\
\subtlespeaks He is a fortunate fellow, that I am sure on.\\
\facespeaks Already, sir, have you found it? Lo thee, Abel!\\
\subtlespeaks And in right way toward riches---\\
\facespeaks \vinphantom{And in right way toward riches---} Sir!\\
\subtlespeaks \vinphantom{And in right way toward riches--- Sir!} This summer\\
He will be of the clothing of his company,\\
And next spring call'd to the scarlet; spend what he can.\\
\facespeaks What, and so little beard?\\
\subtlespeaks \vinphantom{What, and so little beard?} Sir, you must think,\\
He may have a receipt to make hair come:\\
But he'll be wise, preserve his youth, and fine for't;\\
His fortune looks for him another way.\\
\facespeaks 'Slid, doctor, how canst thou know this so soon?\\
I am amused at that!\\
\subtlespeaks \vinphantom{I am amused at that!} By a rule, captain,\\
In metoposcopy, which I do work by;\\
A certain star in the forehead, which you see not.\\
Your chestnut or your olive-colour'd face\\
Does never fail: and your long ear doth promise.\\
I knew't by certain spots, too, in his teeth,\\
And on the nail of his mercurial finger.\\
\facespeaks Which finger's that?\\
\subtlespeaks \vinphantom{Which finger's that?} His little finger. Look.\\
You were born upon a Wednesday?\\
\druggerspeaks \vinphantom{You were born upon a Wednesday?} Yes, indeed, sir.\\
\subtlespeaks The thumb, in chiromancy, we give Venus;\\
The fore-finger, to Jove; the midst, to Saturn;\\
The ring, to Sol; the least, to Mercury,\\
Who was the lord, sir, of his horoscope,\\
His house of life being Libra; which fore-shew'd,\\
He should be a merchant, and should trade with balance.\\
\facespeaks Why, this is strange! Is it not, honest Nab?\\
\subtlespeaks There is a ship now, coming from Ormus,\\
That shall yield him such a commodity\\
Of drugs\\
\direct*{POINTING TO THE PLAN.}
---This is the west, and this the south?\\
\druggerspeaks \vinphantom{---This is the west, and this the south?} Yes, sir.\\
\subtlespeaks And those are your two sides?\\
\druggerspeaks \vinphantom{And those are your two sides?} Ay, sir.\\
\subtlespeaks \vinphantom{And those are your two sides? Ay, sir.} Make me your door, then, south; your broad side, west:\\
And on the east side of your shop, aloft,\\
Write Mathlai, Tarmiel, and Baraborat;\\
Upon the north part, Rael, Velel, Thiel.\\
They are the names of those mercurial spirits,\\
That do fright flies from boxes.\\
\druggerspeaks \vinphantom{That do fright flies from boxes.} Yes, sir.\\
\subtlespeaks \vinphantom{That do fright flies from boxes. Yes, sir.} And\\
Beneath your threshold, bury me a load-stone\\
To draw in gallants that wear spurs: the rest,\\
They'll seem to follow.\\
\facespeaks \vinphantom{They'll seem to follow.} That's a secret, Nab!\\
\subtlespeaks And, on your stall, a puppet, with a vice\\
And a court-fucus to call city-dames:\\
You shall deal much with minerals.\\
\druggerspeaks \vinphantom{You shall deal much with minerals.} Sir, I have.\\
At home, already---\\
\subtlespeaks \vinphantom{At home, already---} Ay, I know you have arsenic,\\
Vitriol, sal-tartar, argaile, alkali,\\
Cinoper: I know all.---This fellow, captain,\\
Will come, in time, to be a great distiller,\\
And give a say---I will not say directly,\\
But very fair---at the philosopher's stone.\\
\facespeaks Why, how now, Abel! is this true?\\
\druggerspeaks \direct*{ASIDE TO FACE} Good captain,\\
What must I give?\\
\facespeaks \vinphantom{What must I give?} Nay, I'll not counsel thee.\\
Thou hear'st what wealth (he says, spend what thou canst,)\\
Thou'rt like to come to.\\
\druggerspeaks \vinphantom{Thou'rt like to come to.} I would gi' him a crown.\\
\facespeaks A crown! and toward such a fortune? heart,\\
Thou shalt rather gi' him thy shop. No gold about thee?\\
\druggerspeaks Yes, I have a portague, I have kept this half-year.\\
\facespeaks Out on thee, Nab! 'Slight, there was such an offer---\\
Shalt keep't no longer, I'll give't him for thee. Doctor,\\
Nab prays your worship to drink this, and swears\\
He will appear more grateful, as your skill\\
Does raise him in the world.\\
\druggerspeaks \vinphantom{Does raise him in the world.} I would entreat\\
Another favour of his worship.\\
\facespeaks What is't, Nab?\\
\druggerspeaks \vinphantom{What is't, Nab?} But to look over, sir, my almanack,\\
And cross out my ill-days, that I may neither\\
Bargain, nor trust upon them.\\
\facespeaks \vinphantom{Bargain, nor trust upon them.} That he shall, Nab:\\
Leave it, it shall be done, 'gainst afternoon.\\
\subtlespeaks And a direction for his shelves.\\
\facespeaks \vinphantom{And a direction for his shelves.} Now, Nab,\\
Art thou well pleased, Nab?\\
\druggerspeaks \vinphantom{Art thou well pleased, Nab?} 'Thank, sir, both your worships.\\
\facespeaks Away.\\
\direct*{EXIT DRUGGER.}
Why, now, you smoaky persecutor of nature!\\
Now do you see, that something's to be done,\\
Beside your beech-coal, and your corsive waters,\\
Your crosslets, crucibles, and cucurbites?\\
You must have stuff brought home to you, to work on:\\
And yet you think, I am at no expense\\
In searching out these veins, then following them,\\
Then trying them out. 'Fore God, my intelligence\\
Costs me more money, than my share oft comes to,\\
In these rare works.\\
\subtlespeaks \vinphantom{In these rare works.} You are pleasant, sir.\\
\direct*{RE-ENTER DOL.}
---How now!\\
What says my dainty Dolkin?\\
\dolspeaks \vinphantom{What says my dainty Dolkin?} Yonder fish-wife\\
Will not away. And there's your giantess,\\
The bawd of Lambeth.\\
\subtlespeaks \vinphantom{The bawd of Lambeth.} Heart, I cannot speak with them.\\
\dolspeaks Not afore night, I have told them in a voice,\\
Thorough the trunk, like one of your familiars.\\
But I have spied sir Epicure Mammon---\\
\subtlespeaks Where?\\
\dolspeaks \vinphantom{Where?} Coming along, at far end of the lane,\\
Slow of his feet, but earnest of his tongue\\
To one that's with him.\\
\subtlespeaks \vinphantom{To one that's with him.} Face, go you and shift.\\
\direct*{EXIT FACE.}
Dol, you must presently make ready, too.\\
\dolspeaks Why, what's the matter?\\
\subtlespeaks \vinphantom{Why, what's the matter?} O, I did look for him\\
With the sun's rising: 'marvel he could sleep,\\
This is the day I am to perfect for him\\
The magisterium, our great work, the stone;\\
And yield it, made, into his hands: of which\\
He has, this month, talked as he were possess'd.\\
And now he's dealing pieces on't away.---\\
Methinks I see him entering ordinaries,\\
Dispensing for the pox, and plaguy houses,\\
Reaching his dose, walking Moorfields for lepers,\\
And offering citizens' wives pomander-bracelets,\\
As his preservative, made of the elixir;\\
Searching the spittal, to make old bawds young;\\
And the highways, for beggars, to make rich.\\
I see no end of his labours. He will make\\
Nature asham'd of her long sleep: when art,\\
Who's but a step-dame, shall do more than she,\\
In her best love to mankind, ever could:\\
If his dream lasts, he'll turn the age to gold.\\
\direct*{EXEUNT.}
\act

\scene

\stagedir{AN OUTER ROOM IN LOVEWIT'S HOUSE. ENTER SIR EPICURE MAMMON AND SURLY.}
\mammonspeaks Come on, sir. Now, you set your foot on shore\\
In Novo Orbe; here's the rich Peru:\\
And there within, sir, are the golden mines,\\
Great Solomon's Ophir! he was sailing to't,\\
Three years, but we have reached it in ten months.\\
This is the day, wherein, to all my friends,\\
I will pronounce the happy word, BE RICH;\\
THIS DAY YOU SHALL BE SPECTATISSIMI.\\
You shall no more deal with the hollow dye,\\
Or the frail card. No more be at charge of keeping\\
The livery-punk for the young heir, that must\\
Seal, at all hours, in his shirt: no more,\\
If he deny, have him beaten to't, as he is\\
That brings him the commodity. No more\\
Shall thirst of satin, or the covetous hunger\\
Of velvet entrails for a rude-spun cloke,\\
To be display'd at madam Augusta's, make\\
The sons of Sword and Hazard fall before\\
The golden calf, and on their knees, whole nights\\
Commit idolatry with wine and trumpets:\\
Or go a feasting after drum and ensign.\\
No more of this. You shall start up young viceroys,\\
And have your punks, and punketees, my Surly.\\
And unto thee I speak it first, BE RICH.\\
Where is my Subtle, there? Within, ho!\\
\facespeaks \direct*{WITHIN} Sir, he'll come to you by and by.\\
\mammonspeaks \vinphantom{Where is my Subtle, there? Within, ho!} That is his fire-drake,\\
His Lungs, his Zephyrus, he that puffs his coals,\\
Till he firk nature up, in her own centre.\\
You are not faithful, sir. This night, I'll change\\
All that is metal, in my house, to gold:\\
And, early in the morning, will I send\\
To all the plumbers and the pewterers,\\
And by their tin and lead up; and to Lothbury\\
For all the copper.\\
\surlyspeaks \vinphantom{For all the copper.} What, and turn that too?\\
\mammonspeaks Yes, and I'll purchase Devonshire and Cornwall,\\
And make them perfect Indies! you admire now?\\
\surlyspeaks No, faith.\\
\mammonspeaks \vinphantom{No, faith.} But when you see th' effects of the Great Medicine,\\
Of which one part projected on a hundred\\
Of Mercury, or Venus, or the moon,\\
Shall turn it to as many of the sun;\\
Nay, to a thousand, so ad infinitum:\\
You will believe me.\\
\surlyspeaks \vinphantom{You will believe me.} Yes, when I see't, I will.\\
But if my eyes do cozen me so, and I\\
Giving them no occasion, sure I'll have\\
A whore, shall piss them out next day.\\
\mammonspeaks \vinphantom{A whore, shall piss them out next day.} Ha! why?\\
Do you think I fable with you? I assure you,\\
He that has once the flower of the sun,\\
The perfect ruby, which we call elixir,\\
Not only can do that, but, by its virtue,\\
Can confer honour, love, respect, long life;\\
Give safety, valour, yea, and victory,\\
To whom he will. In eight and twenty days,\\
I'll make an old man of fourscore, a child.\\
\surlyspeaks No doubt; he's that already.\\
\mammonspeaks \vinphantom{No doubt; he's that already.} Nay, I mean,\\
Restore his years, renew him, like an eagle,\\
To the fifth age; make him get sons and daughters,\\
Young giants; as our philosophers have done,\\
The ancient patriarchs, afore the flood,\\
But taking, once a week, on a knife's point,\\
The quantity of a grain of mustard of it;\\
Become stout Marses, and beget young Cupids.\\
\surlyspeaks The decay'd vestals of Pict-hatch would thank you,\\
That keep the fire alive, there.\\
\mammonspeaks \vinphantom{That keep the fire alive, there.} 'Tis the secret\\
Of nature naturis'd 'gainst all infections,\\
Cures all diseases coming of all causes;\\
A month's grief in a day, a year's in twelve;\\
And, of what age soever, in a month:\\
Past all the doses of your drugging doctors.\\
I'll undertake, withal, to fright the plague\\
Out of the kingdom in three months.\\
\surlyspeaks \vinphantom{Out of the kingdom in three months.} And I'll\\
Be bound, the players shall sing your praises, then,\\
Without their poets.\\
\mammonspeaks \vinphantom{Without their poets.} Sir, I'll do't. Mean time,\\
I'll give away so much unto my man,\\
Shall serve the whole city, with preservative\\
Weekly; each house his dose, and at the rate---\\
\surlyspeaks As he that built the Water-work, does with water?\\
\mammonspeaks You are incredulous.\\
\surlyspeaks \vinphantom{You are incredulous.} Faith I have a humour,\\
I would not willingly be gull'd. Your stone\\
Cannot transmute me.\\
\mammonspeaks \vinphantom{Cannot transmute me.} Pertinax, my Surly,\\
Will you believe antiquity? records?\\
I'll shew you a book where Moses and his sister,\\
And Solomon have written of the art;\\
Ay, and a treatise penn'd by Adam---\\
\surlyspeaks \vinphantom{Ay, and a treatise penn'd by Adam---} How!\\
\mammonspeaks Of the philosopher's stone, and in High Dutch.\\
\surlyspeaks Did Adam write, sir, in High Dutch?\\
\mammonspeaks \vinphantom{Did Adam write, sir, in High Dutch?} He did;\\
Which proves it was the primitive tongue.\\
\surlyspeaks \vinphantom{Which proves it was the primitive tongue.} What paper?\\
\mammonspeaks On cedar board.\\
\surlyspeaks \vinphantom{On cedar board.} O that, indeed, they say,\\
Will last 'gainst worms.\\
\mammonspeaks \vinphantom{Will last 'gainst worms.} 'Tis like your Irish wood,\\
'Gainst cob-webs. I have a piece of Jason's fleece, too,\\
Which was no other than a book of alchemy,\\
Writ in large sheep-skin, a good fat ram-vellum.\\
Such was Pythagoras' thigh, Pandora's tub,\\
And, all that fable of Medea's charms,\\
The manner of our work; the bulls, our furnace,\\
Still breathing fire; our argent-vive, the dragon:\\
The dragon's teeth, mercury sublimate,\\
That keeps the whiteness, hardness, and the biting;\\
And they are gathered into Jason's helm,\\
The alembic, and then sow'd in Mars his field,\\
And thence sublimed so often, till they're fixed.\\
Both this, the Hesperian garden, Cadmus' story,\\
Jove's shower, the boon of Midas, Argus' eyes,\\
Boccace his Demogorgon, thousands more,\\
All abstract riddles of our stone.\\
\direct*{ENTER FACE, AS A SERVANT.}
---How now!\\
Do we succeed? Is our day come? and holds it?\\
\facespeaks The evening will set red upon you, sir;\\
You have colour for it, crimson: the red ferment\\
Has done his office; three hours hence prepare you\\
To see projection.\\
\mammonspeaks \vinphantom{To see projection.} Pertinax, my Surly.\\
Again I say to thee, aloud, Be rich.\\
This day, thou shalt have ingots; and to-morrow,\\
Give lords th' affront.---Is it, my Zephyrus, right?\\
Blushes the bolt's-head?\\
\facespeaks \vinphantom{Blushes the bolt's-head?} Like a wench with child, sir,\\
That were but now discover'd to her master.\\
\mammonspeaks Excellent witty Lungs!---my only care\\
Where to get stuff enough now, to project on;\\
This town will not half serve me.\\
\facespeaks \vinphantom{This town will not half serve me.} No, sir! buy\\
The covering off o' churches.\\
\mammonspeaks \vinphantom{The covering off o' churches.} That's true.\\
\facespeaks Yes.\\
Let them stand bare, as do their auditory;\\
Or cap them, new, with shingles.\\
\mammonspeaks \vinphantom{Or cap them, new, with shingles.} No, good thatch:\\
Thatch will lie light upon the rafters, Lungs.---\\
Lungs, I will manumit thee from the furnace;\\
I will restore thee thy complexion, Puffe,\\
Lost in the embers; and repair this brain,\\
Hurt with the fume o' the metals.\\
\facespeaks \vinphantom{Hurt with the fume o' the metals.} I have blown, sir,\\
Hard for your worship; thrown by many a coal,\\
When 'twas not beech; weigh'd those I put in, just,\\
To keep your heat still even; these blear'd eyes\\
Have wak'd to read your several colours, sir,\\
Of the pale citron, the green lion, the crow,\\
The peacock's tail, the plumed swan.\\
\mammonspeaks \vinphantom{The peacock's tail, the plumed swan.} And, lastly,\\
Thou hast descry'd the flower, the sanguis agni?\\
\facespeaks Yes, sir.\\
\mammonspeaks \vinphantom{Yes, sir.} Where's master?\\
\facespeaks \vinphantom{Yes, sir. Where's master?} At his prayers, sir, he;\\
Good man, he's doing his devotions\\
For the success.\\
\mammonspeaks \vinphantom{For the success.} Lungs, I will set a period\\
To all thy labours; thou shalt be the master\\
Of my seraglio.\\
\facespeaks \vinphantom{Of my seraglio.} Good, sir.\\
\mammonspeaks \vinphantom{Of my seraglio. Good, sir.} But do you hear?\\
I'll geld you, Lungs.\\
\facespeaks \vinphantom{I'll geld you, Lungs.} Yes, sir.\\
\mammonspeaks \vinphantom{I'll geld you, Lungs. Yes, sir.} For I do mean\\
To have a list of wives and concubines,\\
Equal with Solomon, who had the stone\\
Alike with me; and I will make me a back\\
With the elixir, that shall be as tough\\
As Hercules, to encounter fifty a night.---\\
Thou'rt sure thou saw'st it blood?\\
\facespeaks \vinphantom{Thou'rt sure thou saw'st it blood?} Both blood and spirit, sir.\\
\mammonspeaks I will have all my beds blown up, not stuft;\\
Down is too hard: and then, mine oval room\\
Fill'd with such pictures as Tiberius took\\
From Elephantis, and dull Aretine\\
But coldly imitated. Then, my glasses\\
Cut in more subtle angles, to disperse\\
And multiply the figures, as I walk\\
Naked between my succubae. My mists\\
I'll have of perfume, vapour'd 'bout the room,\\
To lose ourselves in; and my baths, like pits\\
To fall into; from whence we will come forth,\\
And roll us dry in gossamer and roses.---\\
Is it arrived at ruby?---Where I spy\\
A wealthy citizen, or a rich lawyer,\\
Have a sublimed pure wife, unto that fellow\\
I'll send a thousand pound to be my cuckold.\\
\facespeaks And I shall carry it?\\
\mammonspeaks \vinphantom{And I shall carry it?} No. I'll have no bawds,\\
But fathers and mothers: they will do it best,\\
Best of all others. And my flatterers\\
Shall be the pure and gravest of divines,\\
That I can get for money. My mere fools,\\
Eloquent burgesses, and then my poets\\
The same that writ so subtly of the fart,\\
Whom I will entertain still for that subject.\\
The few that would give out themselves to be\\
Court and town-stallions, and, each-where, bely\\
Ladies who are known most innocent for them;\\
Those will I beg, to make me eunuchs of:\\
And they shall fan me with ten estrich tails\\
A-piece, made in a plume to gather wind.\\
We will be brave, Puffe, now we have the med'cine.\\
My meat shall all come in, in Indian shells,\\
Dishes of agat set in gold, and studded\\
With emeralds, sapphires, hyacinths, and rubies.\\
The tongues of carps, dormice, and camels' heels,\\
Boil'd in the spirit of sol, and dissolv'd pearl,\\
Apicius' diet, 'gainst the epilepsy:\\
And I will eat these broths with spoons of amber,\\
Headed with diamond and carbuncle.\\
My foot-boy shall eat pheasants, calver'd salmons,\\
Knots, godwits, lampreys: I myself will have\\
The beards of barbels served, instead of sallads;\\
Oil'd mushrooms; and the swelling unctuous paps\\
Of a fat pregnant sow, newly cut off,\\
Drest with an exquisite, and poignant sauce;\\
For which, I'll say unto my cook, "There's gold,\\
Go forth, and be a knight."\\
\facespeaks \vinphantom{Go forth, and be a knight."} Sir, I'll go look\\
A little, how it heightens.\\
\direct*{EXIT.}
\mammonspeaks \vinphantom{A little, how it heightens.} Do.---My shirts\\
I'll have of taffeta-sarsnet, soft and light\\
As cobwebs; and for all my other raiment,\\
It shall be such as might provoke the Persian,\\
Were he to teach the world riot anew.\\
My gloves of fishes' and birds' skins, perfumed\\
With gums of paradise, and eastern air---\\
\surlyspeaks And do you think to have the stone with this?\\
\mammonspeaks No, I do think t' have all this with the stone.\\
\surlyspeaks Why, I have heard he must be homo frugi,\\
A pious, holy, and religious man,\\
One free from mortal sin, a very virgin.\\
\mammonspeaks That makes it, sir; he is so: but I buy it;\\
My venture brings it me. He, honest wretch,\\
A notable, superstitious, good soul,\\
Has worn his knees bare, and his slippers bald,\\
With prayer and fasting for it: and, sir, let him\\
Do it alone, for me, still. Here he comes.\\
Not a profane word afore him: 'tis poison.---\\
\direct*{ENTER SUBTLE.}
Good morrow, father.\\
\subtlespeaks \vinphantom{Good morrow, father.} Gentle son, good morrow,\\
And to your friend there. What is he, is with you?\\
\mammonspeaks An heretic, that I did bring along,\\
In hope, sir, to convert him.\\
\subtlespeaks \vinphantom{In hope, sir, to convert him.} Son, I doubt\\
You are covetous, that thus you meet your time\\
In the just point: prevent your day at morning.\\
This argues something, worthy of a fear\\
Of importune and carnal appetite.\\
Take heed you do not cause the blessing leave you,\\
With your ungovern'd haste. I should be sorry\\
To see my labours, now even at perfection,\\
Got by long watching and large patience,\\
Not prosper where my love and zeal hath placed them.\\
Which (heaven I call to witness, with your self,\\
To whom I have pour'd my thoughts) in all my ends,\\
Have look'd no way, but unto public good,\\
To pious uses, and dear charity\\
Now grown a prodigy with men. Wherein\\
If you, my son, should now prevaricate,\\
And, to your own particular lusts employ\\
So great and catholic a bliss, be sure\\
A curse will follow, yea, and overtake\\
Your subtle and most secret ways.\\
\mammonspeaks \vinphantom{Your subtle and most secret ways.} I know, sir;\\
You shall not need to fear me; I but come,\\
To have you confute this gentleman.\\
\surlyspeaks Who is,\\
Indeed, sir, somewhat costive of belief\\
Toward your stone; would not be gull'd.\\
\subtlespeaks \vinphantom{Toward your stone; would not be gull'd.} Well, son,\\
All that I can convince him in, is this,\\
The WORK IS DONE, bright sol is in his robe.\\
We have a medicine of the triple soul,\\
The glorified spirit. Thanks be to heaven,\\
And make us worthy of it!---Ulen Spiegel!\\
\facespeaks \direct*{WITHIN} Anon, sir.\\
\subtlespeaks Look well to the register.\\
And let your heat still lessen by degrees,\\
To the aludels.\\
\facespeaks \direct*{WITHIN} Yes, sir.\\
\subtlespeaks \vinphantom{To the aludels.} Did you look\\
On the bolt's-head yet?\\
\facespeaks \direct*{WITHIN} Which? on D, sir?\\
\subtlespeaks \vinphantom{On the bolt's-head yet?} Ay;\\
What's the complexion?\\
\facespeaks \direct*{WITHIN} Whitish.\\
\subtlespeaks \vinphantom{What's the complexion?} Infuse vinegar,\\
To draw his volatile substance and his tincture:\\
And let the water in glass E be filter'd,\\
And put into the gripe's egg. Lute him well;\\
And leave him closed in balneo.\\
\facespeaks \direct*{WITHIN} I will, sir.\\
\surlyspeaks \vinphantom{And leave him closed in balneo.} What a brave language here is! next to canting.\\
\subtlespeaks I have another work, you never saw, son,\\
That three days since past the philosopher's wheel,\\
In the lent heat of Athanor; and's become\\
Sulphur of Nature.\\
\mammonspeaks \vinphantom{Sulphur of Nature.} But 'tis for me?\\
\subtlespeaks \vinphantom{Sulphur of Nature. But 'tis for me?} What need you?\\
You have enough in that is perfect.\\
\mammonspeaks \vinphantom{You have enough in that is perfect.} O but---\\
\subtlespeaks Why, this is covetise!\\
\mammonspeaks \vinphantom{Why, this is covetise!} No, I assure you,\\
I shall employ it all in pious uses,\\
Founding of colleges and grammar schools,\\
Marrying young virgins, building hospitals,\\
And now and then a church.\\
\direct*{RE-ENTER FACE.}
\subtlespeaks \vinphantom{And now and then a church.} How now!\\
\facespeaks \vinphantom{And now and then a church. How now!} Sir, please you,\\
Shall I not change the filter?\\
\subtlespeaks \vinphantom{Shall I not change the filter?} Marry, yes;\\
And bring me the complexion of glass B.\\
\direct*{EXIT FACE.}
\mammonspeaks Have you another?\\
\subtlespeaks \vinphantom{Have you another?} Yes, son; were I assured---\\
Your piety were firm, we would not want\\
The means to glorify it: but I hope the best.---\\
I mean to tinct C in sand-heat to-morrow,\\
And give him imbibition.\\
\mammonspeaks \vinphantom{And give him imbibition.} Of white oil?\\
\subtlespeaks No, sir, of red. F is come over the helm too,\\
I thank my Maker, in S. Mary's bath,\\
And shews lac virginis. Blessed be heaven!\\
I sent you of his faeces there calcined:\\
Out of that calx, I have won the salt of mercury.\\
\mammonspeaks By pouring on your rectified water?\\
\subtlespeaks Yes, and reverberating in Athanor.\\
\direct*{RE-ENTER FACE.}
How now! what colour says it?\\
\facespeaks \vinphantom{How now! what colour says it?} The ground black, sir.\\
\mammonspeaks That's your crow's head?\\
\surlyspeaks \vinphantom{That's your crow's head?} Your cock's-comb's, is it not?\\
\subtlespeaks No, 'tis not perfect. Would it were the crow!\\
That work wants something.\\
\surlyspeaks \direct*{ASIDE} O, I looked for this.\\
The hay's a pitching.\\
\subtlespeaks \vinphantom{The hay's a pitching.} Are you sure you loosed them\\
In their own menstrue?\\
\facespeaks \vinphantom{In their own menstrue?} Yes, sir, and then married them,\\
And put them in a bolt's-head nipp'd to digestion,\\
According as you bade me, when I set\\
The liquor of Mars to circulation\\
In the same heat.\\
\subtlespeaks \vinphantom{In the same heat.} The process then was right.\\
\facespeaks Yes, by the token, sir, the retort brake,\\
And what was saved was put into the pellican,\\
And sign'd with Hermes' seal.\\
\subtlespeaks \vinphantom{And sign'd with Hermes' seal.} I think 'twas so.\\
We should have a new amalgama.\\
\surlyspeaks \direct*{ASIDE} O, this ferret\\
Is rank as any pole-cat.\\
\subtlespeaks \vinphantom{Is rank as any pole-cat.} But I care not:\\
Let him e'en die; we have enough beside,\\
In embrion. H has his white shirt on?\\
\facespeaks Yes, sir,\\
He's ripe for inceration, he stands warm,\\
In his ash-fire. I would not you should let\\
Any die now, if I might counsel, sir,\\
For luck's sake to the rest: it is not good.\\
\mammonspeaks He says right.\\
\surlyspeaks \direct*{ASIDE} Ay, are you bolted?\\
\facespeaks \vinphantom{He says right.} Nay, I know't, sir,\\
I have seen the ill fortune. What is some three ounces\\
Of fresh materials?\\
\mammonspeaks \vinphantom{Of fresh materials?} Is't no more?\\
\facespeaks \vinphantom{Of fresh materials? Is't no more?} No more, sir.\\
Of gold, t'amalgame with some six of mercury.\\
\mammonspeaks Away, here's money. What will serve?\\
\facespeaks \vinphantom{Away, here's money. What will serve?} Ask him, sir.\\
\mammonspeaks How much?\\
\subtlespeaks \vinphantom{How much?} Give him nine pound:---you may give him ten.\\
\surlyspeaks Yes, twenty, and be cozen'd, do.\\
\mammonspeaks \vinphantom{Yes, twenty, and be cozen'd, do.} There 'tis.\\
\direct*{GIVES FACE THE MONEY.}
\subtlespeaks This needs not; but that you will have it so,\\
To see conclusions of all: for two\\
Of our inferior works are at fixation,\\
A third is in ascension. Go your ways.\\
Have you set the oil of luna in kemia?\\
\facespeaks Yes, sir.\\
\subtlespeaks \vinphantom{Yes, sir.} And the philosopher's vinegar?\\
\facespeaks Ay.\\
\direct*{EXIT.}
\surlyspeaks \vinphantom{Ay.} We shall have a sallad!\\
\mammonspeaks \vinphantom{Ay. We shall have a sallad!} When do you make projection?\\
\subtlespeaks Son, be not hasty, I exalt our med'cine,\\
By hanging him in balneo vaporoso,\\
And giving him solution; then congeal him;\\
And then dissolve him; then again congeal him;\\
For look, how oft I iterate the work,\\
So many times I add unto his virtue.\\
As, if at first one ounce convert a hundred,\\
After his second loose, he'll turn a thousand;\\
His third solution, ten; his fourth, a hundred:\\
After his fifth, a thousand thousand ounces\\
Of any imperfect metal, into pure\\
Silver or gold, in all examinations,\\
As good as any of the natural mine.\\
Get you your stuff here against afternoon,\\
Your brass, your pewter, and your andirons.\\
\mammonspeaks Not those of iron?\\
\subtlespeaks \vinphantom{Not those of iron?} Yes, you may bring them too:\\
We'll change all metals.\\
\surlyspeaks \vinphantom{We'll change all metals.} I believe you in that.\\
\mammonspeaks Then I may send my spits?\\
\subtlespeaks \vinphantom{Then I may send my spits?} Yes, and your racks.\\
\surlyspeaks And dripping-pans, and pot-hangers, and hooks?\\
Shall he not?\\
\subtlespeaks \vinphantom{Shall he not?} If he please.\\
\surlyspeaks \vinphantom{Shall he not? If he please.} ---To be an ass.\\
\subtlespeaks How, sir!\\
\mammonspeaks \vinphantom{How, sir!} This gentleman you must bear withal:\\
I told you he had no faith.\\
\surlyspeaks \vinphantom{I told you he had no faith.} And little hope, sir;\\
But much less charity, should I gull myself.\\
\subtlespeaks Why, what have you observ'd, sir, in our art,\\
Seems so impossible?\\
\surlyspeaks \vinphantom{Seems so impossible?} But your whole work, no more.\\
That you should hatch gold in a furnace, sir,\\
As they do eggs in Egypt!\\
\subtlespeaks \vinphantom{As they do eggs in Egypt!} Sir, do you\\
Believe that eggs are hatch'd so?\\
\surlyspeaks \vinphantom{Believe that eggs are hatch'd so?} If I should?\\
\subtlespeaks Why, I think that the greater miracle.\\
No egg but differs from a chicken more\\
Than metals in themselves.\\
\surlyspeaks \vinphantom{Than metals in themselves.} That cannot be.\\
The egg's ordain'd by nature to that end,\\
And is a chicken in potentia.\\
\subtlespeaks The same we say of lead and other metals,\\
Which would be gold, if they had time.\\
\mammonspeaks \vinphantom{Which would be gold, if they had time.} And that\\
Our art doth further.\\
\subtlespeaks \vinphantom{Our art doth further.} Ay, for 'twere absurb\\
To think that nature in the earth bred gold\\
Perfect in the instant: something went before.\\
There must be remote matter.\\
\surlyspeaks \vinphantom{There must be remote matter.} Ay, what is that?\\
\subtlespeaks Marry, we say---\\
\mammonspeaks \vinphantom{Marry, we say---} Ay, now it heats: stand, father,\\
Pound him to dust.\\
\subtlespeaks \vinphantom{Pound him to dust.} It is, of the one part,\\
A humid exhalation, which we call\\
Material liquida, or the unctuous water;\\
On the other part, a certain crass and vicious\\
Portion of earth; both which, concorporate,\\
Do make the elementary matter of gold;\\
Which is not yet propria materia,\\
But common to all metals and all stones;\\
For, where it is forsaken of that moisture,\\
And hath more driness, it becomes a stone:\\
Where it retains more of the humid fatness,\\
It turns to sulphur, or to quicksilver,\\
Who are the parents of all other metals.\\
Nor can this remote matter suddenly\\
Progress so from extreme unto extreme,\\
As to grow gold, and leap o'er all the means.\\
Nature doth first beget the imperfect, then\\
Proceeds she to the perfect. Of that airy\\
And oily water, mercury is engender'd;\\
Sulphur of the fat and earthy part; the one,\\
Which is the last, supplying the place of male,\\
The other of the female, in all metals.\\
Some do believe hermaphrodeity,\\
That both do act and suffer. But these two\\
Make the rest ductile, malleable, extensive.\\
And even in gold they are; for we do find\\
Seeds of them, by our fire, and gold in them;\\
And can produce the species of each metal\\
More perfect thence, than nature doth in earth.\\
Beside, who doth not see in daily practice\\
Art can beget bees, hornets, beetles, wasps,\\
Out of the carcases and dung of creatures;\\
Yea, scorpions of an herb, being rightly placed?\\
And these are living creatures, far more perfect\\
And excellent than metals.\\
\mammonspeaks \vinphantom{And excellent than metals.} Well said, father!\\
Nay, if he take you in hand, sir, with an argument,\\
He'll bray you in a mortar.\\
\surlyspeaks \vinphantom{He'll bray you in a mortar.} Pray you, sir, stay.\\
Rather than I'll be brayed, sir, I'll believe\\
That Alchemy is a pretty kind of game,\\
Somewhat like tricks o' the cards, to cheat a man\\
With charming.\\
\subtlespeaks \vinphantom{With charming.} Sir?\\
\surlyspeaks \vinphantom{With charming. Sir?} What else are all your terms,\\
Whereon no one of your writers 'grees with other?\\
Of your elixir, your lac virginis,\\
Your stone, your med'cine, and your chrysosperm,\\
Your sal, your sulphur, and your mercury,\\
Your oil of height, your tree of life, your blood,\\
Your marchesite, your tutie, your magnesia,\\
Your toad, your crow, your dragon, and your panther;\\
Your sun, your moon, your firmament, your adrop,\\
Your lato, azoch, zernich, chibrit, heautarit,\\
And then your red man, and your white woman,\\
With all your broths, your menstrues, and materials,\\
Of piss and egg-shells, women's terms, man's blood,\\
Hair o' the head, burnt clouts, chalk, merds, and clay,\\
Powder of bones, scalings of iron, glass,\\
And worlds of other strange ingredients,\\
Would burst a man to name?\\
\subtlespeaks \vinphantom{Would burst a man to name?} And all these named,\\
Intending but one thing; which art our writers\\
Used to obscure their art.\\
\mammonspeaks \vinphantom{Used to obscure their art.} Sir, so I told him---\\
Because the simple idiot should not learn it,\\
And make it vulgar.\\
\subtlespeaks \vinphantom{And make it vulgar.} Was not all the knowledge\\
Of the Aegyptians writ in mystic symbols?\\
Speak not the scriptures oft in parables?\\
Are not the choicest fables of the poets,\\
That were the fountains and first springs of wisdom,\\
Wrapp'd in perplexed allegories?\\
\mammonspeaks \vinphantom{Wrapp'd in perplexed allegories?} I urg'd that,\\
And clear'd to him, that Sisyphus was damn'd\\
To roll the ceaseless stone, only because\\
He would have made Ours common.\\
\dolspeaks \direct*{APPEARS AT THE DOOR}---\\
Who is this?\\
\subtlespeaks \vinphantom{Who is this?} 'Sprecious!---What do you mean? go in, good lady,\\
Let me entreat you.\\
\direct*{DOL RETIRES.}
---Where's this varlet?\\
\direct*{RE-ENTER FACE.}
\facespeaks \vinphantom{---Where's this varlet?} Sir.\\
\subtlespeaks \vinphantom{---Where's this varlet? Sir.} You very knave! do you use me thus?\\
\facespeaks Wherein, sir?\\
\subtlespeaks \vinphantom{Wherein, sir?} Go in and see, you traitor. Go!\\
\direct*{EXIT FACE.}
\mammonspeaks Who is it, sir?\\
\subtlespeaks \vinphantom{Who is it, sir?} Nothing, sir; nothing.\\
\mammonspeaks \vinphantom{Who is it, sir? Nothing, sir; nothing.} What's the matter, good sir?\\
I have not seen you thus distemper'd: who is't?\\
\subtlespeaks All arts have still had, sir, their adversaries;\\
But ours the most ignorant.---\\
\direct*{RE-ENTER FACE.}
What now?\\
\facespeaks \vinphantom{What now?} 'Twas not my fault, sir; she would speak with you.\\
\subtlespeaks Would she, sir! Follow me.\\
\direct*{EXIT.}
\mammonspeaks \direct*{STOPPING HIM} Stay, Lungs.\\
\facespeaks \vinphantom{Would she, sir! Follow me.} I dare not, sir.\\
\mammonspeaks Stay, man; what is she?\\
\facespeaks \vinphantom{Stay, man; what is she?} A lord's sister, sir.\\
\mammonspeaks How! pray thee, stay.\\
\facespeaks \vinphantom{How! pray thee, stay.} She's mad, sir, and sent hither---\\
He'll be mad too.---\\
\mammonspeaks \vinphantom{He'll be mad too.---} I warrant thee.---\\
Why sent hither?\\
\facespeaks \vinphantom{Why sent hither?} Sir, to be cured.\\
\subtlespeaks \direct*{WITHIN} Why, rascal!\\
\facespeaks \vinphantom{Why sent hither? Sir, to be cured.} Lo you!---Here, sir!\\
\direct*{EXIT.}
\mammonspeaks 'Fore God, a Bradamante, a brave piece.\\
\surlyspeaks Heart, this is a bawdy-house! I will be burnt else.\\
\mammonspeaks O, by this light, no: do not wrong him. He's\\
Too scrupulous that way: it is his vice.\\
No, he's a rare physician, do him right,\\
An excellent Paracelsian, and has done\\
Strange cures with mineral physic. He deals all\\
With spirits, he; he will not hear a word\\
Of Galen; or his tedious recipes.---\\
\direct*{RE-ENTER FACE.}
How now, Lungs!\\
\facespeaks \vinphantom{How now, Lungs!} Softly, sir; speak softly. I meant\\
To have told your worship all. This must not hear.\\
\mammonspeaks No, he will not be "gull'd;" let him alone.\\
\facespeaks You are very right, sir, she is a most rare scholar,\\
And is gone mad with studying Broughton's works.\\
If you but name a word touching the Hebrew,\\
She falls into her fit, and will discourse\\
So learnedly of genealogies,\\
As you would run mad too, to hear her, sir.\\
\mammonspeaks How might one do t' have conference with her, Lungs?\\
\facespeaks O divers have run mad upon the conference:\\
I do not know, sir. I am sent in haste,\\
To fetch a vial.\\
\surlyspeaks \vinphantom{To fetch a vial.} Be not gull'd, sir Mammon.\\
\mammonspeaks Wherein? pray ye, be patient.\\
\surlyspeaks \vinphantom{Wherein? pray ye, be patient.} Yes, as you are,\\
And trust confederate knaves and bawds and whores.\\
\mammonspeaks You are too foul, believe it.---Come here, Ulen,\\
One word.\\
\facespeaks \vinphantom{One word.} I dare not, in good faith.\\
\direct*{GOING.}
\mammonspeaks \vinphantom{One word. I dare not, in good faith.} Stay, knave.\\
\facespeaks He is extreme angry that you saw her, sir.\\
\mammonspeaks Drink that.\\
\direct*{GIVES HIM MONEY.}
What is she when she's out of her fit?\\
\facespeaks \vinphantom{What is she when she's out of her fit?} O, the most affablest creature, sir! so merry!\\
So pleasant! she'll mount you up, like quicksilver,\\
Over the helm; and circulate like oil,\\
A very vegetal: discourse of state,\\
Of mathematics, bawdry, any thing---\\
\mammonspeaks \vinphantom{Of mathematics, bawdry, any thing---} Is she no way accessible? no means,\\
No trick to give a man a taste of her---wit---\\
Or so?\\
\subtlespeaks \direct*{WITHIN} Ulen!\\
\facespeaks \vinphantom{Or so?} I'll come to you again, sir.\\
\direct*{EXIT.}
\mammonspeaks \vinphantom{Or so? I'll come to you again, sir.} Surly, I did not think one of your breeding\\
Would traduce personages of worth.\\
\surlyspeaks Sir Epicure,\\
Your friend to use; yet still loth to be gull'd:\\
I do not like your philosophical bawds.\\
Their stone is letchery enough to pay for,\\
Without this bait.\\
\mammonspeaks \vinphantom{Without this bait.} 'Heart, you abuse yourself.\\
I know the lady, and her friends, and means,\\
The original of this disaster. Her brother\\
Has told me all.\\
\surlyspeaks \vinphantom{Has told me all.} And yet you never saw her\\
Till now!\\
\mammonspeaks \vinphantom{Till now!} O yes, but I forgot. I have, believe it,\\
One of the treacherousest memories, I do think,\\
Of all mankind.\\
\surlyspeaks \vinphantom{Of all mankind.} What call you her brother?\\
\mammonspeaks My lord---\\
He will not have his name known, now I think on't.\\
\surlyspeaks A very treacherous memory!\\
\mammonspeaks \vinphantom{A very treacherous memory!} On my faith---\\
\surlyspeaks Tut, if you have it not about you, pass it,\\
Till we meet next.\\
\mammonspeaks \vinphantom{Till we meet next.} Nay, by this hand, 'tis true.\\
He's one I honour, and my noble friend;\\
And I respect his house.\\
\surlyspeaks \vinphantom{And I respect his house.} Heart! can it be,\\
That a grave sir, a rich, that has no need,\\
A wise sir, too, at other times, should thus,\\
With his own oaths, and arguments, make hard means\\
To gull himself? An this be your elixir,\\
Your lapis mineralis, and your lunary,\\
Give me your honest trick yet at primero,\\
Or gleek; and take your lutum sapientis,\\
Your menstruum simplex! I'll have gold before you,\\
And with less danger of the quicksilver,\\
Or the hot sulphur.\\
\direct*{RE-ENTER FACE.}
\facespeaks \vinphantom{Or the hot sulphur.} Here's one from Captain Face, sir,\\
\direct*{TO SURLY.}
Desires you meet him in the Temple-church,\\
Some half-hour hence, and upon earnest business.\\
Sir,\\
\direct*{WHISPERS MAMMON.}
if you please to quit us, now; and come\\
Again within two hours, you shall have\\
My master busy examining o' the works;\\
And I will steal you in, unto the party,\\
That you may see her converse.---Sir, shall I say,\\
You'll meet the captain's worship?\\
\surlyspeaks \vinphantom{You'll meet the captain's worship?} Sir, I will.---\\
\direct*{WALKS ASIDE.}
But, by attorney, and to a second purpose.\\
Now, I am sure it is a bawdy-house;\\
I'll swear it, were the marshal here to thank me:\\
The naming this commander doth confirm it.\\
Don Face! why, he's the most authentic dealer\\
In these commodities, the superintendant\\
To all the quainter traffickers in town!\\
He is the visitor, and does appoint,\\
Who lies with whom, and at what hour; what price;\\
Which gown, and in what smock; what fall; what tire.\\
Him will I prove, by a third person, to find\\
The subtleties of this dark labyrinth:\\
Which if I do discover, dear sir Mammon,\\
You'll give your poor friend leave, though no philosopher,\\
To laugh: for you that are, 'tis thought, shall weep.\\
\facespeaks Sir, he does pray, you'll not forget.\\
\surlyspeaks \vinphantom{Sir, he does pray, you'll not forget.} I will not, sir.\\
Sir Epicure, I shall leave you.\\
\direct*{EXIT.}
\mammonspeaks \vinphantom{Sir Epicure, I shall leave you.} I follow you, straight.\\
\facespeaks But do so, good sir, to avoid suspicion.\\
This gentleman has a parlous head.\\
\mammonspeaks But wilt thou Ulen,\\
Be constant to thy promise?\\
\facespeaks \vinphantom{Be constant to thy promise?} As my life, sir.\\
\mammonspeaks And wilt thou insinuate what I am, and praise me,\\
And say, I am a noble fellow?\\
\facespeaks \vinphantom{And say, I am a noble fellow?} O, what else, sir?\\
And that you'll make her royal with the stone,\\
An empress; and yourself, King of Bantam.\\
\mammonspeaks Wilt thou do this?\\
\facespeaks \vinphantom{Wilt thou do this?} Will I, sir!\\
\mammonspeaks \vinphantom{Wilt thou do this? Will I, sir!} Lungs, my Lungs!\\
I love thee.\\
\facespeaks \vinphantom{I love thee.} Send your stuff, sir, that my master\\
May busy himself about projection.\\
\mammonspeaks Thou hast witch'd me, rogue: take, go.\\
\direct*{GIVES HIM MONEY.}
\facespeaks \vinphantom{Thou hast witch'd me, rogue: take, go.} Your jack, and all, sir.\\
\mammonspeaks Thou art a villain---I will send my jack,\\
And the weights too. Slave, I could bite thine ear.\\
Away, thou dost not care for me.\\
\facespeaks \vinphantom{Away, thou dost not care for me.} Not I, sir!\\
\mammonspeaks Come, I was born to make thee, my good weasel,\\
Set thee on a bench, and have thee twirl a chain\\
With the best lord's vermin of 'em all.\\
\facespeaks \vinphantom{With the best lord's vermin of 'em all.} Away, sir.\\
\mammonspeaks A count, nay, a count palatine---\\
\facespeaks \vinphantom{A count, nay, a count palatine---} Good, sir, go.\\
\mammonspeaks Shall not advance thee better: no, nor faster.\\
\direct*{EXIT.}
\direct*{RE-ENTER SUBTLE AND DOL.}
\subtlespeaks Has he bit? has he bit?\\
\facespeaks \vinphantom{Has he bit? has he bit?} And swallowed, too, my Subtle.\\
I have given him line, and now he plays, i'faith.\\
\subtlespeaks And shall we twitch him?\\
\facespeaks \vinphantom{And shall we twitch him?} Thorough both the gills.\\
A wench is a rare bait, with which a man\\
No sooner's taken, but he straight firks mad.\\
\subtlespeaks Dol, my Lord What'ts'hums sister, you must now\\
Bear yourself statelich.\\
\dolspeaks \vinphantom{Bear yourself statelich.} O let me alone.\\
I'll not forget my race, I warrant you.\\
I'll keep my distance, laugh and talk aloud;\\
Have all the tricks of a proud scurvy lady,\\
And be as rude as her woman.\\
\facespeaks \vinphantom{And be as rude as her woman.} Well said, sanguine!\\
\subtlespeaks But will he send his andirons?\\
\facespeaks \vinphantom{But will he send his andirons?} His jack too,\\
And's iron shoeing-horn; I have spoke to him. Well,\\
I must not lose my wary gamester yonder.\\
\subtlespeaks O monsieur Caution, that WILL NOT BE GULL'D?\\
\facespeaks Ay,\\
If I can strike a fine hook into him, now!\\
The Temple-church, there I have cast mine angle.\\
Well, pray for me. I'll about it.\\
\direct*{KNOCKING WITHOUT.}
\subtlespeaks \vinphantom{Well, pray for me. I'll about it.} What, more gudgeons!\\
Dol, scout, scout!\\
\direct*{DOL GOES TO THE WINDOW.}
Stay, Face, you must go to the door,\\
'Pray God it be my anabaptist---Who is't, Dol?\\
\dolspeaks I know him not: he looks like a gold-endman.\\
\subtlespeaks Ods so! 'tis he, he said he would send what call you him?\\
The sanctified elder, that should deal\\
For Mammon's jack and andirons. Let him in.\\
Stay, help me off, first, with my gown.\\
\direct*{EXIT FACE WITH THE GOWN.}
Away,\\
Madam, to your withdrawing chamber.\\
\direct*{EXIT DOL.}
Now,\\
In a new tune, new gesture, but old language.---\\
This fellow is sent from one negociates with me\\
About the stone too, for the holy brethren\\
Of Amsterdam, the exiled saints, that hope\\
To raise their discipline by it. I must use him\\
In some strange fashion, now, to make him admire me.---\\
\direct*{ENTER ANANIAS.}
\direct*{ALOUD.}
Where is my drudge?\\
\direct*{RE-ENTER FACE.}
\facespeaks \vinphantom{Where is my drudge?} Sir!\\
\subtlespeaks \vinphantom{Where is my drudge? Sir!} Take away the recipient,\\
And rectify your menstrue from the phlegma.\\
Then pour it on the Sol, in the cucurbite,\\
And let them macerate together.\\
\facespeaks Yes, sir.\\
And save the ground?\\
\subtlespeaks \vinphantom{And save the ground?} No: terra damnata\\
Must not have entrance in the work.---Who are you?\\
\ananiasspeaks A faithful brother, if it please you.\\
\subtlespeaks \vinphantom{A faithful brother, if it please you.} What's that?\\
A Lullianist? a Ripley? Filius artis?\\
Can you sublime and dulcify? calcine?\\
Know you the sapor pontic? sapor stiptic?\\
Or what is homogene, or heterogene?\\
\ananiasspeaks I understand no heathen language, truly.\\
\subtlespeaks Heathen! you Knipper-doling? is Ars sacra,\\
Or chrysopoeia, or spagyrica,\\
Or the pamphysic, or panarchic knowledge,\\
A heathen language?\\
\ananiasspeaks \vinphantom{A heathen language?} Heathen Greek, I take it.\\
\subtlespeaks How! heathen Greek?\\
\ananiasspeaks \vinphantom{How! heathen Greek?} All's heathen but the Hebrew.\\
\subtlespeaks Sirrah, my varlet, stand you forth and speak to him,\\
Like a philosopher: answer in the language.\\
Name the vexations, and the martyrisations\\
Of metals in the work.\\
\facespeaks \vinphantom{Of metals in the work.} Sir, putrefaction,\\
Solution, ablution, sublimation,\\
Cohobation, calcination, ceration, and\\
Fixation.\\
\subtlespeaks \vinphantom{Fixation.} This is heathen Greek to you, now!---\\
And when comes vivification?\\
\facespeaks \vinphantom{And when comes vivification?} After mortification.\\
\subtlespeaks What's cohobation?\\
\facespeaks \vinphantom{What's cohobation?} 'Tis the pouring on\\
Your aqua regis, and then drawing him off,\\
To the trine circle of the seven spheres.\\
\subtlespeaks What's the proper passion of metals?\\
\facespeaks \vinphantom{What's the proper passion of metals?} Malleation.\\
\subtlespeaks What's your ultimum supplicium auri?\\
\facespeaks Antimonium.\\
\subtlespeaks \vinphantom{Antimonium.} This is heathen Greek to you!---And what's your mercury?\\
\facespeaks A very fugitive, he will be gone, sir.\\
\subtlespeaks How know you him?\\
\facespeaks \vinphantom{How know you him?} By his viscosity,\\
His oleosity, and his suscitability.\\
\subtlespeaks How do you sublime him?\\
\facespeaks \vinphantom{How do you sublime him?} With the calce of egg-shells,\\
White marble, talc.\\
\subtlespeaks \vinphantom{White marble, talc.} Your magisterium now,\\
What's that?\\
\facespeaks \vinphantom{What's that?} Shifting, sir, your elements,\\
Dry into cold, cold into moist, moist into hot,\\
Hot into dry.\\
\subtlespeaks \vinphantom{Hot into dry.} This is heathen Greek to you still!\\
Your lapis philosophicus?\\
\facespeaks \vinphantom{Your lapis philosophicus?} 'Tis a stone,\\
And not a stone; a spirit, a soul, and a body:\\
Which if you do dissolve, it is dissolved;\\
If you coagulate, it is coagulated;\\
If you make it to fly, it flieth.\\
\subtlespeaks \vinphantom{If you make it to fly, it flieth.} Enough.\\
\direct*{EXIT FACE.}
This is heathen Greek to you! What are you, sir?\\
\ananiasspeaks Please you, a servant of the exiled brethren,\\
That deal with widows' and with orphans' goods,\\
And make a just account unto the saints:\\
A deacon.\\
\subtlespeaks \vinphantom{A deacon.} O, you are sent from master Wholesome,\\
Your teacher?\\
\ananiasspeaks \vinphantom{Your teacher?} From Tribulation Wholesome,\\
Our very zealous pastor.\\
\subtlespeaks \vinphantom{Our very zealous pastor.} Good! I have\\
Some orphans' goods to come here.\\
\ananiasspeaks \vinphantom{Some orphans' goods to come here.} Of what kind, sir?\\
\subtlespeaks Pewter and brass, andirons and kitchen-ware,\\
Metals, that we must use our medicine on:\\
Wherein the brethren may have a pennyworth\\
For ready money.\\
\ananiasspeaks \vinphantom{For ready money.} Were the orphans' parents\\
Sincere professors?\\
\subtlespeaks \vinphantom{Sincere professors?} Why do you ask?\\
\ananiasspeaks \vinphantom{Sincere professors? Why do you ask?} Because\\
We then are to deal justly, and give, in truth,\\
Their utmost value.\\
\subtlespeaks \vinphantom{Their utmost value.} 'Slid, you'd cozen else,\\
And if their parents were not of the faithful!---\\
I will not trust you, now I think on it,\\
'Till I have talked with your pastor. Have you brought money\\
To buy more coals?\\
\ananiasspeaks \vinphantom{To buy more coals?} No, surely.\\
\subtlespeaks \vinphantom{To buy more coals? No, surely.} No! how so?\\
\ananiasspeaks The brethren bid me say unto you, sir,\\
Surely, they will not venture any more,\\
Till they may see projection.\\
\subtlespeaks \vinphantom{Till they may see projection.} How!\\
\ananiasspeaks \vinphantom{Till they may see projection. How!} You have had,\\
For the instruments, as bricks, and lome, and glasses,\\
Already thirty pound; and for materials,\\
They say, some ninety more: and they have heard since,\\
That one at Heidelberg, made it of an egg,\\
And a small paper of pin-dust.\\
\subtlespeaks \vinphantom{And a small paper of pin-dust.} What's your name?\\
\ananiasspeaks My name is Ananias.\\
\subtlespeaks \vinphantom{My name is Ananias.} Out, the varlet\\
That cozen'd the apostles! Hence, away!\\
Flee, mischief! had your holy consistory\\
No name to send me, of another sound,\\
Than wicked Ananias? send your elders\\
Hither to make atonement for you quickly,\\
And give me satisfaction; or out goes\\
The fire; and down th' alembics, and the furnace,\\
Piger Henricus, or what not. Thou wretch!\\
Both sericon and bufo shall be lost,\\
Tell them. All hope of rooting out the bishops,\\
Or the antichristian hierarchy, shall perish,\\
If they stay threescore minutes: the aqueity,\\
Terreity, and sulphureity\\
Shall run together again, and all be annull'd,\\
Thou wicked Ananias!\\
\direct*{EXIT ANANIAS.}
This will fetch 'em,\\
And make them haste towards their gulling more.\\
A man must deal like a rough nurse, and fright\\
Those that are froward, to an appetite.\\
\direct*{RE-ENTER FACE, IN HIS UNIFORM, FOLLOWED BY DRUGGER.}
\facespeaks He is busy with his spirits, but we'll upon him.\\
\subtlespeaks How now! what mates, what Baiards have we here?\\
\facespeaks I told you, he would be furious.---Sir, here's Nab,\\
Has brought you another piece of gold to look on:\\
---We must appease him. Give it me,---and prays you,\\
You would devise---what is it, Nab?\\
\druggerspeaks \vinphantom{You would devise---what is it, Nab?} A sign, sir.\\
\facespeaks Ay, a good lucky one, a thriving sign, doctor.\\
\subtlespeaks I was devising now.\\
\facespeaks \vinphantom{I was devising now.} 'Slight, do not say so,\\
He will repent he gave you any more---\\
What say you to his constellation, doctor,\\
The Balance?\\
\subtlespeaks \vinphantom{The Balance?} No, that way is stale, and common.\\
A townsman born in Taurus, gives the bull,\\
Or the bull's-head: in Aries, the ram,\\
A poor device! No, I will have his name\\
Form'd in some mystic character; whose radii,\\
Striking the senses of the passers by,\\
Shall, by a virtual influence, breed affections,\\
That may result upon the party owns it:\\
As thus---\\
\facespeaks \vinphantom{As thus---} Nab!\\
\subtlespeaks \vinphantom{As thus--- Nab!} He shall have "a bell," that's "Abel;"\\
And by it standing one whose name is "Dee,"\\
In a "rug" gown, there's "D," and "Rug," that's "drug:"\\
And right anenst him a dog snarling "er;"\\
There's "Drugger," Abel Drugger. That's his sign.\\
And here's now mystery and hieroglyphic!\\
\facespeaks Abel, thou art made.\\
\druggerspeaks \vinphantom{Abel, thou art made.} Sir, I do thank his worship.\\
\facespeaks Six o' thy legs more will not do it, Nab.\\
He has brought you a pipe of tobacco, doctor.\\
\druggerspeaks Yes, sir;\\
I have another thing I would impart---\\
\facespeaks Out with it, Nab.\\
\druggerspeaks \vinphantom{Out with it, Nab.} Sir, there is lodged, hard by me,\\
A rich young widow---\\
\facespeaks \vinphantom{A rich young widow---} Good! a bona roba?\\
\druggerspeaks But nineteen, at the most.\\
\facespeaks \vinphantom{But nineteen, at the most.} Very good, Abel.\\
\druggerspeaks Marry, she's not in fashion yet; she wears\\
A hood, but it stands a cop.\\
\facespeaks \vinphantom{A hood, but it stands a cop.} No matter, Abel.\\
\druggerspeaks And I do now and then give her a fucus---\\
\facespeaks What! dost thou deal, Nab?\\
\subtlespeaks \vinphantom{What! dost thou deal, Nab?} I did tell you, captain.\\
\druggerspeaks And physic too, sometime, sir; for which she trusts me\\
With all her mind. She's come up here of purpose\\
To learn the fashion.\\
\facespeaks \vinphantom{To learn the fashion.} Good (his match too!)---On, Nab.\\
\druggerspeaks And she does strangely long to know her fortune.\\
\facespeaks Ods lid, Nab, send her to the doctor, hither.\\
\druggerspeaks Yes, I have spoke to her of his worship already;\\
But she's afraid it will be blown abroad,\\
And hurt her marriage.\\
\facespeaks \vinphantom{And hurt her marriage.} Hurt it! 'tis the way\\
To heal it, if 'twere hurt; to make it more\\
Follow'd and sought: Nab, thou shalt tell her this.\\
She'll be more known, more talk'd of; and your widows\\
Are ne'er of any price till they be famous;\\
Their honour is their multitude of suitors.\\
Send her, it may be thy good fortune. What!\\
Thou dost not know.\\
\druggerspeaks \vinphantom{Thou dost not know.} No, sir, she'll never marry\\
Under a knight: her brother has made a vow.\\
\facespeaks What! and dost thou despair, my little Nab,\\
Knowing what the doctor has set down for thee,\\
And seeing so many of the city dubb'd?\\
One glass o' thy water, with a madam I know,\\
Will have it done, Nab: what's her brother, a knight?\\
\druggerspeaks No, sir, a gentleman newly warm in his land, sir,\\
Scarce cold in his one and twenty, that does govern\\
His sister here; and is a man himself\\
Of some three thousand a year, and is come up\\
To learn to quarrel, and to live by his wits,\\
And will go down again, and die in the country.\\
\facespeaks How! to quarrel?\\
\druggerspeaks \vinphantom{How! to quarrel?} Yes, sir, to carry quarrels,\\
As gallants do; to manage them by line.\\
\facespeaks 'Slid, Nab, the doctor is the only man\\
In Christendom for him. He has made a table,\\
With mathematical demonstrations,\\
Touching the art of quarrels: he will give him\\
An instrument to quarrel by. Go, bring them both,\\
Him and his sister. And, for thee, with her\\
The doctor happ'ly may persuade. Go to:\\
'Shalt give his worship a new damask suit\\
Upon the premises.\\
\subtlespeaks \vinphantom{Upon the premises.} O, good captain!\\
\facespeaks He shall;\\
He is the honestest fellow, doctor.---Stay not,\\
No offers; bring the damask, and the parties.\\
\druggerspeaks I'll try my power, sir.\\
\facespeaks \vinphantom{I'll try my power, sir.} And thy will too, Nab.\\
\subtlespeaks 'Tis good tobacco, this! What is't an ounce?\\
\facespeaks He'll send you a pound, doctor.\\
\subtlespeaks \vinphantom{He'll send you a pound, doctor.} O no.\\
\facespeaks \vinphantom{He'll send you a pound, doctor. O no.} He will do't.\\
It is the goodest soul!---Abel, about it.\\
Thou shalt know more anon. Away, be gone.\\
\direct*{EXIT ABEL.}
A miserable rogue, and lives with cheese,\\
And has the worms. That was the cause, indeed,\\
Why he came now: he dealt with me in private,\\
To get a med'cine for them.\\
\subtlespeaks \vinphantom{To get a med'cine for them.} And shall, sir. This works.\\
\facespeaks A wife, a wife for one on us, my dear Subtle!\\
We'll e'en draw lots, and he that fails, shall have\\
The more in goods, the other has in tail.\\
\subtlespeaks Rather the less: for she may be so light\\
She may want grains.\\
\facespeaks \vinphantom{She may want grains.} Ay, or be such a burden,\\
A man would scarce endure her for the whole.\\
\subtlespeaks Faith, best let's see her first, and then determine.\\
\facespeaks Content: but Dol must have no breath on't.\\
\subtlespeaks \vinphantom{Content: but Dol must have no breath on't.} Mum.\\
Away you, to your Surly yonder, catch him.\\
\facespeaks 'Pray God I have not staid too long.\\
\subtlespeaks \vinphantom{'Pray God I have not staid too long.} I fear it.\\
\direct*{EXEUNT.}
\act

\scene

\stagedir{THE LANE BEFORE LOVEWIT'S HOUSE. ENTER TRIBULATION WHOLESOME AND ANANIAS.}
\tribulationspeaks These chastisements are common to the saints,\\
And such rebukes, we of the separation\\
Must bear with willing shoulders, as the trials\\
Sent forth to tempt our frailties.\\
\ananiasspeaks \vinphantom{Sent forth to tempt our frailties.} In pure zeal,\\
I do not like the man; he is a heathen,\\
And speaks the language of Canaan, truly.\\
\tribulationspeaks I think him a profane person indeed.\\
\ananiasspeaks He bears\\
The visible mark of the beast in his forehead.\\
And for his stone, it is a work of darkness,\\
And with philosophy blinds the eyes of man.\\
\tribulationspeaks Good brother, we must bend unto all means,\\
That may give furtherance to the holy cause.\\
\ananiasspeaks Which his cannot: the sanctified cause\\
Should have a sanctified course.\\
\tribulationspeaks \vinphantom{Should have a sanctified course.} Not always necessary:\\
The children of perdition are oft-times\\
Made instruments even of the greatest works:\\
Beside, we should give somewhat to man's nature,\\
The place he lives in, still about the fire,\\
And fume of metals, that intoxicate\\
The brain of man, and make him prone to passion.\\
Where have you greater atheists than your cooks?\\
Or more profane, or choleric, than your glass-men?\\
More antichristian than your bell-founders?\\
What makes the devil so devilish, I would ask you,\\
Sathan, our common enemy, but his being\\
Perpetually about the fire, and boiling\\
Brimstone and arsenic? We must give, I say,\\
Unto the motives, and the stirrers up\\
Of humours in the blood. It may be so,\\
When as the work is done, the stone is made,\\
This heat of his may turn into a zeal,\\
And stand up for the beauteous discipline,\\
Against the menstruous cloth and rag of Rome.\\
We must await his calling, and the coming\\
Of the good spirit. You did fault, t' upbraid him\\
With the brethren's blessing of Heidelberg, weighing\\
What need we have to hasten on the work,\\
For the restoring of the silenced saints,\\
Which ne'er will be, but by the philosopher's stone.\\
And so a learned elder, one of Scotland,\\
Assured me; aurum potabile being\\
The only med'cine, for the civil magistrate,\\
T' incline him to a feeling of the cause;\\
And must be daily used in the disease.\\
\ananiasspeaks I have not edified more, truly, by man;\\
Not since the beautiful light first shone on me:\\
And I am sad my zeal hath so offended.\\
\tribulationspeaks Let us call on him then.\\
\ananiasspeaks \vinphantom{Let us call on him then.} The motion's good,\\
And of the spirit; I will knock first.\\
\direct*{KNOCKS.}
Peace be within!\\
\direct*{THE DOOR IS OPENED, AND THEY ENTER.}
\scene

\stagedir{A ROOM IN LOVEWIT'S HOUSE. ENTER SUBTLE, FOLLOWED BY TRIBULATION AND ANANIAS.}
\subtlespeaks \vinphantom{Peace be within!} O, are you come? 'twas time. Your threescore minutes\\
Were at last thread, you see: and down had gone\\
Furnus acediae, turris circulatorius:\\
Lembec, bolt's-head, retort and pelican\\
Had all been cinders.---Wicked Ananias!\\
Art thou return'd? nay then, it goes down yet.\\
\tribulationspeaks Sir, be appeased; he is come to humble\\
Himself in spirit, and to ask your patience,\\
If too much zeal hath carried him aside\\
From the due path.\\
\subtlespeaks \vinphantom{From the due path.} Why, this doth qualify!\\
\tribulationspeaks The brethren had no purpose, verily,\\
To give you the least grievance; but are ready\\
To lend their willing hands to any project\\
The spirit and you direct.\\
\subtlespeaks \vinphantom{The spirit and you direct.} This qualifies more!\\
\tribulationspeaks And for the orphans' goods, let them be valued,\\
Or what is needful else to the holy work,\\
It shall be numbered; here, by me, the saints,\\
Throw down their purse before you.\\
\subtlespeaks \vinphantom{Throw down their purse before you.} This qualifies most!\\
Why, thus it should be, now you understand.\\
Have I discours'd so unto you of our stone,\\
And of the good that it shall bring your cause?\\
Shew'd you (beside the main of hiring forces\\
Abroad, drawing the Hollanders, your friends,\\
From the Indies, to serve you, with all their fleet)\\
That even the med'cinal use shall make you a faction,\\
And party in the realm? As, put the case,\\
That some great man in state, he have the gout,\\
Why, you but send three drops of your elixir,\\
You help him straight: there you have made a friend.\\
Another has the palsy or the dropsy,\\
He takes of your incombustible stuff,\\
He's young again: there you have made a friend,\\
A lady that is past the feat of body,\\
Though not of mind, and hath her face decay'd\\
Beyond all cure of paintings, you restore,\\
With the oil of talc: there you have made a friend;\\
And all her friends. A lord that is a leper,\\
A knight that has the bone-ache, or a squire\\
That hath both these, you make them smooth and sound,\\
With a bare fricace of your med'cine: still\\
You increase your friends.\\
\tribulationspeaks \vinphantom{You increase your friends.} Ay, it is very pregnant.\\
\subtlespeaks And then the turning of this lawyer's pewter\\
To plate at Christmas.---\\
\ananiasspeaks \vinphantom{To plate at Christmas.---} Christ-tide, I pray you.\\
\subtlespeaks Yet, Ananias!\\
\ananiasspeaks \vinphantom{Yet, Ananias!} I have done.\\
\subtlespeaks \vinphantom{Yet, Ananias! I have done.} Or changing\\
His parcel gilt to massy gold. You cannot\\
But raise you friends. Withal, to be of power\\
To pay an army in the field, to buy\\
The king of France out of his realms, or Spain\\
Out of his Indies. What can you not do\\
Against lords spiritual or temporal,\\
That shall oppone you?\\
\tribulationspeaks \vinphantom{That shall oppone you?} Verily, 'tis true.\\
We may be temporal lords ourselves, I take it.\\
\subtlespeaks You may be any thing, and leave off to make\\
Long-winded exercises; or suck up\\
Your "ha!" and "hum!" in a tune. I not deny,\\
But such as are not graced in a state,\\
May, for their ends, be adverse in religion,\\
And get a tune to call the flock together:\\
For, to say sooth, a tune does much with women,\\
And other phlegmatic people; it is your bell.\\
\ananiasspeaks Bells are profane; a tune may be religious.\\
\subtlespeaks No warning with you! then farewell my patience.\\
'Slight, it shall down: I will not be thus tortured.\\
\tribulationspeaks I pray you, sir.\\
\subtlespeaks \vinphantom{I pray you, sir.} All shall perish. I have spoken it.\\
\tribulationspeaks Let me find grace, sir, in your eyes; the man\\
He stands corrected: neither did his zeal,\\
But as your self, allow a tune somewhere.\\
Which now, being tow'rd the stone, we shall not need.\\
\subtlespeaks No, nor your holy vizard, to win widows\\
To give you legacies; or make zealous wives\\
To rob their husbands for the common cause:\\
Nor take the start of bonds broke but one day,\\
And say, they were forfeited by providence.\\
Nor shall you need o'er night to eat huge meals,\\
To celebrate your next day's fast the better;\\
The whilst the brethren and the sisters humbled,\\
Abate the stiffness of the flesh. Nor cast\\
Before your hungry hearers scrupulous bones;\\
As whether a Christian may hawk or hunt,\\
Or whether matrons of the holy assembly\\
May lay their hair out, or wear doublets,\\
Or have that idol starch about their linen.\\
\ananiasspeaks It is indeed an idol.\\
\tribulationspeaks \vinphantom{It is indeed an idol.} Mind him not, sir.\\
I do command thee, spirit of zeal, but trouble,\\
To peace within him! Pray you, sir, go on.\\
\subtlespeaks Nor shall you need to libel 'gainst the prelates,\\
And shorten so your ears against the hearing\\
Of the next wire-drawn grace. Nor of necessity\\
Rail against plays, to please the alderman\\
Whose daily custard you devour; nor lie\\
With zealous rage till you are hoarse. Not one\\
Of these so singular arts. Nor call yourselves\\
By names of Tribulation, Persecution,\\
Restraint, Long-patience, and such-like, affected\\
By the whole family or wood of you,\\
Only for glory, and to catch the ear\\
Of the disciple.\\
\tribulationspeaks \vinphantom{Of the disciple.} Truly, sir, they are\\
Ways that the godly brethren have invented,\\
For propagation of the glorious cause,\\
As very notable means, and whereby also\\
Themselves grow soon, and profitably, famous.\\
\subtlespeaks O, but the stone, all's idle to it! nothing!\\
The art of angels' nature's miracle,\\
The divine secret that doth fly in clouds\\
From east to west: and whose tradition\\
Is not from men, but spirits.\\
\ananiasspeaks \vinphantom{Is not from men, but spirits.} I hate traditions;\\
I do not trust them---\\
\tribulationspeaks \vinphantom{I do not trust them---} Peace!\\
\ananiasspeaks \vinphantom{I do not trust them--- Peace!} They are popish all.\\
I will not peace: I will not---\\
\tribulationspeaks \vinphantom{I will not peace: I will not---} Ananias!\\
\ananiasspeaks Please the profane, to grieve the godly; I may not.\\
\subtlespeaks Well, Ananias, thou shalt overcome.\\
\tribulationspeaks It is an ignorant zeal that haunts him, sir;\\
But truly, else, a very faithful brother,\\
A botcher, and a man, by revelation,\\
That hath a competent knowledge of the truth.\\
\subtlespeaks Has he a competent sum there in the bag\\
To buy the goods within? I am made guardian,\\
And must, for charity, and conscience sake,\\
Now see the most be made for my poor orphan;\\
Though I desire the brethren too good gainers:\\
There they are within. When you have view'd and bought 'em,\\
And ta'en the inventory of what they are,\\
They are ready for projection; there's no more\\
To do: cast on the med'cine, so much silver\\
As there is tin there, so much gold as brass,\\
I'll give't you in by weight.\\
\tribulationspeaks \vinphantom{I'll give't you in by weight.} But how long time,\\
Sir, must the saints expect yet?\\
\subtlespeaks \vinphantom{Sir, must the saints expect yet?} Let me see,\\
How's the moon now? Eight, nine, ten days hence,\\
He will be silver potate; then three days\\
Before he citronise: Some fifteen days,\\
The magisterium will be perfected.\\
\ananiasspeaks About the second day of the third week,\\
In the ninth month?\\
\subtlespeaks \vinphantom{In the ninth month?} Yes, my good Ananias.\\
\tribulationspeaks What will the orphan's goods arise to, think you?\\
\subtlespeaks Some hundred marks, as much as fill'd three cars,\\
Unladed now: you'll make six millions of them.---\\
But I must have more coals laid in.\\
\tribulationspeaks \vinphantom{But I must have more coals laid in.} How?\\
\subtlespeaks \vinphantom{But I must have more coals laid in. How?} Another load,\\
And then we have finish'd. We must now increase\\
Our fire to ignis ardens; we are past\\
Fimus equinus, balnei, cineris,\\
And all those lenter heats. If the holy purse\\
Should with this draught fall low, and that the saints\\
Do need a present sum, I have a trick\\
To melt the pewter, you shall buy now, instantly,\\
And with a tincture make you as good Dutch dollars\\
As any are in Holland.\\
\tribulationspeaks \vinphantom{As any are in Holland.} Can you so?\\
\subtlespeaks Ay, and shall 'bide the third examination.\\
\ananiasspeaks It will be joyful tidings to the brethren.\\
\subtlespeaks But you must carry it secret.\\
\tribulationspeaks \vinphantom{But you must carry it secret.} Ay; but stay,\\
This act of coining, is it lawful?\\
\ananiasspeaks \vinphantom{This act of coining, is it lawful?} Lawful!\\
We know no magistrate; or, if we did,\\
This is foreign coin.\\
\subtlespeaks \vinphantom{This is foreign coin.} It is no coining, sir.\\
It is but casting.\\
\tribulationspeaks \vinphantom{It is but casting.} Ha! you distinguish well:\\
Casting of money may be lawful.\\
\ananiasspeaks \vinphantom{Casting of money may be lawful.} 'Tis, sir.\\
\tribulationspeaks Truly, I take it so.\\
\subtlespeaks \vinphantom{Truly, I take it so.} There is no scruple,\\
Sir, to be made of it; believe Ananias:\\
This case of conscience he is studied in.\\
\tribulationspeaks I'll make a question of it to the brethren.\\
\ananiasspeaks The brethren shall approve it lawful, doubt not.\\
Where shall it be done?\\
\direct*{KNOCKING WITHOUT.}
\subtlespeaks \vinphantom{Where shall it be done?} For that we'll talk anon.\\
There's some to speak with me. Go in, I pray you,\\
And view the parcels. That's the inventory.\\
I'll come to you straight.\\
\direct*{EXEUNT TRIB. AND ANA.}
Who is it?---Face! appear.\\
\direct*{ENTER FACE IN HIS UNIFORM.}
How now! good prize?\\
\facespeaks \vinphantom{How now! good prize?} Good pox! yond' costive cheater\\
Never came on.\\
\subtlespeaks \vinphantom{Never came on.} How then?\\
\facespeaks \vinphantom{Never came on. How then?} I have walk'd the round\\
Till now, and no such thing.\\
\subtlespeaks \vinphantom{Till now, and no such thing.} And have you quit him?\\
\facespeaks Quit him! an hell would quit him too, he were happy.\\
'Slight! would you have me stalk like a mill-jade,\\
All day, for one that will not yield us grains?\\
I know him of old.\\
\subtlespeaks \vinphantom{I know him of old.} O, but to have gull'd him,\\
Had been a mastery.\\
\facespeaks \vinphantom{Had been a mastery.} Let him go, black boy!\\
And turn thee, that some fresh news may possess thee.\\
A noble count, a don of Spain, my dear\\
Delicious compeer, and my party-bawd,\\
Who is come hither private for his conscience,\\
And brought munition with him, six great slops,\\
Bigger than three Dutch hoys, beside round trunks,\\
Furnished with pistolets, and pieces of eight,\\
Will straight be here, my rogue, to have thy bath,\\
(That is the colour,) and to make his battery\\
Upon our Dol, our castle, our cinque-port,\\
Our Dover pier, our what thou wilt. Where is she?\\
She must prepare perfumes, delicate linen,\\
The bath in chief, a banquet, and her wit,\\
For she must milk his epididimis.\\
Where is the doxy?\\
\subtlespeaks \vinphantom{Where is the doxy?} I'll send her to thee:\\
And but despatch my brace of little John Leydens,\\
And come again my self.\\
\facespeaks \vinphantom{And come again my self.} Are they within then?\\
\subtlespeaks Numbering the sum.\\
\facespeaks \vinphantom{Numbering the sum.} How much?\\
\subtlespeaks \vinphantom{Numbering the sum. How much?} A hundred marks, boy.\\
\direct*{EXIT.}
\facespeaks Why, this is a lucky day. Ten pounds of Mammon!\\
Three of my clerk! A portague of my grocer!\\
This of the brethren! beside reversions,\\
And states to come in the widow, and my count!\\
My share to-day will not be bought for forty---\\
\direct*{ENTER DOL.}
\dolspeaks What?\\
\facespeaks \vinphantom{What?} Pounds, dainty Dorothy! art thou so near?\\
\dolspeaks Yes; say, lord general, how fares our camp?\\
\facespeaks As with the few that had entrench'd themselves\\
Safe, by their discipline, against a world, Dol,\\
And laugh'd within those trenches, and grew fat\\
With thinking on the booties, Dol, brought in\\
Daily by their small parties. This dear hour,\\
A doughty don is taken with my Dol;\\
And thou mayst make his ransom what thou wilt,\\
My Dousabel; he shall be brought here fetter'd\\
With thy fair looks, before he sees thee; and thrown\\
In a down-bed, as dark as any dungeon;\\
Where thou shalt keep him waking with thy drum;\\
Thy drum, my Dol, thy drum; till he be tame\\
As the poor black-birds were in the great frost,\\
Or bees are with a bason; and so hive him\\
In the swan-skin coverlid, and cambric sheets,\\
Till he work honey and wax, my little God's-gift.\\
\dolspeaks What is he, general?\\
\facespeaks \vinphantom{What is he, general?} An adalantado,\\
A grandee, girl. Was not my Dapper here yet?\\
\dolspeaks No.\\
\facespeaks \vinphantom{No.} Nor my Drugger?\\
\dolspeaks \vinphantom{No. Nor my Drugger?} Neither.\\
\facespeaks \vinphantom{No. Nor my Drugger? Neither.} A pox on 'em,\\
They are so long a furnishing! such stinkards\\
Would not be seen upon these festival days.---\\
\direct*{RE-ENTER SUBTLE.}
How now! have you done?\\
\subtlespeaks \vinphantom{How now! have you done?} Done. They are gone: the sum\\
Is here in bank, my Face. I would we knew\\
Another chapman now would buy 'em outright.\\
\facespeaks 'Slid, Nab shall do't against he have the widow,\\
To furnish household.\\
\subtlespeaks \vinphantom{To furnish household.} Excellent, well thought on:\\
Pray God he come!\\
\facespeaks \vinphantom{Pray God he come!} I pray he keep away\\
Till our new business be o'erpast.\\
\subtlespeaks But, Face,\\
How cam'st thou by this secret don?\\
\facespeaks \vinphantom{How cam'st thou by this secret don?} A spirit\\
Brought me th' intelligence in a paper here,\\
As I was conjuring yonder in my circle\\
For Surly; I have my flies abroad. Your bath\\
Is famous, Subtle, by my means. Sweet Dol,\\
You must go tune your virginal, no losing\\
O' the least time: and, do you hear? good action.\\
Firk, like a flounder; kiss, like a scallop, close;\\
And tickle him with thy mother tongue. His great\\
Verdugoship has not a jot of language;\\
So much the easier to be cozen'd, my Dolly.\\
He will come here in a hired coach, obscure,\\
And our own coachman, whom I have sent as guide,\\
No creature else.\\
\direct*{KNOCKING WITHOUT.}
Who's that?\\
\direct*{EXIT DOL.}
\subtlespeaks \vinphantom{Who's that?} It is not he?\\
\facespeaks \vinphantom{Who's that? It is not he?} O no, not yet this hour.\\
\direct*{RE-ENTER DOL.}
\subtlespeaks Who is't?\\
\dolspeaks \vinphantom{Who is't?} Dapper,\\
Your clerk.\\
\facespeaks \vinphantom{Your clerk.} God's will then, queen of Fairy,\\
On with your tire;\\
\direct*{EXIT DOL.}
and, doctor, with your robes.\\
Let's dispatch him for God's sake.\\
\subtlespeaks \vinphantom{Let's dispatch him for God's sake.} 'Twill be long.\\
\facespeaks I warrant you, take but the cues I give you,\\
It shall be brief enough.\\
\direct*{GOES TO THE WINDOW.}
'Slight, here are more!\\
Abel, and I think the angry boy, the heir,\\
That fain would quarrel.\\
\subtlespeaks \vinphantom{That fain would quarrel.} And the widow?\\
\facespeaks \vinphantom{That fain would quarrel. And the widow?} No,\\
Not that I see. Away!\\
\direct*{EXIT SUB.}
\direct*{ENTER DAPPER.}
O sir, you are welcome.\\
The doctor is within a moving for you;\\
I have had the most ado to win him to it!---\\
He swears you'll be the darling of the dice:\\
He never heard her highness dote till now.\\
Your aunt has given you the most gracious words\\
That can be thought on.\\
\dapperspeaks \vinphantom{That can be thought on.} Shall I see her grace?\\
\facespeaks See her, and kiss her too.---\\
\direct*{ENTER ABEL, FOLLOWED BY KASTRIL.}
What, honest Nab!\\
Hast brought the damask?\\
\druggerspeaks \vinphantom{Hast brought the damask?} No, sir; here's tobacco.\\
\facespeaks 'Tis well done, Nab; thou'lt bring the damask too?\\
\druggerspeaks Yes: here's the gentleman, captain, master Kastril,\\
I have brought to see the doctor.\\
\facespeaks \vinphantom{I have brought to see the doctor.} Where's the widow?\\
\druggerspeaks Sir, as he likes, his sister, he says, shall come.\\
\facespeaks O, is it so? good time. Is your name Kastril, sir?\\
\kastrilspeaks Ay, and the best of the Kastrils, I'd be sorry else,\\
By fifteen hundred a year. Where is the doctor?\\
My mad tobacco-boy, here, tells me of one\\
That can do things: has he any skill?\\
\facespeaks \vinphantom{That can do things: has he any skill?} Wherein, sir?\\
\kastrilspeaks To carry a business, manage a quarrel fairly,\\
Upon fit terms.\\
\facespeaks \vinphantom{Upon fit terms.} It seems, sir, you are but young\\
About the town, that can make that a question.\\
\kastrilspeaks Sir, not so young, but I have heard some speech\\
Of the angry boys, and seen them take tobacco;\\
And in his shop; and I can take it too.\\
And I would fain be one of 'em, and go down\\
And practise in the country.\\
\facespeaks \vinphantom{And practise in the country.} Sir, for the duello,\\
The doctor, I assure you, shall inform you,\\
To the least shadow of a hair; and shew you\\
An instrument he has of his own making,\\
Wherewith no sooner shall you make report\\
Of any quarrel, but he will take the height on't\\
Most instantly, and tell in what degree\\
Of safety it lies in, or mortality.\\
And how it may be borne, whether in a right line,\\
Or a half circle; or may else be cast\\
Into an angle blunt, if not acute:\\
And this he will demonstrate. And then, rules\\
To give and take the lie by.\\
\kastrilspeaks \vinphantom{To give and take the lie by.} How! to take it?\\
\facespeaks Yes, in oblique he'll shew you, or in circle;\\
But never in diameter. The whole town\\
Study his theorems, and dispute them ordinarily\\
At the eating academies.\\
\kastrilspeaks \vinphantom{At the eating academies.} But does he teach\\
Living by the wits too?\\
\facespeaks \vinphantom{Living by the wits too?} Anything whatever.\\
You cannot think that subtlety, but he reads it.\\
He made me a captain. I was a stark pimp,\\
Just of your standing, 'fore I met with him;\\
It is not two months since. I'll tell you his method:\\
First, he will enter you at some ordinary.\\
\kastrilspeaks No, I'll not come there: you shall pardon me.\\
\facespeaks For why, sir?\\
\kastrilspeaks \vinphantom{For why, sir?} There's gaming there, and tricks.\\
\facespeaks \vinphantom{For why, sir? There's gaming there, and tricks.} Why, would you be\\
A gallant, and not game?\\
\kastrilspeaks \vinphantom{A gallant, and not game?} Ay, 'twill spend a man.\\
\facespeaks Spend you! it will repair you when you are spent:\\
How do they live by their wits there, that have vented\\
Six times your fortunes?\\
\kastrilspeaks \vinphantom{Six times your fortunes?} What, three thousand a-year!\\
\facespeaks Ay, forty thousand.\\
\kastrilspeaks \vinphantom{Ay, forty thousand.} Are there such?\\
\facespeaks \vinphantom{Ay, forty thousand. Are there such?} Ay, sir,\\
And gallants yet. Here's a young gentleman\\
Is born to nothing,---\\
\direct*{POINTS TO DAPPER.}
forty marks a year,\\
Which I count nothing:---he is to be initiated,\\
And have a fly of the doctor. He will win you,\\
By unresistible luck, within this fortnight,\\
Enough to buy a barony. They will set him\\
Upmost, at the groom porter's, all the Christmas:\\
And for the whole year through, at every place,\\
Where there is play, present him with the chair;\\
The best attendance, the best drink; sometimes\\
Two glasses of Canary, and pay nothing;\\
The purest linen, and the sharpest knife,\\
The partridge next his trencher: and somewhere\\
The dainty bed, in private, with the dainty.\\
You shall have your ordinaries bid for him,\\
As play-houses for a poet; and the master\\
Pray him aloud to name what dish he affects,\\
Which must be butter'd shrimps: and those that drink\\
To no mouth else, will drink to his, as being\\
The goodly president mouth of all the board.\\
\kastrilspeaks Do you not gull one?\\
\facespeaks \vinphantom{Do you not gull one?} 'Ods my life! do you think it?\\
You shall have a cast commander, (can but get\\
In credit with a glover, or a spurrier,\\
For some two pair of either's ware aforehand,)\\
Will, by most swift posts, dealing \direct*{but} with him,\\
Arrive at competent means to keep himself,\\
His punk and naked boy, in excellent fashion,\\
And be admired for't.\\
\kastrilspeaks \vinphantom{And be admired for't.} Will the doctor teach this?\\
\facespeaks He will do more, sir: when your land is gone,\\
As men of spirit hate to keep earth long,\\
In a vacation, when small money is stirring,\\
And ordinaries suspended till the term,\\
He'll shew a perspective, where on one side\\
You shall behold the faces and the persons\\
Of all sufficient young heirs in town,\\
Whose bonds are current for commodity;\\
On th' other side, the merchants' forms, and others,\\
That without help of any second broker,\\
Who would expect a share, will trust such parcels:\\
In the third square, the very street and sign\\
Where the commodity dwells, and does but wait\\
To be deliver'd, be it pepper, soap,\\
Hops, or tobacco, oatmeal, woad, or cheeses.\\
All which you may so handle, to enjoy\\
To your own use, and never stand obliged.\\
\kastrilspeaks I'faith! is he such a fellow?\\
\facespeaks \vinphantom{I'faith! is he such a fellow?} Why, Nab here knows him.\\
And then for making matches for rich widows,\\
Young gentlewomen, heirs, the fortunat'st man!\\
He's sent to, far and near, all over England,\\
To have his counsel, and to know their fortunes.\\
\kastrilspeaks God's will, my suster shall see him.\\
\facespeaks \vinphantom{God's will, my suster shall see him.} I'll tell you, sir,\\
What he did tell me of Nab. It's a strange thing:---\\
By the way, you must eat no cheese, Nab, it breeds melancholy,\\
And that same melancholy breeds worms; but pass it:---\\
He told me, honest Nab here was ne'er at tavern\\
But once in's life!\\
\druggerspeaks \vinphantom{But once in's life!} Truth, and no more I was not.\\
\facespeaks And then he was so sick---\\
\druggerspeaks \vinphantom{And then he was so sick---} Could he tell you that too?\\
\facespeaks How should I know it?\\
\druggerspeaks \vinphantom{How should I know it?} In troth we had been a shooting,\\
And had a piece of fat ram-mutton to supper,\\
That lay so heavy o' my stomach---\\
\facespeaks \vinphantom{That lay so heavy o' my stomach---} And he has no head\\
To bear any wine; for what with the noise of the fidlers,\\
And care of his shop, for he dares keep no servants---\\
\druggerspeaks My head did so ach---\\
\facespeaks \vinphantom{My head did so ach---} And he was fain to be brought home,\\
The doctor told me: and then a good old woman---\\
\druggerspeaks Yes, faith, she dwells in Sea-coal-lane,---did cure me,\\
With sodden ale, and pellitory of the wall;\\
Cost me but two-pence. I had another sickness\\
Was worse than that.\\
\facespeaks \vinphantom{Was worse than that.} Ay, that was with the grief\\
Thou took'st for being cess'd at eighteen-pence,\\
For the water-work.\\
\druggerspeaks \vinphantom{For the water-work.} In truth, and it was like\\
T' have cost me almost my life.\\
\facespeaks \vinphantom{T' have cost me almost my life.} Thy hair went off?\\
\druggerspeaks Yes, sir; 'twas done for spight.\\
\facespeaks \vinphantom{Yes, sir; 'twas done for spight.} Nay, so says the doctor.\\
\kastrilspeaks Pray thee, tobacco-boy, go fetch my suster;\\
I'll see this learned boy before I go;\\
And so shall she.\\
\facespeaks \vinphantom{And so shall she.} Sir, he is busy now:\\
But if you have a sister to fetch hither,\\
Perhaps your own pains may command her sooner;\\
And he by that time will be free.\\
\kastrilspeaks \vinphantom{And he by that time will be free.} I go.\\
\direct*{EXIT.}
\facespeaks Drugger, she's thine: the damask!---\\
\direct*{EXIT ABEL.}
Subtle and I\\
Must wrestle for her.\\
\direct*{ASIDE.}
---Come on, master Dapper,\\
You see how I turn clients here away,\\
To give your cause dispatch; have you perform'd\\
The ceremonies were enjoin'd you?\\
\dapperspeaks \vinphantom{The ceremonies were enjoin'd you?} Yes, of the vinegar,\\
And the clean shirt.\\
\facespeaks \vinphantom{And the clean shirt.} 'Tis well: that shirt may do you\\
More worship than you think. Your aunt's a-fire,\\
But that she will not shew it, t' have a sight of you.\\
Have you provided for her grace's servants?\\
\dapperspeaks Yes, here are six score Edward shillings.\\
\facespeaks \vinphantom{Yes, here are six score Edward shillings.} Good!\\
\dapperspeaks And an old Harry's sovereign.\\
\facespeaks \vinphantom{And an old Harry's sovereign.} Very good!\\
\dapperspeaks And three James shillings, and an Elizabeth groat,\\
Just twenty nobles.\\
\facespeaks \vinphantom{Just twenty nobles.} O, you are too just.\\
I would you had had the other noble in Maries.\\
\dapperspeaks I have some Philip and Maries.\\
\facespeaks \vinphantom{I have some Philip and Maries.} Ay, those same\\
Are best of all: where are they? Hark, the doctor.\\
\direct*{ENTER SUBTLE, DISGUISED LIKE A PRIEST OF FAIRY, WITH A STRIPE OF CLOTH.}
\subtlespeaks \direct*{IN A FEIGNED VOICE} Is yet her grace's cousin come?\\
\facespeaks He is come.\\
\subtlespeaks \vinphantom{He is come.} And is he fasting?\\
\facespeaks \vinphantom{He is come. And is he fasting?} Yes.\\
\subtlespeaks \vinphantom{He is come. And is he fasting? Yes.} And hath cried hum?\\
\facespeaks Thrice, you must answer.\\
\dapperspeaks \vinphantom{Thrice, you must answer.} Thrice.\\
\subtlespeaks \vinphantom{Thrice, you must answer. Thrice.} And as oft buz?\\
\facespeaks If you have, say.\\
\dapperspeaks \vinphantom{If you have, say.} I have.\\
\subtlespeaks \vinphantom{If you have, say. I have.} Then, to her cuz,\\
Hoping that he hath vinegar'd his senses,\\
As he was bid, the Fairy queen dispenses,\\
By me, this robe, the petticoat of fortune;\\
Which that he straight put on, she doth importune.\\
And though to fortune near be her petticoat,\\
Yet nearer is her smock, the queen doth note:\\
And therefore, ev'n of that a piece she hath sent\\
Which, being a child, to wrap him in was rent;\\
And prays him for a scarf he now will wear it,\\
With as much love as then her grace did tear it,\\
About his eyes,\\
\direct*{THEY BLIND HIM WITH THE RAG,}
to shew he is fortunate.\\
And, trusting unto her to make his state,\\
He'll throw away all worldly pelf about him;\\
Which that he will perform, she doth not doubt him.\\
\facespeaks She need not doubt him, sir. Alas, he has nothing,\\
But what he will part withal as willingly,\\
Upon her grace's word---throw away your purse---\\
As she would ask it;---handkerchiefs and all---\\
\direct*{HE THROWS AWAY, AS THEY BID HIM.}
She cannot bid that thing, but he'll obey.---\\
If you have a ring about you, cast it off,\\
Or a silver seal at your wrist; her grace will send\\
Her fairies here to search you, therefore deal\\
Directly with her highness: if they find\\
That you conceal a mite, you are undone.\\
\dapperspeaks Truly, there's all.\\
\facespeaks \vinphantom{Truly, there's all.} All what?\\
\dapperspeaks \vinphantom{Truly, there's all. All what?} My money; truly.\\
\facespeaks Keep nothing that is transitory about you.\\
\direct*{ASIDE TO SUBTLE.}
Bid Dol play music.---\\
\direct*{DOL PLAYS ON THE CITTERN WITHIN.}
Look, the elves are come.\\
To pinch you, if you tell not truth. Advise you.\\
\direct*{THEY PINCH HIM.}
\dapperspeaks O! I have a paper with a spur-ryal in't.\\
\facespeaks Ti, ti.\\
They knew't, they say.\\
\subtlespeaks \vinphantom{They knew't, they say.} Ti, ti, ti, ti. He has more yet.\\
\facespeaks Ti, ti-ti-ti.\\
\direct*{ASIDE TO SUB.}
In the other pocket.\\
\subtlespeaks \vinphantom{In the other pocket.} Titi, titi, titi, titi, titi.\\
They must pinch him or he will never confess, they say.\\
\direct*{THEY PINCH HIM AGAIN.}
\dapperspeaks O, O!\\
\facespeaks \vinphantom{O, O!} Nay, pray you, hold: he is her grace's nephew,\\
Ti, ti, ti? What care you? good faith, you shall care.---\\
Deal plainly, sir, and shame the fairies. Shew\\
You are innocent.\\
\dapperspeaks \vinphantom{You are innocent.} By this good light, I have nothing.\\
\subtlespeaks Ti, ti, ti, ti, to, ta. He does equivocate she says:\\
Ti, ti do ti, ti ti do, ti da;\\
and swears by the LIGHT when he is blinded.\\
\dapperspeaks By this good DARK, I have nothing but a half-crown\\
Of gold about my wrist, that my love gave me;\\
And a leaden heart I wore since she forsook me.\\
\facespeaks I thought 'twas something. And would you incur\\
Your aunt's displeasure for these trifles? Come,\\
I had rather you had thrown away twenty half-crowns.\\
\direct*{TAKES IT OFF.}
You may wear your leaden heart still.---\\
\direct*{ENTER DOL HASTILY.}
How now!\\
\subtlespeaks \vinphantom{How now!} What news, Dol?\\
\dolspeaks \vinphantom{How now! What news, Dol?} Yonder's your knight, sir Mammon.\\
\facespeaks 'Ods lid, we never thought of him till now!\\
Where is he?\\
\dolspeaks \vinphantom{Where is he?} Here hard by: he is at the door.\\
\subtlespeaks And you are not ready now! Dol, get his suit.\\
\direct*{EXIT DOL.}
He must not be sent back.\\
\facespeaks \vinphantom{He must not be sent back.} O, by no means.\\
What shall we do with this same puffin here,\\
Now he's on the spit?\\
\subtlespeaks \vinphantom{Now he's on the spit?} Why, lay him back awhile,\\
With some device.\\
\direct*{RE-ENTER DOL, WITH FACE'S CLOTHES.}
---Ti, ti, ti, ti, ti, ti, Would her grace speak with me?\\
I come.---Help, Dol!\\
\direct*{KNOCKING WITHOUT.}
\facespeaks \direct*{SPEAKS THROUGH THE KEYHOLE} Who's there? sir Epicure,\\
My master's in the way. Please you to walk\\
Three or four turns, but till his back be turned,\\
And I am for you.---Quickly, Dol!\\
\subtlespeaks \vinphantom{And I am for you.---Quickly, Dol!} Her grace\\
Commends her kindly to you, master Dapper.\\
\dapperspeaks I long to see her grace.\\
\subtlespeaks \vinphantom{I long to see her grace.} She now is set\\
At dinner in her bed, and she has sent you\\
From her own private trencher, a dead mouse,\\
And a piece of gingerbread, to be merry withal,\\
And stay your stomach, lest you faint with fasting:\\
Yet if you could hold out till she saw you, she says,\\
It would be better for you.\\
\facespeaks \vinphantom{It would be better for you.} Sir, he shall\\
Hold out, an 'twere this two hours, for her highness;\\
I can assure you that. We will not lose\\
All we have done.---\\
\subtlespeaks \vinphantom{All we have done.---} He must not see, nor speak\\
To any body, till then.\\
\facespeaks \vinphantom{To any body, till then.} For that we'll put, sir,\\
A stay in's mouth.\\
\subtlespeaks \vinphantom{A stay in's mouth.} Of what?\\
\facespeaks \vinphantom{A stay in's mouth. Of what?} Of gingerbread.\\
Make you it fit. He that hath pleas'd her grace\\
Thus far, shall not now crincle for a little.---\\
Gape, sir, and let him fit you.\\
\direct*{THEY THRUST A GAG OF GINGERBREAD IN HIS MOUTH.}
\subtlespeaks \vinphantom{Gape, sir, and let him fit you.} Where shall we now\\
Bestow him?\\
\dolspeaks \vinphantom{Bestow him?} In the privy.\\
\subtlespeaks \vinphantom{Bestow him? In the privy.} Come along, sir,\\
I now must shew you Fortune's privy lodgings.\\
\facespeaks Are they perfumed, and his bath ready?\\
\subtlespeaks \vinphantom{Are they perfumed, and his bath ready?} All:\\
Only the fumigation's somewhat strong.\\
\facespeaks \direct*{SPEAKING THROUGH THE KEYHOLE}
Sir Epicure, I am yours, sir, by and by.\\
\direct*{EXEUNT WITH DAPPER.}
\act

\scene

\stagedir{A ROOM IN LOVEWIT'S HOUSE. ENTER FACE AND MAMMON.}
\facespeaks O sir, you're come in the only finest time.---\\
\mammonspeaks Where's master?\\
\facespeaks \vinphantom{Where's master?} Now preparing for projection, sir.\\
Your stuff will be all changed shortly.\\
\mammonspeaks \vinphantom{Your stuff will be all changed shortly.} Into gold?\\
\facespeaks To gold and silver, sir.\\
\mammonspeaks \vinphantom{To gold and silver, sir.} Silver I care not for.\\
\facespeaks Yes, sir, a little to give beggars.\\
\mammonspeaks \vinphantom{Yes, sir, a little to give beggars.} Where's the lady?\\
\facespeaks At hand here. I have told her such brave things of you,\\
Touching your bounty, and your noble spirit---\\
\mammonspeaks Hast thou?\\
\facespeaks \vinphantom{Hast thou?} As she is almost in her fit to see you.\\
But, good sir, no divinity in your conference,\\
For fear of putting her in rage.---\\
\mammonspeaks \vinphantom{For fear of putting her in rage.---} I warrant thee.\\
\facespeaks Six men \direct*{sir} will not hold her down: and then,\\
If the old man should hear or see you---\\
\mammonspeaks \vinphantom{If the old man should hear or see you---} Fear not.\\
\facespeaks The very house, sir, would run mad. You know it,\\
How scrupulous he is, and violent,\\
'Gainst the least act of sin. Physic, or mathematics,\\
Poetry, state, or bawdry, as I told you,\\
She will endure, and never startle; but\\
No word of controversy.\\
\mammonspeaks \vinphantom{No word of controversy.} I am school'd, good Ulen.\\
\facespeaks And you must praise her house, remember that,\\
And her nobility.\\
\mammonspeaks \vinphantom{And her nobility.} Let me alone:\\
No herald, no, nor antiquary, Lungs,\\
Shall do it better. Go.\\
\facespeaks \direct*{ASIDE} Why, this is yet\\
A kind of modern happiness, to have\\
Dol Common for a great lady.\\
\direct*{EXIT.}
\mammonspeaks \vinphantom{Dol Common for a great lady.} Now, Epicure,\\
Heighten thyself, talk to her all in gold;\\
Rain her as many showers as Jove did drops\\
Unto his Danae; shew the god a miser,\\
Compared with Mammon. What! the stone will do't.\\
She shall feel gold, taste gold, hear gold, sleep gold;\\
Nay, we will concumbere gold: I will be puissant,\\
And mighty in my talk to her.---\\
\direct*{RE-ENTER FACE, WITH DOL RICHLY DRESSED.}
Here she comes.\\
\facespeaks \vinphantom{Here she comes.} To him, Dol, suckle him.---This is the noble knight,\\
I told your ladyship---\\
\mammonspeaks \vinphantom{I told your ladyship---} Madam, with your pardon,\\
I kiss your vesture.\\
\dolspeaks \vinphantom{I kiss your vesture.} Sir, I were uncivil\\
If I would suffer that; my lip to you, sir.\\
\mammonspeaks I hope my lord your brother be in health, lady.\\
\dolspeaks My lord, my brother is, though I no lady, sir.\\
\facespeaks \direct*{ASIDE} Well said, my Guinea bird.\\
\mammonspeaks Right noble madam---\\
\facespeaks \direct*{ASIDE} O, we shall have most fierce idolatry.\\
\mammonspeaks \vinphantom{Right noble madam---} 'Tis your prerogative.\\
\dolspeaks Rather your courtesy.\\
\mammonspeaks \vinphantom{Rather your courtesy.} Were there nought else to enlarge your virtues to me,\\
These answers speak your breeding and your blood.\\
\dolspeaks Blood we boast none, sir, a poor baron's daughter.\\
\mammonspeaks Poor! and gat you? profane not. Had your father\\
Slept all the happy remnant of his life\\
After that act, lien but there still, and panted,\\
He had done enough to make himself, his issue,\\
And his posterity noble.\\
\dolspeaks \vinphantom{And his posterity noble.} Sir, although\\
We may be said to want the gilt and trappings,\\
The dress of honour, yet we strive to keep\\
The seeds and the materials.\\
\mammonspeaks \vinphantom{The seeds and the materials.} I do see\\
The old ingredient, virtue, was not lost,\\
Nor the drug money used to make your compound.\\
There is a strange nobility in your eye,\\
This lip, that chin! methinks you do resemble\\
One of the Austriac princes.\\
\facespeaks \vinphantom{One of the Austriac princes.} Very like!\\
\direct*{ASIDE.}
Her father was an Irish costermonger.\\
\mammonspeaks The house of Valois just had such a nose,\\
And such a forehead yet the Medici\\
Of Florence boast.\\
\dolspeaks \vinphantom{Of Florence boast.} Troth, and I have been liken'd\\
To all these princes.\\
\facespeaks \direct*{ASIDE} I'll be sworn, I heard it.\\
\mammonspeaks \vinphantom{To all these princes.} I know not how! it is not any one,\\
But e'en the very choice of all their features.\\
\facespeaks \direct*{ASIDE} I'll in, and laugh.\\
\direct*{EXIT.}
\mammonspeaks A certain touch, or air,\\
That sparkles a divinity, beyond\\
An earthly beauty!\\
\dolspeaks \vinphantom{An earthly beauty!} O, you play the courtier.\\
\mammonspeaks Good lady, give me leave---\\
\dolspeaks \vinphantom{Good lady, give me leave---} In faith, I may not,\\
To mock me, sir.\\
\mammonspeaks \vinphantom{To mock me, sir.} To burn in this sweet flame;\\
The phoenix never knew a nobler death.\\
\dolspeaks Nay, now you court the courtier, and destroy\\
What you would build. This art, sir, in your words,\\
Calls your whole faith in question.\\
\mammonspeaks \vinphantom{Calls your whole faith in question.} By my soul---\\
\dolspeaks Nay, oaths are made of the same air, sir.\\
\mammonspeaks \vinphantom{Nay, oaths are made of the same air, sir.} Nature\\
Never bestow'd upon mortality\\
A more unblamed, a more harmonious feature;\\
She play'd the step-dame in all faces else:\\
Sweet Madam, let me be particular---\\
\dolspeaks Particular, sir! I pray you know your distance.\\
\mammonspeaks In no ill sense, sweet lady; but to ask\\
How your fair graces pass the hours? I see\\
You are lodged here, in the house of a rare man,\\
An excellent artist; but what's that to you?\\
\dolspeaks Yes, sir; I study here the mathematics,\\
And distillation.\\
\mammonspeaks \vinphantom{And distillation.} O, I cry your pardon.\\
He's a divine instructor! can extract\\
The souls of all things by his art; call all\\
The virtues, and the miracles of the sun,\\
Into a temperate furnace; teach dull nature\\
What her own forces are. A man, the emperor\\
Has courted above Kelly; sent his medals\\
And chains, to invite him.\\
\dolspeaks \vinphantom{And chains, to invite him.} Ay, and for his physic, sir---\\
\mammonspeaks Above the art of Aesculapius,\\
That drew the envy of the thunderer!\\
I know all this, and more.\\
\dolspeaks \vinphantom{I know all this, and more.} Troth, I am taken, sir,\\
Whole with these studies, that contemplate nature.\\
\mammonspeaks It is a noble humour; but this form\\
Was not intended to so dark a use.\\
Had you been crooked, foul, of some coarse mould\\
A cloister had done well; but such a feature\\
That might stand up the glory of a kingdom,\\
To live recluse! is a mere soloecism,\\
Though in a nunnery. It must not be.\\
I muse, my lord your brother will permit it:\\
You should spend half my land first, were I he.\\
Does not this diamond better on my finger,\\
Than in the quarry?\\
\dolspeaks \vinphantom{Than in the quarry?} Yes.\\
\mammonspeaks \vinphantom{Than in the quarry? Yes.} Why, you are like it.\\
You were created, lady, for the light.\\
Here, you shall wear it; take it, the first pledge\\
Of what I speak, to bind you to believe me.\\
\dolspeaks In chains of adamant?\\
\mammonspeaks \vinphantom{In chains of adamant?} Yes, the strongest bands.\\
And take a secret too---here, by your side,\\
Doth stand this hour, the happiest man in Europe.\\
\dolspeaks You are contended, sir!\\
\mammonspeaks \vinphantom{You are contended, sir!} Nay, in true being,\\
The envy of princes and the fear of states.\\
\dolspeaks Say you so, sir Epicure?\\
\mammonspeaks \vinphantom{Say you so, sir Epicure?} Yes, and thou shalt prove it,\\
Daughter of honour. I have cast mine eye\\
Upon thy form, and I will rear this beauty\\
Above all styles.\\
\dolspeaks \vinphantom{Above all styles.} You mean no treason, sir?\\
\mammonspeaks No, I will take away that jealousy.\\
I am the lord of the philosopher's stone,\\
And thou the lady.\\
\dolspeaks \vinphantom{And thou the lady.} How, sir! have you that?\\
\mammonspeaks I am the master of the mystery.\\
This day the good old wretch here o' the house\\
Has made it for us: now he's at projection.\\
Think therefore thy first wish now, let me hear it;\\
And it shall rain into thy lap, no shower,\\
But floods of gold, whole cataracts, a deluge,\\
To get a nation on thee.\\
\dolspeaks \vinphantom{To get a nation on thee.} You are pleased, sir,\\
To work on the ambition of our sex.\\
\mammonspeaks I am pleased the glory of her sex should know,\\
This nook, here, of the Friars is no climate\\
For her to live obscurely in, to learn\\
Physic and surgery, for the constable's wife\\
Of some odd hundred in Essex; but come forth,\\
And taste the air of palaces; eat, drink\\
The toils of empirics, and their boasted practice;\\
Tincture of pearl, and coral, gold, and amber;\\
Be seen at feasts and triumphs; have it ask'd,\\
What miracle she is; set all the eyes\\
Of court a-fire, like a burning glass,\\
And work them into cinders, when the jewels\\
Of twenty states adorn thee, and the light\\
Strikes out the stars! that when thy name is mention'd,\\
Queens may look pale; and we but shewing our love,\\
Nero's Poppaea may be lost in story!\\
Thus will we have it.\\
\dolspeaks \vinphantom{Thus will we have it.} I could well consent, sir.\\
But, in a monarchy, how will this be?\\
The prince will soon take notice, and both seize\\
You and your stone, it being a wealth unfit\\
For any private subject.\\
\mammonspeaks \vinphantom{For any private subject.} If he knew it.\\
\dolspeaks Yourself do boast it, sir.\\
\mammonspeaks \vinphantom{Yourself do boast it, sir.} To thee, my life.\\
\dolspeaks O, but beware, sir! You may come to end\\
The remnants of your days in a loth'd prison,\\
By speaking of it.\\
\mammonspeaks \vinphantom{By speaking of it.} 'Tis no idle fear.\\
We'll therefore go withal, my girl, and live\\
In a free state, where we will eat our mullets,\\
Soused in high-country wines, sup pheasants' eggs,\\
And have our cockles boil'd in silver shells;\\
Our shrimps to swim again, as when they liv'd,\\
In a rare butter made of dolphins' milk,\\
Whose cream does look like opals; and with these\\
Delicate meats set ourselves high for pleasure,\\
And take us down again, and then renew\\
Our youth and strength with drinking the elixir,\\
And so enjoy a perpetuity\\
Of life and lust! And thou shalt have thy wardrobe\\
Richer than nature's, still to change thy self,\\
And vary oftener, for thy pride, than she,\\
Or art, her wise and almost-equal servant.\\
\direct*{RE-ENTER FACE.}
\facespeaks Sir, you are too loud. I hear you every word\\
Into the laboratory. Some fitter place;\\
The garden, or great chamber above. How like you her?\\
\mammonspeaks Excellent! Lungs. There's for thee.\\
\direct*{GIVES HIM MONEY.}
\facespeaks \vinphantom{Excellent! Lungs. There's for thee.} But do you hear?\\
Good sir, beware, no mention of the rabbins.\\
\mammonspeaks We think not on 'em.\\
\direct*{EXEUNT MAM. AND DOL.}
\facespeaks \vinphantom{We think not on 'em.} O, it is well, sir.---Subtle!\\
\direct*{ENTER SUBTLE.}
Dost thou not laugh?\\
\subtlespeaks \vinphantom{Dost thou not laugh?} Yes; are they gone?\\
\facespeaks \vinphantom{Dost thou not laugh? Yes; are they gone?} All's clear.\\
\subtlespeaks The widow is come.\\
\facespeaks \vinphantom{The widow is come.} And your quarrelling disciple?\\
\subtlespeaks Ay.\\
\facespeaks \vinphantom{Ay.} I must to my captainship again then.\\
\subtlespeaks Stay, bring them in first.\\
\facespeaks \vinphantom{Stay, bring them in first.} So I meant. What is she?\\
A bonnibel?\\
\subtlespeaks \vinphantom{A bonnibel?} I know not.\\
\facespeaks \vinphantom{A bonnibel? I know not.} We'll draw lots:\\
You'll stand to that?\\
\subtlespeaks \vinphantom{You'll stand to that?} What else?\\
\facespeaks \vinphantom{You'll stand to that? What else?} O, for a suit,\\
To fall now like a curtain, flap!\\
\subtlespeaks \vinphantom{To fall now like a curtain, flap!} To the door, man.\\
\facespeaks You'll have the first kiss, 'cause I am not ready.\\
\direct*{EXIT.}
\subtlespeaks Yes, and perhaps hit you through both the nostrils.\\
\facespeaks \direct*{WITHIN} Who would you speak with?\\
\kastrilspeaks \direct*{WITHIN} Where's the captain?\\
\facespeaks \direct*{WITHIN} Gone, sir,\\
About some business.\\
\kastrilspeaks \direct*{WITHIN} Gone!\\
\facespeaks \direct*{WITHIN} He'll return straight.\\
But master doctor, his lieutenant, is here.\\
\direct*{ENTER KASTRIL, FOLLOWED BY DAME PLIANT.}
\subtlespeaks Come near, my worshipful boy, my terrae fili,\\
That is, my boy of land; make thy approaches:\\
Welcome; I know thy lusts, and thy desires,\\
And I will serve and satisfy them. Begin,\\
Charge me from thence, or thence, or in this line;\\
Here is my centre: ground thy quarrel.\\
\kastrilspeaks \vinphantom{Here is my centre: ground thy quarrel.} You lie.\\
\subtlespeaks How, child of wrath and anger! the loud lie?\\
For what, my sudden boy?\\
\kastrilspeaks \vinphantom{For what, my sudden boy?} Nay, that look you to,\\
I am afore-hand.\\
\subtlespeaks \vinphantom{I am afore-hand.} O, this is no true grammar,\\
And as ill logic! You must render causes, child,\\
Your first and second intentions, know your canons\\
And your divisions, moods, degrees, and differences,\\
Your predicaments, substance, and accident,\\
Series, extern and intern, with their causes,\\
Efficient, material, formal, final,\\
And have your elements perfect.\\
\kastrilspeaks \direct*{ASIDE} What is this?\\
The angry tongue he talks in?\\
\subtlespeaks \vinphantom{The angry tongue he talks in?} That false precept,\\
Of being afore-hand, has deceived a number,\\
And made them enter quarrels, often-times,\\
Before they were aware; and afterward,\\
Against their wills.\\
\kastrilspeaks \vinphantom{Against their wills.} How must I do then, sir?\\
\subtlespeaks I cry this lady mercy: she should first\\
Have been saluted.\\
\direct*{KISSES HER.}
I do call you lady,\\
Because you are to be one, ere't be long,\\
My soft and buxom widow.\\
\kastrilspeaks \vinphantom{My soft and buxom widow.} Is she, i'faith?\\
\subtlespeaks Yes, or my art is an egregious liar.\\
\kastrilspeaks How know you?\\
\subtlespeaks \vinphantom{How know you?} By inspection on her forehead,\\
And subtlety of her lip, which must be tasted\\
Often to make a judgment.\\
\direct*{KISSES HER AGAIN.}
'Slight, she melts\\
Like a myrobolane:---here is yet a line,\\
In rivo frontis, tells me he is no knight.\\
\pliantspeaks What is he then, sir?\\
\subtlespeaks \vinphantom{What is he then, sir?} Let me see your hand.\\
O, your linea fortunae makes it plain;\\
And stella here in monte Veneris.\\
But, most of all, junctura annularis.\\
He is a soldier, or a man of art, lady,\\
But shall have some great honour shortly.\\
\pliantspeaks Brother,\\
He's a rare man, believe me!\\
\direct*{RE-ENTER FACE, IN HIS UNIFORM.}
\kastrilspeaks \vinphantom{He's a rare man, believe me!} Hold your peace.\\
Here comes the t'other rare man.---'Save you, captain.\\
\facespeaks Good master Kastril! Is this your sister?\\
\kastrilspeaks Ay, sir.\\
Please you to kuss her, and be proud to know her.\\
\facespeaks I shall be proud to know you, lady.\\
\direct*{KISSES HER.}
\pliantspeaks \vinphantom{I shall be proud to know you, lady.} Brother,\\
He calls me lady too.\\
\kastrilspeaks \vinphantom{He calls me lady too.} Ay, peace: I heard it.\\
\direct*{TAKES HER ASIDE.}
\facespeaks The count is come.\\
\subtlespeaks \vinphantom{The count is come.} Where is he?\\
\facespeaks \vinphantom{The count is come. Where is he?} At the door.\\
\subtlespeaks Why, you must entertain him.\\
\facespeaks \vinphantom{Why, you must entertain him.} What will you do\\
With these the while?\\
\subtlespeaks \vinphantom{With these the while?} Why, have them up, and shew them\\
Some fustian book, or the dark glass.\\
\facespeaks \vinphantom{Some fustian book, or the dark glass.} 'Fore God,\\
She is a delicate dab-chick! I must have her.\\
\direct*{EXIT.}
\subtlespeaks Must you! ay, if your fortune will, you must.---\\
Come, sir, the captain will come to us presently:\\
I'll have you to my chamber of demonstrations,\\
Where I will shew you both the grammar and logic,\\
And rhetoric of quarrelling; my whole method\\
Drawn out in tables; and my instrument,\\
That hath the several scales upon't, shall make you\\
Able to quarrel at a straw's-breadth by moon-light.\\
And, lady, I'll have you look in a glass,\\
Some half an hour, but to clear your eye-sight,\\
Against you see your fortune; which is greater,\\
Than I may judge upon the sudden, trust me.\\
\direct*{EXIT, FOLLOWED BY KAST. AND DAME P.}
\direct*{RE-ENTER FACE.}
\facespeaks Where are you, doctor?\\
\subtlespeaks \direct*{WITHIN} I'll come to you presently.\\
\facespeaks \vinphantom{Where are you, doctor?} I will have this same widow, now I have seen her,\\
On any composition.\\
\direct*{RE-ENTER SUBTLE.}
\subtlespeaks \vinphantom{On any composition.} What do you say?\\
\facespeaks Have you disposed of them?\\
\subtlespeaks \vinphantom{Have you disposed of them?} I have sent them up.\\
\facespeaks Subtle, in troth, I needs must have this widow.\\
\subtlespeaks Is that the matter?\\
\facespeaks \vinphantom{Is that the matter?} Nay, but hear me.\\
\subtlespeaks \vinphantom{Is that the matter? Nay, but hear me.} Go to.\\
If you rebel once, Dol shall know it all:\\
Therefore be quiet, and obey your chance.\\
\facespeaks Nay, thou art so violent now---Do but conceive,\\
Thou art old, and canst not serve---\\
\subtlespeaks \vinphantom{Thou art old, and canst not serve---} Who cannot? I?\\
'Slight, I will serve her with thee, for a---\\
\facespeaks \vinphantom{'Slight, I will serve her with thee, for a---} Nay,\\
But understand: I'll give you composition.\\
\subtlespeaks I will not treat with thee; what! sell my fortune?\\
'Tis better than my birth-right. Do not murmur:\\
Win her, and carry her. If you grumble, Dol\\
Knows it directly.\\
\facespeaks \vinphantom{Knows it directly.} Well, sir, I am silent.\\
Will you go help to fetch in Don in state?\\
\direct*{EXIT.}
\subtlespeaks I follow you, sir. We must keep Face in awe,\\
Or he will over-look us like a tyrant.\\
\direct*{RE-ENTER FACE, INTRODUCING SURLY DISGUISED AS A SPANIARD.}
Brain of a tailor! who comes here? Don John!\\
\surlyspeaks Senores, beso las manos a vuestras mercedes.\\
\subtlespeaks Would you had stoop'd a little, and kist our anos!\\
\facespeaks Peace, Subtle.\\
\subtlespeaks \vinphantom{Peace, Subtle.} Stab me; I shall never hold, man.\\
He looks in that deep ruff like a head in a platter,\\
Serv'd in by a short cloke upon two trestles.\\
\facespeaks Or, what do you say to a collar of brawn, cut down\\
Beneath the souse, and wriggled with a knife?\\
\subtlespeaks 'Slud, he does look too fat to be a Spaniard.\\
\facespeaks Perhaps some Fleming or some Hollander got him\\
In d'Alva's time; count Egmont's bastard.\\
\subtlespeaks \vinphantom{In d'Alva's time; count Egmont's bastard.} Don,\\
Your scurvy, yellow, Madrid face is welcome.\\
\surlyspeaks Gratia.\\
\subtlespeaks \vinphantom{Gratia.} He speaks out of a fortification.\\
Pray God he have no squibs in those deep sets.\\
\surlyspeaks Por dios, senores, muy linda casa!\\
\subtlespeaks What says he?\\
\facespeaks \vinphantom{What says he?} Praises the house, I think;\\
I know no more but's action.\\
\subtlespeaks \vinphantom{I know no more but's action.} Yes, the casa,\\
My precious Diego, will prove fair enough\\
To cozen you in. Do you mark? you shall\\
Be cozen'd, Diego.\\
\facespeaks \vinphantom{Be cozen'd, Diego.} Cozen'd, do you see,\\
My worthy Donzel, cozen'd.\\
\surlyspeaks \vinphantom{My worthy Donzel, cozen'd.} Entiendo.\\
\subtlespeaks Do you intend it? so do we, dear Don.\\
Have you brought pistolets, or portagues,\\
My solemn Don?---Dost thou feel any?\\
\facespeaks \direct*{FEELS HIS POCKETS} Full.\\
\subtlespeaks \vinphantom{My solemn Don?---Dost thou feel any?} You shall be emptied, Don, pumped and drawn\\
Dry, as they say.\\
\facespeaks \vinphantom{Dry, as they say.} Milked, in troth, sweet Don.\\
\subtlespeaks \vinphantom{Dry, as they say. Milked, in troth, sweet Don.} See all the monsters; the great lion of all, Don.\\
\surlyspeaks Con licencia, se puede ver a esta senora?\\
\subtlespeaks What talks he now?\\
\facespeaks \vinphantom{What talks he now?} Of the sennora.\\
\subtlespeaks \vinphantom{What talks he now? Of the sennora.} O, Don,\\
This is the lioness, which you shall see\\
Also, my Don.\\
\facespeaks \vinphantom{Also, my Don.} 'Slid, Subtle, how shall we do?\\
\subtlespeaks For what?\\
\facespeaks \vinphantom{For what?} Why Dol's employ'd, you know.\\
\subtlespeaks \vinphantom{For what? Why Dol's employ'd, you know.} That's true.\\
'Fore heaven, I know not: he must stay, that's all.\\
\facespeaks Stay! that he must not by no means.\\
\subtlespeaks \vinphantom{Stay! that he must not by no means.} No! why?\\
\facespeaks Unless you'll mar all. 'Slight, he will suspect it:\\
And then he will not pay, not half so well.\\
This is a travelled punk-master, and does know\\
All the delays; a notable hot rascal,\\
And looks already rampant.\\
\subtlespeaks \vinphantom{And looks already rampant.} 'Sdeath, and Mammon\\
Must not be troubled.\\
\facespeaks \vinphantom{Must not be troubled.} Mammon! in no case.\\
\subtlespeaks What shall we do then?\\
\facespeaks \vinphantom{What shall we do then?} Think: you must be sudden.\\
\surlyspeaks Entiendo que la senora es tan hermosa, que codicio tan\\
verla, como la bien aventuranza de mi vida.\\
\facespeaks Mi vida! 'Slid, Subtle, he puts me in mind of the widow.\\
What dost thou say to draw her to it, ha!\\
And tell her 'tis her fortune? all our venture\\
Now lies upon't. It is but one man more,\\
Which of us chance to have her: and beside,\\
There is no maidenhead to be fear'd or lost.\\
What dost thou think on't, Subtle?\\
\subtlespeaks \vinphantom{What dost thou think on't, Subtle?} Who? I? why---\\
\facespeaks The credit of our house too is engaged.\\
\subtlespeaks You made me an offer for my share erewhile.\\
What wilt thou give me, i'faith?\\
\facespeaks \vinphantom{What wilt thou give me, i'faith?} O, by that light\\
I'll not buy now: You know your doom to me.\\
E'en take your lot, obey your chance, sir; win her,\\
And wear her out, for me.\\
\subtlespeaks \vinphantom{And wear her out, for me.} 'Slight, I'll not work her then.\\
\facespeaks It is the common cause; therefore bethink you.\\
Dol else must know it, as you said.\\
\subtlespeaks \vinphantom{Dol else must know it, as you said.} I care not.\\
\surlyspeaks Senores, porque se tarda tanto?\\
\subtlespeaks Faith, I am not fit, I am old.\\
\facespeaks \vinphantom{Faith, I am not fit, I am old.} That's now no reason, sir.\\
\surlyspeaks Puede ser de hazer burla de mi amor?\\
\facespeaks You hear the Don too? by this air, I call,\\
And loose the hinges: Dol!\\
\subtlespeaks \vinphantom{And loose the hinges: Dol!} A plague of hell---\\
\facespeaks Will you then do?\\
\subtlespeaks \vinphantom{Will you then do?} You are a terrible rogue!\\
I'll think of this: will you, sir, call the widow?\\
\facespeaks Yes, and I'll take her too with all her faults,\\
Now I do think on't better.\\
\subtlespeaks \vinphantom{Now I do think on't better.} With all my heart, sir;\\
Am I discharged o' the lot?\\
\facespeaks \vinphantom{Am I discharged o' the lot?} As you please.\\
\subtlespeaks Hands.\\
\direct*{THEY TAKE HANDS.}
\facespeaks \vinphantom{Hands.} Remember now, that upon any change,\\
You never claim her.\\
\subtlespeaks \vinphantom{You never claim her.} Much good joy, and health to you, sir,\\
Marry a whore! fate, let me wed a witch first.\\
\surlyspeaks Por estas honradas barbas---\\
\subtlespeaks \vinphantom{Por estas honradas barbas---} He swears by his beard.\\
Dispatch, and call the brother too.\\
\direct*{EXIT FACE.}
\surlyspeaks \vinphantom{Dispatch, and call the brother too.} Tengo duda, senores, que no me hagan alguna traycion.\\
\subtlespeaks How, issue on? yes, praesto, sennor. Please you\\
Enthratha the chambrata, worthy don:\\
Where if you please the fates, in your bathada,\\
You shall be soked, and stroked, and tubb'd and rubb'd,\\
And scrubb'd, and fubb'd, dear don, before you go.\\
You shall in faith, my scurvy baboon don,\\
Be curried, claw'd, and flaw'd, and taw'd, indeed.\\
I will the heartlier go about it now,\\
And make the widow a punk so much the sooner,\\
To be revenged on this impetuous Face:\\
The quickly doing of it is the grace.\\
\direct*{EXEUNT SUB. AND SURLY.}
\scene

\stagedir{ANOTHER ROOM IN THE SAME. ENTER FACE, KASTRIL, AND DAME PLIANT.}
\facespeaks Come, lady: I knew the Doctor would not leave,\\
Till he had found the very nick of her fortune.\\
\kastrilspeaks To be a countess, say you, a Spanish countess, sir?\\
\pliantspeaks Why, is that better than an English countess?\\
\facespeaks Better! 'Slight, make you that a question, lady?\\
\kastrilspeaks Nay, she is a fool, captain, you must pardon her.\\
\facespeaks Ask from your courtier, to your inns-of-court-man,\\
To your mere milliner; they will tell you all,\\
Your Spanish gennet is the best horse; your Spanish\\
Stoup is the best garb; your Spanish beard\\
Is the best cut; your Spanish ruffs are the best\\
Wear; your Spanish pavin the best dance;\\
Your Spanish titillation in a glove\\
The best perfume: and for your Spanish pike,\\
And Spanish blade, let your poor captain speak---\\
Here comes the doctor.\\
\direct*{ENTER SUBTLE, WITH A PAPER.}
\subtlespeaks \vinphantom{Here comes the doctor.} My most honour'd lady,\\
For so I am now to style you, having found\\
By this my scheme, you are to undergo\\
An honourable fortune, very shortly.\\
What will you say now, if some---\\
\facespeaks \vinphantom{What will you say now, if some---} I have told her all, sir,\\
And her right worshipful brother here, that she shall be\\
A countess; do not delay them, sir; a Spanish countess.\\
\subtlespeaks Still, my scarce-worshipful captain, you can keep\\
No secret! Well, since he has told you, madam,\\
Do you forgive him, and I do.\\
\kastrilspeaks \vinphantom{Do you forgive him, and I do.} She shall do that, sir;\\
I'll look to it, 'tis my charge.\\
\subtlespeaks \vinphantom{I'll look to it, 'tis my charge.} Well then: nought rests\\
But that she fit her love now to her fortune.\\
\pliantspeaks Truly I shall never brook a Spaniard.\\
\subtlespeaks No!\\
\pliantspeaks \vinphantom{No!} Never since eighty-eight could I abide them,\\
And that was some three year afore I was born, in truth.\\
\subtlespeaks Come, you must love him, or be miserable,\\
Choose which you will.\\
\facespeaks \vinphantom{Choose which you will.} By this good rush, persuade her,\\
She will cry strawberries else within this twelvemonth.\\
\subtlespeaks Nay, shads and mackerel, which is worse.\\
\facespeaks \vinphantom{Nay, shads and mackerel, which is worse.} Indeed, sir!\\
\kastrilspeaks Od's lid, you shall love him, or I'll kick you.\\
\pliantspeaks Why,\\
I'll do as you will have me, brother.\\
\kastrilspeaks \vinphantom{I'll do as you will have me, brother.} Do,\\
Or by this hand I'll maul you.\\
\facespeaks \vinphantom{Or by this hand I'll maul you.} Nay, good sir,\\
Be not so fierce.\\
\subtlespeaks \vinphantom{Be not so fierce.} No, my enraged child;\\
She will be ruled. What, when she comes to taste\\
The pleasures of a countess! to be courted---\\
\facespeaks And kiss'd, and ruffled!\\
\subtlespeaks \vinphantom{And kiss'd, and ruffled!} Ay, behind the hangings.\\
\facespeaks And then come forth in pomp!\\
\subtlespeaks \vinphantom{And then come forth in pomp!} And know her state!\\
\facespeaks Of keeping all the idolaters of the chamber\\
Barer to her, than at their prayers!\\
\subtlespeaks \vinphantom{Barer to her, than at their prayers!} Is serv'd\\
Upon the knee!\\
\facespeaks \vinphantom{Upon the knee!} And has her pages, ushers,\\
Footmen, and coaches---\\
\subtlespeaks \vinphantom{Footmen, and coaches---} Her six mares---\\
\facespeaks \vinphantom{Footmen, and coaches--- Her six mares---} Nay, eight!\\
\subtlespeaks To hurry her through London, to the Exchange,\\
Bethlem, the china-houses---\\
\facespeaks \vinphantom{Bethlem, the china-houses---} Yes, and have\\
The citizens gape at her, and praise her tires,\\
And my lord's goose-turd bands, that ride with her!\\
\kastrilspeaks Most brave! By this hand, you are not my suster,\\
If you refuse.\\
\pliantspeaks \vinphantom{If you refuse.} I will not refuse, brother.\\
\direct*{ENTER SURLY.}
\surlyspeaks Que es esto, senores, que no venga?\\
Esta tardanza me mata!\\
\facespeaks \vinphantom{Esta tardanza me mata!} It is the count come:\\
The doctor knew he would be here, by his art.\\
\subtlespeaks En gallanta madama, Don! gallantissima!\\
\surlyspeaks Por todos los dioses, la mas acabada hermosura, que he visto\\
en mi vida!\\
\facespeaks \vinphantom{en mi vida!} Is't not a gallant language that they speak?\\
\kastrilspeaks An admirable language! Is't not French?\\
\facespeaks No, Spanish, sir.\\
\kastrilspeaks \vinphantom{No, Spanish, sir.} It goes like law-French,\\
And that, they say, is the courtliest language.\\
\facespeaks List, sir.\\
\surlyspeaks \vinphantom{List, sir.} El sol ha perdido su lumbre, con el esplandor que trae\\
esta dama! Valgame dios!\\
\facespeaks \vinphantom{esta dama! Valgame dios!} He admires your sister.\\
\kastrilspeaks Must not she make curt'sy?\\
\subtlespeaks \vinphantom{Must not she make curt'sy?} Ods will, she must go to him, man, and kiss him!\\
It is the Spanish fashion, for the women\\
To make first court.\\
\facespeaks \vinphantom{To make first court.} 'Tis true he tells you, sir:\\
His art knows all.\\
\surlyspeaks \vinphantom{His art knows all.} Porque no se acude?\\
\kastrilspeaks He speaks to her, I think.\\
\facespeaks \vinphantom{He speaks to her, I think.} That he does, sir.\\
\surlyspeaks Por el amor de dios, que es esto que se tarda?\\
\kastrilspeaks Nay, see: she will not understand him! gull,\\
Noddy.\\
\pliantspeaks \vinphantom{Noddy.} What say you, brother?\\
\kastrilspeaks \vinphantom{Noddy. What say you, brother?} Ass, my suster.\\
Go kuss him, as the cunning man would have you;\\
I'll thrust a pin in your buttocks else.\\
\facespeaks \vinphantom{I'll thrust a pin in your buttocks else.} O no, sir.\\
\surlyspeaks Senora mia, mi persona esta muy indigna de allegar\\
a tanta hermosura.\\
\facespeaks \vinphantom{a tanta hermosura.} Does he not use her bravely?\\
\kastrilspeaks Bravely, i'faith!\\
\facespeaks \vinphantom{Bravely, i'faith!} Nay, he will use her better.\\
\kastrilspeaks Do you think so?\\
\surlyspeaks \vinphantom{Do you think so?} Senora, si sera servida, entremonos.\\
\direct*{EXIT WITH DAME PLIANT.}
\kastrilspeaks Where does he carry her?\\
\facespeaks \vinphantom{Where does he carry her?} Into the garden, sir;\\
Take you no thought: I must interpret for her.\\
\subtlespeaks Give Dol the word.\\
\direct*{ASIDE TO FACE, WHO GOES OUT.}
---Come, my fierce child, advance,\\
We'll to our quarrelling lesson again.\\
\kastrilspeaks Agreed.\\
I love a Spanish boy with all my heart.\\
\subtlespeaks Nay, and by this means, sir, you shall be brother\\
To a great count.\\
\kastrilspeaks \vinphantom{To a great count.} Ay, I knew that at first,\\
This match will advance the house of the Kastrils.\\
\subtlespeaks 'Pray God your sister prove but pliant!\\
\kastrilspeaks \vinphantom{'Pray God your sister prove but pliant!} Why,\\
Her name is so, by her other husband.\\
\subtlespeaks How!\\
\kastrilspeaks \vinphantom{How!} The widow Pliant. Knew you not that?\\
\subtlespeaks No, faith, sir;\\
Yet, by erection of her figure, I guest it.\\
Come, let's go practise.\\
\kastrilspeaks \vinphantom{Come, let's go practise.} Yes, but do you think, doctor,\\
I e'er shall quarrel well?\\
\subtlespeaks \vinphantom{I e'er shall quarrel well?} I warrant you.\\
\direct*{EXEUNT.}
\scene

\stagedir{ANOTHER ROOM IN THE SAME. ENTER DOL IN HER FIT OF RAVING, FOLLOWED BY MAMMON.}
\dolspeaks "For after Alexander's death"---\\
\mammonspeaks \vinphantom{"For after Alexander's death"---} Good lady---\\
\dolspeaks "That Perdiccas and Antigonus, were slain,\\
The two that stood, Seleuc', and Ptolomee"---\\
\mammonspeaks Madam---\\
\dolspeaks \vinphantom{Madam---} "Made up the two legs, and the fourth beast,\\
That was Gog-north, and Egypt-south: which after\\
Was call'd Gog-iron-leg and South-iron-leg"---\\
\mammonspeaks Lady---\\
\dolspeaks \vinphantom{Lady---} "And then Gog-horned. So was Egypt, too:\\
Then Egypt-clay-leg, and Gog-clay-leg"---\\
\mammonspeaks \vinphantom{Then Egypt-clay-leg, and Gog-clay-leg"---} Sweet madam---\\
\dolspeaks "And last Gog-dust, and Egypt-dust, which fall\\
In the last link of the fourth chain. And these\\
Be stars in story, which none see, or look at"---\\
\mammonspeaks What shall I do?\\
\dolspeaks \vinphantom{What shall I do?} "For," as he says, "except\\
We call the rabbins, and the heathen Greeks"---\\
\mammonspeaks Dear lady---\\
\dolspeaks \vinphantom{Dear lady---} "To come from Salem, and from Athens,\\
And teach the people of Great Britain"---\\
\direct*{ENTER FACE, HASTILY, IN HIS SERVANT'S DRESS.}
\facespeaks \vinphantom{And teach the people of Great Britain"---} What's the matter, sir?\\
\dolspeaks "To speak the tongue of Eber, and Javan"---\\
\mammonspeaks O,\\
She's in her fit.\\
\dolspeaks \vinphantom{She's in her fit.} "We shall know nothing"---\\
\facespeaks \vinphantom{She's in her fit. "We shall know nothing"---} Death, sir,\\
We are undone!\\
\dolspeaks \vinphantom{We are undone!} "Where then a learned linguist\\
Shall see the ancient used communion\\
Of vowels and consonants"---\\
\facespeaks \vinphantom{Of vowels and consonants"---} My master will hear!\\
\dolspeaks "A wisdom, which Pythagoras held most high"---\\
\mammonspeaks Sweet honourable lady!\\
\dolspeaks \vinphantom{Sweet honourable lady!} "To comprise\\
All sounds of voices, in few marks of letters"---\\
\facespeaks Nay, you must never hope to lay her now.\\
\direct*{THEY ALL SPEAK TOGETHER.}
\dolspeaks "And so we may arrive by Talmud skill,\\
And profane Greek, to raise the building up\\
Of Helen's house against the Ismaelite,\\
King of Thogarma, and his habergions\\
Brimstony, blue, and fiery; and the force\\
Of king Abaddon, and the beast of Cittim:\\
Which rabbi David Kimchi, Onkelos,\\
And Aben Ezra do interpret Rome."\\
\facespeaks How did you put her into't?\\
\mammonspeaks \vinphantom{How did you put her into't?} Alas, I talk'd\\
Of a fifth monarchy I would erect,\\
With the philosopher's stone, by chance, and she\\
Falls on the other four straight.\\
\facespeaks \vinphantom{Falls on the other four straight.} Out of Broughton!\\
I told you so. 'Slid, stop her mouth.\\
\mammonspeaks \vinphantom{I told you so. 'Slid, stop her mouth.} Is't best?\\
\facespeaks She'll never leave else. If the old man hear her,\\
We are but faeces, ashes.\\
\subtlespeaks \direct*{WITHIN} What's to do there?\\
\facespeaks \vinphantom{We are but faeces, ashes.} O, we are lost! Now she hears him, she is quiet.\\
\direct*{ENTER SUBTLE, THEY RUN DIFFERENT WAYS.}
\mammonspeaks Where shall I hide me!\\
\subtlespeaks \vinphantom{Where shall I hide me!} How! what sight is here?\\
Close deeds of darkness, and that shun the light!\\
Bring him again. Who is he? What, my son!\\
O, I have lived too long.\\
\mammonspeaks \vinphantom{O, I have lived too long.} Nay, good, dear father,\\
There was no unchaste purpose.\\
\subtlespeaks \vinphantom{There was no unchaste purpose.} Not? and flee me\\
When I come in?\\
\mammonspeaks \vinphantom{When I come in?} That was my error.\\
\subtlespeaks \vinphantom{When I come in? That was my error.} Error?\\
Guilt, guilt, my son: give it the right name. No marvel,\\
If I found check in our great work within,\\
When such affairs as these were managing!\\
\mammonspeaks Why, have you so?\\
\subtlespeaks \vinphantom{Why, have you so?} It has stood still this half hour:\\
And all the rest of our less works gone back.\\
Where is the instrument of wickedness,\\
My lewd false drudge?\\
\mammonspeaks \vinphantom{My lewd false drudge?} Nay, good sir, blame not him;\\
Believe me, 'twas against his will or knowledge:\\
I saw her by chance.\\
\subtlespeaks \vinphantom{I saw her by chance.} Will you commit more sin,\\
To excuse a varlet?\\
\mammonspeaks \vinphantom{To excuse a varlet?} By my hope, 'tis true, sir.\\
\subtlespeaks Nay, then I wonder less, if you, for whom\\
The blessing was prepared, would so tempt heaven,\\
And lose your fortunes.\\
\mammonspeaks \vinphantom{And lose your fortunes.} Why, sir?\\
\subtlespeaks \vinphantom{And lose your fortunes. Why, sir?} This will retard\\
The work a month at least.\\
\mammonspeaks \vinphantom{The work a month at least.} Why, if it do,\\
What remedy? But think it not, good father:\\
Our purposes were honest.\\
\subtlespeaks \vinphantom{Our purposes were honest.} As they were,\\
So the reward will prove.\\
\direct*{A LOUD EXPLOSION WITHIN.}
---How now! ah me!\\
God, and all saints be good to us.---\\
\direct*{RE-ENTER FACE.}
What's that?\\
\facespeaks \vinphantom{What's that?} O, sir, we are defeated! all the works\\
Are flown in fumo, every glass is burst;\\
Furnace, and all rent down, as if a bolt\\
Of thunder had been driven through the house.\\
Retorts, receivers, pelicans, bolt-heads,\\
All struck in shivers!\\
\direct*{SUBTLE FALLS DOWN AS IN A SWOON.}
Help, good sir! alas,\\
Coldness and death invades him. Nay, sir Mammon,\\
Do the fair offices of a man! you stand,\\
As you were readier to depart than he.\\
\direct*{KNOCKING WITHIN.}
Who's there? my lord her brother is come.\\
\mammonspeaks \vinphantom{Who's there? my lord her brother is come.} Ha, Lungs!\\
\facespeaks His coach is at the door. Avoid his sight,\\
For he's as furious as his sister's mad.\\
\mammonspeaks Alas!\\
\facespeaks \vinphantom{Alas!} My brain is quite undone with the fume, sir,\\
I ne'er must hope to be mine own man again.\\
\mammonspeaks Is all lost, Lungs? will nothing be preserv'd\\
Of all our cost?\\
\facespeaks \vinphantom{Of all our cost?} Faith, very little, sir;\\
A peck of coals or so, which is cold comfort, sir.\\
\mammonspeaks O, my voluptuous mind! I am justly punish'd.\\
\facespeaks And so am I, sir.\\
\mammonspeaks \vinphantom{And so am I, sir.} Cast from all my hopes---\\
\facespeaks Nay, certainties, sir.\\
\mammonspeaks \vinphantom{Nay, certainties, sir.} By mine own base affections.\\
\subtlespeaks \direct*{SEEMING TO COME TO HIMSELF}
O, the curst fruits of vice and lust!\\
\mammonspeaks \vinphantom{O, the curst fruits of vice and lust!} Good father,\\
It was my sin. Forgive it.\\
\subtlespeaks \vinphantom{It was my sin. Forgive it.} Hangs my roof\\
Over us still, and will not fall, O justice,\\
Upon us, for this wicked man!\\
\facespeaks \vinphantom{Upon us, for this wicked man!} Nay, look, sir,\\
You grieve him now with staying in his sight:\\
Good sir, the nobleman will come too, and take you,\\
And that may breed a tragedy.\\
\mammonspeaks \vinphantom{And that may breed a tragedy.} I'll go.\\
\facespeaks Ay, and repent at home, sir. It may be,\\
For some good penance you may have it yet;\\
A hundred pound to the box at Bethlem---\\
\mammonspeaks Yes.\\
\facespeaks \vinphantom{Yes.} For the restoring such as---have their wits.\\
\mammonspeaks I'll do't.\\
\facespeaks \vinphantom{I'll do't.} I'll send one to you to receive it.\\
\mammonspeaks Do.\\
Is no projection left?\\
\facespeaks \vinphantom{Is no projection left?} All flown, or stinks, sir.\\
\mammonspeaks Will nought be sav'd that's good for med'cine,\\
think'st thou?\\
\facespeaks \vinphantom{think'st thou?} I cannot tell, sir. There will be perhaps,\\
Something about the scraping of the shards,\\
Will cure the itch,---though not your itch of mind, sir.\\
\direct*{ASIDE.}
It shall be saved for you, and sent home. Good sir,\\
This way, for fear the lord should meet you.\\
\direct*{EXIT MAMMON.}
\subtlespeaks \direct*{RAISING HIS HEAD} Face!\\
\facespeaks \vinphantom{This way, for fear the lord should meet you.} Ay.\\
\subtlespeaks Is he gone?\\
\facespeaks \vinphantom{Is he gone?} Yes, and as heavily\\
As all the gold he hoped for were in's blood.\\
Let us be light though.\\
\subtlespeaks \direct*{LEAPING UP} Ay, as balls, and bound\\
And hit our heads against the roof for joy:\\
There's so much of our care now cast away.\\
\facespeaks Now to our don.\\
\subtlespeaks \vinphantom{Now to our don.} Yes, your young widow by this time\\
Is made a countess, Face; she has been in travail\\
Of a young heir for you.\\
\facespeaks \vinphantom{Of a young heir for you.} Good sir.\\
\subtlespeaks \vinphantom{Of a young heir for you. Good sir.} Off with your case,\\
And greet her kindly, as a bridegroom should,\\
After these common hazards.\\
\facespeaks \vinphantom{After these common hazards.} Very well, sir.\\
Will you go fetch Don Diego off, the while?\\
\subtlespeaks And fetch him over too, if you'll be pleased, sir:\\
Would Dol were in her place, to pick his pockets now!\\
\facespeaks Why, you can do't as well, if you would set to't.\\
I pray you prove your virtue.\\
\subtlespeaks \vinphantom{I pray you prove your virtue.} For your sake sir.\\
\direct*{EXEUNT.}
\scene

\stagedir{ANOTHER ROOM IN THE SAME.}
\direct*{ENTER SURLY AND DAME PLIANT.}
\surlyspeaks Lady, you see into what hands you are fall'n;\\
'Mongst what a nest of villains! and how near\\
Your honour was t' have catch'd a certain clap,\\
Through your credulity, had I but been\\
So punctually forward, as place, time,\\
And other circumstances would have made a man;\\
For you're a handsome woman: would you were wise too!\\
I am a gentleman come here disguised,\\
Only to find the knaveries of this citadel;\\
And where I might have wrong'd your honour, and have not,\\
I claim some interest in your love. You are,\\
They say, a widow, rich: and I'm a batchelor,\\
Worth nought: your fortunes may make me a man,\\
As mine have preserv'd you a woman. Think upon it,\\
And whether I have deserv'd you or no.\\
\pliantspeaks I will, sir.\\
\surlyspeaks \vinphantom{I will, sir.} And for these household-rogues, let me alone\\
To treat with them.\\
\direct*{ENTER SUBTLE.}
\subtlespeaks \vinphantom{To treat with them.} How doth my noble Diego,\\
And my dear madam countess? hath the count\\
Been courteous, lady? liberal, and open?\\
Donzel, methinks you look melancholic,\\
After your coitum, and scurvy: truly,\\
I do not like the dulness of your eye;\\
It hath a heavy cast, 'tis upsee Dutch,\\
And says you are a lumpish whore-master.\\
Be lighter, and I will make your pockets so.\\
\direct*{ATTEMPTS TO PICK THEM.}
\surlyspeaks \direct*{THROWS OPEN HIS CLOAK} Will you, don bawd and\\
pickpurse?\\
\direct*{STRIKES HIM DOWN.}
how now! reel you?\\
Stand up, sir, you shall find, since I am so heavy,\\
I'll give you equal weight.\\
\subtlespeaks \vinphantom{I'll give you equal weight.} Help! murder!\\
\surlyspeaks \vinphantom{I'll give you equal weight. Help! murder!} No, sir,\\
There's no such thing intended: a good cart,\\
And a clean whip shall ease you of that fear.\\
I am the Spanish don "that should be cozen'd,\\
Do you see, cozen'd?" Where's your Captain Face,\\
That parcel broker, and whole-bawd, all rascal!\\
\direct*{ENTER FACE, IN HIS UNIFORM.}
\facespeaks How, Surly!\\
\surlyspeaks \vinphantom{How, Surly!} O, make your approach, good captain.\\
I have found from whence your copper rings and spoons\\
Come, now, wherewith you cheat abroad in taverns.\\
'Twas here you learned t' anoint your boot with brimstone,\\
Then rub men's gold on't for a kind of touch,\\
And say 'twas naught, when you had changed the colour,\\
That you might have't for nothing. And this doctor,\\
Your sooty, smoky-bearded compeer, he\\
Will close you so much gold, in a bolt's-head,\\
And, on a turn, convey in the stead another\\
With sublimed mercury, that shall burst in the heat,\\
And fly out all in fumo! Then weeps Mammon;\\
Then swoons his worship.\\
\direct*{FACE SLIPS OUT.}
Or, he is the Faustus,\\
That casteth figures and can conjure, cures\\
Plagues, piles, and pox, by the ephemerides,\\
And holds intelligence with all the bawds\\
And midwives of three shires: while you send in---\\
Captain!---what! is he gone?---damsels with child,\\
Wives that are barren, or the waiting-maid\\
With the green sickness.\\
\direct*{SEIZES SUBTLE AS HE IS RETIRING.}
---Nay, sir, you must tarry,\\
Though he be scaped; and answer by the ears, sir.\\
\direct*{RE-ENTER FACE, WITH KASTRIL.}
\facespeaks Why, now's the time, if ever you will quarrel\\
Well, as they say, and be a true-born child:\\
The doctor and your sister both are abused.\\
\kastrilspeaks Where is he? which is he? he is a slave,\\
Whate'er he is, and the son of a whore.---Are you\\
The man, sir, I would know?\\
\surlyspeaks \vinphantom{The man, sir, I would know?} I should be loth, sir,\\
To confess so much.\\
\kastrilspeaks \vinphantom{To confess so much.} Then you lie in your throat.\\
\surlyspeaks How!\\
\facespeaks \direct*{TO KASTRIL} A very errant rogue, sir, and a cheater,\\
Employ'd here by another conjurer\\
That does not love the doctor, and would cross him,\\
If he knew how.\\
\surlyspeaks \vinphantom{If he knew how.} Sir, you are abused.\\
\kastrilspeaks \vinphantom{If he knew how. Sir, you are abused.} You lie:\\
And 'tis no matter.\\
\facespeaks \vinphantom{And 'tis no matter.} Well said, sir! He is\\
The impudent'st rascal---\\
\surlyspeaks \vinphantom{The impudent'st rascal---} You are indeed: Will you hear me, sir?\\
\facespeaks By no means: bid him be gone.\\
\kastrilspeaks \vinphantom{By no means: bid him be gone.} Begone, sir, quickly.\\
\surlyspeaks This 's strange!---Lady, do you inform your brother.\\
\facespeaks There is not such a foist in all the town,\\
The doctor had him presently; and finds yet,\\
The Spanish count will come here.\\
\direct*{ASIDE.}
---Bear up, Subtle.\\
\subtlespeaks \vinphantom{---Bear up, Subtle.} Yes, sir, he must appear within this hour.\\
\facespeaks And yet this rogue would come in a disguise,\\
By the temptation of another spirit,\\
To trouble our art, though he could not hurt it!\\
\kastrilspeaks Ay,\\
I know---Away,\\
\direct*{TO HIS SISTER.}
you talk like a foolish mauther.\\
\surlyspeaks \vinphantom{you talk like a foolish mauther.} Sir, all is truth she says.\\
\facespeaks Do not believe him, sir.\\
He is the lying'st swabber! Come your ways, sir.\\
\surlyspeaks You are valiant out of company!\\
\kastrilspeaks \vinphantom{You are valiant out of company!} Yes, how then, sir?\\
\direct*{ENTER DRUGGER, WITH A PIECE OF DAMASK.}
\facespeaks Nay, here's an honest fellow, too, that knows him,\\
And all his tricks. Make good what I say, Abel,\\
This cheater would have cozen'd thee o' the widow.---\\
\direct*{ASIDE TO DRUG.}
He owes this honest Drugger here, seven pound,\\
He has had on him, in two-penny'orths of tobacco.\\
\druggerspeaks Yes, sir.\\
And he has damn'd himself three terms to pay me.\\
\facespeaks And what does he owe for lotium?\\
\druggerspeaks \vinphantom{And what does he owe for lotium?} Thirty shillings, sir;\\
And for six syringes.\\
\surlyspeaks \vinphantom{And for six syringes.} Hydra of villainy!\\
\facespeaks Nay, sir, you must quarrel him out o' the house.\\
\kastrilspeaks I will:\\
---Sir, if you get not out of doors, you lie;\\
And you are a pimp.\\
\surlyspeaks \vinphantom{And you are a pimp.} Why, this is madness, sir,\\
Not valour in you; I must laugh at this.\\
\kastrilspeaks It is my humour: you are a pimp and a trig,\\
And an Amadis de Gaul, or a Don Quixote.\\
\druggerspeaks Or a knight o' the curious coxcomb, do you see?\\
\direct*{ENTER ANANIAS.}
\ananiasspeaks Peace to the household!\\
\kastrilspeaks \vinphantom{Peace to the household!} I'll keep peace for no man.\\
\ananiasspeaks Casting of dollars is concluded lawful.\\
\kastrilspeaks Is he the constable?\\
\subtlespeaks \vinphantom{Is he the constable?} Peace, Ananias.\\
\facespeaks No, sir.\\
\kastrilspeaks \vinphantom{No, sir.} Then you are an otter, and a shad, a whit,\\
A very tim.\\
\surlyspeaks \vinphantom{A very tim.} You'll hear me, sir?\\
\kastrilspeaks \vinphantom{A very tim. You'll hear me, sir?} I will not.\\
\ananiasspeaks What is the motive?\\
\subtlespeaks \vinphantom{What is the motive?} Zeal in the young gentleman,\\
Against his Spanish slops.\\
\ananiasspeaks \vinphantom{Against his Spanish slops.} They are profane,\\
Lewd, superstitious, and idolatrous breeches.\\
\surlyspeaks New rascals!\\
\kastrilspeaks \vinphantom{New rascals!} Will you begone, sir?\\
\ananiasspeaks \vinphantom{New rascals! Will you begone, sir?} Avoid, Sathan!\\
Thou art not of the light: That ruff of pride\\
About thy neck, betrays thee; and is the same\\
With that which the unclean birds, in seventy-seven,\\
Were seen to prank it with on divers coasts:\\
Thou look'st like antichrist, in that lewd hat.\\
\surlyspeaks I must give way.\\
\kastrilspeaks \vinphantom{I must give way.} Be gone, sir.\\
\surlyspeaks \vinphantom{I must give way. Be gone, sir.} But I'll take\\
A course with you---\\
\ananiasspeaks \vinphantom{A course with you---} Depart, proud Spanish fiend!\\
\surlyspeaks Captain and doctor.\\
\ananiasspeaks \vinphantom{Captain and doctor.} Child of perdition!\\
\kastrilspeaks Hence, sir!---\\
\direct*{EXIT SURLY.}
Did I not quarrel bravely?\\
\facespeaks \vinphantom{Did I not quarrel bravely?} Yes, indeed, sir.\\
\kastrilspeaks Nay, an I give my mind to't, I shall do't.\\
\facespeaks O, you must follow, sir, and threaten him tame:\\
He'll turn again else.\\
\kastrilspeaks \vinphantom{He'll turn again else.} I'll re-turn him then.\\
\direct*{EXIT.}
\direct*{SUBTLE TAKES ANANIAS ASIDE.}
\facespeaks Drugger, this rogue prevented us for thee:\\
We had determin'd that thou should'st have come\\
In a Spanish suit, and have carried her so; and he,\\
A brokerly slave! goes, puts it on himself.\\
Hast brought the damask?\\
\druggerspeaks \vinphantom{Hast brought the damask?} Yes, sir.\\
\facespeaks \vinphantom{Hast brought the damask? Yes, sir.} Thou must borrow\\
A Spanish suit. Hast thou no credit with the players?\\
\druggerspeaks Yes, sir; did you never see me play the Fool?\\
\facespeaks I know not, Nab:---Thou shalt, if I can help it.---\\
\direct*{ASIDE.}
Hieronimo's old cloak, ruff, and hat will serve;\\
I'll tell thee more when thou bring'st 'em.\\
\direct*{EXIT DRUGGER.}
\ananiasspeaks \vinphantom{I'll tell thee more when thou bring'st 'em.} Sir, I know\\
The Spaniard hates the brethren, and hath spies\\
Upon their actions: and that this was one\\
I make no scruple.---But the holy synod\\
Have been in prayer and meditation for it;\\
And 'tis revealed no less to them than me,\\
That casting of money is most lawful.\\
\subtlespeaks True.\\
But here I cannot do it: if the house\\
Shou'd chance to be suspected, all would out,\\
And we be locked up in the Tower for ever,\\
To make gold there for the state, never come out;\\
And then are you defeated.\\
\ananiasspeaks \vinphantom{And then are you defeated.} I will tell\\
This to the elders and the weaker brethren,\\
That the whole company of the separation\\
May join in humble prayer again.\\
\subtlespeaks \vinphantom{May join in humble prayer again.} And fasting.\\
\ananiasspeaks Yea, for some fitter place. The peace of mind\\
Rest with these walls!\\
\direct*{EXIT.}
\subtlespeaks \vinphantom{Rest with these walls!} Thanks, courteous Ananias.\\
\facespeaks What did he come for?\\
\subtlespeaks \vinphantom{What did he come for?} About casting dollars,\\
Presently out of hand. And so I told him,\\
A Spanish minister came here to spy,\\
Against the faithful---\\
\facespeaks \vinphantom{Against the faithful---} I conceive. Come, Subtle,\\
Thou art so down upon the least disaster!\\
How wouldst thou ha' done, if I had not help't thee out?\\
\subtlespeaks I thank thee, Face, for the angry boy, i'faith.\\
\facespeaks Who would have look'd it should have been that rascal,\\
Surly? he had dyed his beard and all. Well, sir.\\
Here's damask come to make you a suit.\\
\subtlespeaks \vinphantom{Here's damask come to make you a suit.} Where's Drugger?\\
\facespeaks He is gone to borrow me a Spanish habit;\\
I'll be the count, now.\\
\subtlespeaks \vinphantom{I'll be the count, now.} But where's the widow?\\
\facespeaks Within, with my lord's sister; madam Dol\\
Is entertaining her.\\
\subtlespeaks \vinphantom{Is entertaining her.} By your favour, Face,\\
Now she is honest, I will stand again.\\
\facespeaks You will not offer it.\\
\subtlespeaks \vinphantom{You will not offer it.} Why?\\
\facespeaks \vinphantom{You will not offer it. Why?} Stand to your word,\\
Or---here comes Dol, she knows---\\
\subtlespeaks \vinphantom{Or---here comes Dol, she knows---} You are tyrannous still.\\
\direct*{ENTER DOL, HASTILY.}
\facespeaks Strict for my right.---How now, Dol!\\
Hast \direct*{thou} told her,\\
The Spanish count will come?\\
\dolspeaks \vinphantom{The Spanish count will come?} Yes; but another is come,\\
You little look'd for!\\
\facespeaks \vinphantom{You little look'd for!} Who's that?\\
\dolspeaks \vinphantom{You little look'd for! Who's that?} Your master;\\
The master of the house.\\
\subtlespeaks \vinphantom{The master of the house.} How, Dol!\\
\facespeaks \vinphantom{The master of the house. How, Dol!} She lies,\\
This is some trick. Come, leave your quiblins, Dorothy.\\
\dolspeaks Look out, and see.\\
\direct*{FACE GOES TO THE WINDOW.}
\subtlespeaks \vinphantom{Look out, and see.} Art thou in earnest?\\
\dolspeaks \vinphantom{Look out, and see. Art thou in earnest?} 'Slight,\\
Forty of the neighbours are about him, talking.\\
\facespeaks 'Tis he, by this good day.\\
\dolspeaks \vinphantom{'Tis he, by this good day.} 'Twill prove ill day\\
For some on us.\\
\facespeaks \vinphantom{For some on us.} We are undone, and taken.\\
\dolspeaks Lost, I'm afraid.\\
\subtlespeaks \vinphantom{Lost, I'm afraid.} You said he would not come,\\
While there died one a week within the liberties.\\
\facespeaks No: 'twas within the walls.\\
\subtlespeaks \vinphantom{No: 'twas within the walls.} Was't so! cry you mercy.\\
I thought the liberties. What shall we do now, Face?\\
\facespeaks Be silent: not a word, if he call or knock.\\
I'll into mine old shape again and meet him,\\
Of Jeremy, the butler. In the mean time,\\
Do you two pack up all the goods and purchase,\\
That we can carry in the two trunks. I'll keep him\\
Off for to-day, if I cannot longer: and then\\
At night, I'll ship you both away to Ratcliff,\\
Where we will meet to-morrow, and there we'll share.\\
Let Mammon's brass and pewter keep the cellar;\\
We'll have another time for that. But, Dol,\\
'Prythee go heat a little water quickly;\\
Subtle must shave me: all my captain's beard\\
Must off, to make me appear smooth Jeremy.\\
You'll do it?\\
\subtlespeaks \vinphantom{You'll do it?} Yes, I'll shave you, as well as I can.\\
\facespeaks And not cut my throat, but trim me?\\
\subtlespeaks \vinphantom{And not cut my throat, but trim me?} You shall see, sir.\\
\direct*{EXEUNT.}
\act

\scene

\stagedir{BEFORE LOVEWIT'S DOOR. ENTER LOVEWIT, WITH SEVERAL OF THE NEIGHBOURS.}
\lovewitspeaks Has there been such resort, say you?\\
\neighonespeaks \vinphantom{Has there been such resort, say you?} Daily, sir.\\
\neightwospeaks And nightly, too.\\
\neighthreespeaks \vinphantom{And nightly, too.} Ay, some as brave as lords.\\
\neighfourspeaks Ladies and gentlewomen.\\
\neighfivespeaks \vinphantom{Ladies and gentlewomen.} Citizens' wives.\\
\neighonespeaks And knights.\\
\neighsixspeaks \vinphantom{And knights.} In coaches.\\
\neightwospeaks \vinphantom{And knights. In coaches.} Yes, and oyster women.\\
\neighonespeaks Beside other gallants.\\
\neighthreespeaks \vinphantom{Beside other gallants.} Sailors' wives.\\
\neighfourspeaks \vinphantom{Beside other gallants. Sailors' wives.} Tobacco men.\\
\neighfivespeaks Another Pimlico!\\
\lovewitspeaks \vinphantom{Another Pimlico!} What should my knave advance,\\
To draw this company? he hung out no banners\\
Of a strange calf with five legs to be seen,\\
Or a huge lobster with six claws?\\
\neighsixspeaks \vinphantom{Or a huge lobster with six claws?} No, sir.\\
\neighthreespeaks We had gone in then, sir.\\
\lovewitspeaks \vinphantom{We had gone in then, sir.} He has no gift\\
Of teaching in the nose that e'er I knew of.\\
You saw no bills set up that promised cure\\
Of agues, or the tooth-ach?\\
\neightwospeaks \vinphantom{Of agues, or the tooth-ach?} No such thing, sir!\\
\lovewitspeaks Nor heard a drum struck for baboons or puppets?\\
\neighfivespeaks Neither, sir.\\
\lovewitspeaks \vinphantom{Neither, sir.} What device should he bring forth now?\\
I love a teeming wit as I love my nourishment:\\
'Pray God he have not kept such open house,\\
That he hath sold my hangings, and my bedding!\\
I left him nothing else. If he have eat them,\\
A plague o' the moth, say I! Sure he has got\\
Some bawdy pictures to call all this ging!\\
The friar and the nun; or the new motion\\
Of the knight's courser covering the parson's mare;\\
Or 't may be, he has the fleas that run at tilt\\
Upon a table, or some dog to dance.\\
When saw you him?\\
\neighonespeaks \vinphantom{When saw you him?} Who, sir, Jeremy?\\
\neightwospeaks \vinphantom{When saw you him? Who, sir, Jeremy?} Jeremy butler?\\
We saw him not this month.\\
\lovewitspeaks \vinphantom{We saw him not this month.} How!\\
\neighfourspeaks \vinphantom{We saw him not this month. How!} Not these five weeks, sir.\\
\neighsixspeaks These six weeks at the least.\\
\lovewitspeaks \vinphantom{These six weeks at the least.} You amaze me, neighbours!\\
\neighfivespeaks Sure, if your worship know not where he is,\\
He's slipt away.\\
\neighsixspeaks \vinphantom{He's slipt away.} Pray God, he be not made away.\\
\lovewitspeaks Ha! it's no time to question, then.\\
\direct*{KNOCKS AT THE DOOR.}
\neighsixspeaks \vinphantom{Ha! it's no time to question, then.} About\\
Some three weeks since, I heard a doleful cry,\\
As I sat up a mending my wife's stockings.\\
\lovewitspeaks 'Tis strange that none will answer! Didst thou hear\\
A cry, sayst thou?\\
\neighsixspeaks \vinphantom{A cry, sayst thou?} Yes, sir, like unto a man\\
That had been strangled an hour, and could not speak.\\
\neightwospeaks I heard it too, just this day three weeks, at two o'clock\\
Next morning.\\
\lovewitspeaks \vinphantom{Next morning.} These be miracles, or you make them so!\\
A man an hour strangled, and could not speak,\\
And both you heard him cry?\\
\neighthreespeaks \vinphantom{And both you heard him cry?} Yes, downward, sir.\\
Love, Thou art a wise fellow. Give me thy hand, I pray thee.\\
What trade art thou on?\\
\neighthreespeaks \vinphantom{What trade art thou on?} A smith, an't please your worship.\\
\lovewitspeaks A smith! then lend me thy help to get this door open.\\
\neighthreespeaks That I will presently, sir, but fetch my tools---\\
\direct*{EXIT.}
\neighonespeaks Sir, best to knock again, afore you break it.\\
\lovewitspeaks \direct*{KNOCKS AGAIN} I will.\\
\direct*{ENTER FACE, IN HIS BUTLER'S LIVERY.}
\facespeaks What mean you, sir?\\
1, 2, 4 NEI. O, here's Jeremy!\\
\facespeaks \vinphantom{What mean you, sir?} Good sir, come from the door.\\
\lovewitspeaks Why, what's the matter?\\
\facespeaks \vinphantom{Why, what's the matter?} Yet farther, you are too near yet.\\
\lovewitspeaks In the name of wonder,\\
What means the fellow!\\
\facespeaks \vinphantom{What means the fellow!} The house, sir, has been visited.\\
\lovewitspeaks What, with the plague? stand thou then farther.\\
\facespeaks \vinphantom{What, with the plague? stand thou then farther.} No, sir,\\
I had it not.\\
\lovewitspeaks \vinphantom{I had it not.} Who had it then? I left\\
None else but thee in the house.\\
\facespeaks \vinphantom{None else but thee in the house.} Yes, sir, my fellow,\\
The cat that kept the buttery, had it on her\\
A week before I spied it; but I got her\\
Convey'd away in the night: and so I shut\\
The house up for a month---\\
\lovewitspeaks \vinphantom{The house up for a month---} How!\\
\facespeaks \vinphantom{The house up for a month--- How!} Purposing then, sir,\\
To have burnt rose-vinegar, treacle, and tar,\\
And have made it sweet, that you shou'd ne'er have known it;\\
Because I knew the news would but afflict you, sir.\\
\lovewitspeaks Breathe less, and farther off! Why this is stranger:\\
The neighbours tell me all here that the doors\\
Have still been open---\\
\facespeaks \vinphantom{Have still been open---} How, sir!\\
\lovewitspeaks \vinphantom{Have still been open--- How, sir!} Gallants, men and women,\\
And of all sorts, tag-rag, been seen to flock here\\
In threaves, these ten weeks, as to a second Hogsden,\\
In days of Pimlico and Eye-bright.\\
\facespeaks \vinphantom{In days of Pimlico and Eye-bright.} Sir,\\
Their wisdoms will not say so.\\
\lovewitspeaks \vinphantom{Their wisdoms will not say so.} To-day they speak\\
Of coaches and gallants; one in a French hood\\
Went in, they tell me; and another was seen\\
In a velvet gown at the window: divers more\\
Pass in and out.\\
\facespeaks \vinphantom{Pass in and out.} They did pass through the doors then,\\
Or walls, I assure their eye-sights, and their spectacles;\\
For here, sir, are the keys, and here have been,\\
In this my pocket, now above twenty days:\\
And for before, I kept the fort alone there.\\
But that 'tis yet not deep in the afternoon,\\
I should believe my neighbours had seen double\\
Through the black pot, and made these apparitions!\\
For, on my faith to your worship, for these three weeks\\
And upwards the door has not been open'd.\\
\lovewitspeaks Strange!\\
\neighonespeaks \vinphantom{Strange!} Good faith, I think I saw a coach.\\
\neightwospeaks \vinphantom{Strange! Good faith, I think I saw a coach.} And I too,\\
I'd have been sworn.\\
\lovewitspeaks \vinphantom{I'd have been sworn.} Do you but think it now?\\
And but one coach?\\
\neighfourspeaks \vinphantom{And but one coach?} We cannot tell, sir: Jeremy\\
Is a very honest fellow.\\
\facespeaks \vinphantom{Is a very honest fellow.} Did you see me at all?\\
\neighonespeaks No; that we are sure on.\\
\neightwospeaks \vinphantom{No; that we are sure on.} I'll be sworn o' that.\\
\lovewitspeaks Fine rogues to have your testimonies built on!\\
\direct*{RE-ENTER THIRD NEIGHBOUR, WITH HIS TOOLS.}
\neighthreespeaks Is Jeremy come!\\
\neighonespeaks \vinphantom{Is Jeremy come!} O yes; you may leave your tools;\\
We were deceived, he says.\\
\neightwospeaks \vinphantom{We were deceived, he says.} He has had the keys;\\
And the door has been shut these three weeks.\\
\neighthreespeaks \vinphantom{And the door has been shut these three weeks.} Like enough.\\
\lovewitspeaks Peace, and get hence, you changelings.\\
\direct*{ENTER SURLY AND MAMMON.}
\facespeaks \direct*{ASIDE} Surly come!\\
And Mammon made acquainted! they'll tell all.\\
How shall I beat them off? what shall I do?\\
Nothing's more wretched than a guilty conscience.\\
\surlyspeaks No, sir, he was a great physician. This,\\
It was no bawdy-house, but a mere chancel!\\
You knew the lord and his sister.\\
\mammonspeaks \vinphantom{You knew the lord and his sister.} Nay, good Surly.---\\
\surlyspeaks The happy word, BE RICH---\\
\mammonspeaks \vinphantom{The happy word, BE RICH---} Play not the tyrant.---\\
\surlyspeaks "Should be to-day pronounced to all your friends."\\
And where be your andirons now? and your brass pots,\\
That should have been golden flagons, and great wedges?\\
\mammonspeaks Let me but breathe. What, they have shut their doors,\\
Methinks!\\
\surlyspeaks \vinphantom{Methinks!} Ay, now 'tis holiday with them.\\
\mammonspeaks Rogues,\\
\direct*{HE AND SURLY KNOCK.}
Cozeners, impostors, bawds!\\
\facespeaks \vinphantom{Cozeners, impostors, bawds!} What mean you, sir?\\
\mammonspeaks To enter if we can.\\
\facespeaks \vinphantom{To enter if we can.} Another man's house!\\
Here is the owner, sir: turn you to him,\\
And speak your business.\\
\mammonspeaks \vinphantom{And speak your business.} Are you, sir, the owner?\\
\lovewitspeaks Yes, sir.\\
\mammonspeaks \vinphantom{Yes, sir.} And are those knaves within your cheaters!\\
\lovewitspeaks What knaves, what cheaters?\\
\mammonspeaks \vinphantom{What knaves, what cheaters?} Subtle and his Lungs.\\
\facespeaks The gentleman is distracted, sir! No lungs,\\
Nor lights have been seen here these three weeks, sir,\\
Within these doors, upon my word.\\
\surlyspeaks \vinphantom{Within these doors, upon my word.} Your word,\\
Groom arrogant!\\
\facespeaks \vinphantom{Groom arrogant!} Yes, sir, I am the housekeeper,\\
And know the keys have not been out of my hands.\\
\surlyspeaks This is a new Face.\\
\facespeaks \vinphantom{This is a new Face.} You do mistake the house, sir:\\
What sign was't at?\\
\surlyspeaks \vinphantom{What sign was't at?} You rascal! this is one\\
Of the confederacy. Come, let's get officers,\\
And force the door.\\
\lovewitspeaks \vinphantom{And force the door.} 'Pray you stay, gentlemen.\\
\surlyspeaks No, sir, we'll come with warrant.\\
\mammonspeaks \vinphantom{No, sir, we'll come with warrant.} Ay, and then\\
We shall have your doors open.\\
\direct*{EXEUNT MAM. AND SUR.}
\lovewitspeaks \vinphantom{We shall have your doors open.} What means this?\\
\facespeaks I cannot tell, sir.\\
I NEI. These are two of the gallants\\
That we do think we saw.\\
\facespeaks \vinphantom{That we do think we saw.} Two of the fools!\\
Your talk as idly as they. Good faith, sir,\\
I think the moon has crazed 'em all.---\\
\direct*{ASIDE.}
O me,\\
\direct*{ENTER KASTRIL.}
The angry boy come too! He'll make a noise,\\
And ne'er away till he have betray'd us all.\\
\kastrilspeaks \direct*{KNOCKING} What rogues, bawds, slaves,\\
you'll open the door, anon!\\
Punk, cockatrice, my suster! By this light\\
I'll fetch the marshal to you. You are a whore\\
To keep your castle---\\
\facespeaks \vinphantom{To keep your castle---} Who would you speak with, sir?\\
\kastrilspeaks The bawdy doctor, and the cozening captain,\\
And puss my suster.\\
\lovewitspeaks \vinphantom{And puss my suster.} This is something, sure.\\
\facespeaks Upon my trust, the doors were never open, sir.\\
\kastrilspeaks I have heard all their tricks told me twice over,\\
By the fat knight and the lean gentleman.\\
\lovewitspeaks Here comes another.\\
\direct*{ENTER ANANIAS AND TRIBULATION.}
\facespeaks \vinphantom{Here comes another.} Ananias too!\\
And his pastor!\\
\tribulationspeaks \direct*{BEATING AT THE DOOR} The doors are shut against us.\\
\ananiasspeaks \vinphantom{And his pastor!} Come forth, you seed of sulphur, sons of fire!\\
Your stench it is broke forth; abomination\\
Is in the house.\\
\kastrilspeaks \vinphantom{Is in the house.} Ay, my suster's there.\\
\ananiasspeaks \vinphantom{Is in the house. Ay, my suster's there.} The place,\\
It is become a cage of unclean birds.\\
\kastrilspeaks Yes, I will fetch the scavenger, and the constable.\\
\tribulationspeaks You shall do well.\\
\ananiasspeaks \vinphantom{You shall do well.} We'll join to weed them out.\\
\kastrilspeaks You will not come then, punk devise, my sister!\\
\ananiasspeaks Call her not sister; she's a harlot verily.\\
\kastrilspeaks I'll raise the street.\\
\lovewitspeaks \vinphantom{I'll raise the street.} Good gentlemen, a word.\\
\ananiasspeaks Satan avoid, and hinder not our zeal!\\
\direct*{EXEUNT ANA., TRIB., AND KAST.}
\lovewitspeaks The world's turn'd Bethlem.\\
\facespeaks \vinphantom{The world's turn'd Bethlem.} These are all broke loose,\\
Out of St. Katherine's, where they use to keep\\
The better sort of mad-folks.\\
\neighonespeaks \vinphantom{The better sort of mad-folks.} All these persons\\
We saw go in and out here.\\
\neightwospeaks \vinphantom{We saw go in and out here.} Yes, indeed, sir.\\
\neighthreespeaks These were the parties.\\
\facespeaks \vinphantom{These were the parties.} Peace, you drunkards! Sir,\\
I wonder at it: please you to give me leave\\
To touch the door, I'll try an the lock be chang'd.\\
\lovewitspeaks It mazes me!\\
\facespeaks \direct*{GOES TO THE DOOR} Good faith, sir, I believe\\
There's no such thing: 'tis all deceptio visus.---\\
\direct*{ASIDE.}
Would I could get him away.\\
\dapperspeaks \direct*{WITHIN} Master captain! master doctor!\\
\lovewitspeaks \vinphantom{Would I could get him away.} Who's that?\\
\facespeaks \vinphantom{Would I could get him away. Who's that?} Our clerk within, that I forgot!\\
\direct*{ASIDE.}
I know not, sir.\\
\dapperspeaks \direct*{WITHIN} For God's sake, when will her grace be at leisure?\\
\facespeaks \vinphantom{I know not, sir.} Ha!\\
Illusions, some spirit o' the air---\\
\direct*{ASIDE.}
His gag is melted,\\
And now he sets out the throat.\\
\dapperspeaks \direct*{WITHIN} I am almost stifled---\\
\facespeaks \direct*{ASIDE} Would you were altogether.\\
\lovewitspeaks \vinphantom{And now he sets out the throat.} 'Tis in the house.\\
Ha! list.\\
\facespeaks \vinphantom{Ha! list.} Believe it, sir, in the air.\\
\lovewitspeaks \vinphantom{Ha! list. Believe it, sir, in the air.} Peace, you.\\
\dapperspeaks \direct*{WITHIN} Mine aunt's grace does not use me well.\\
\subtlespeaks \direct*{WITHIN} You fool,\\
Peace, you'll mar all.\\
\facespeaks \direct*{SPEAKS THROUGH THE KEYHOLE, WHILE LOVEWIT ADVANCES TO THE DOOR UNOBSERVED}
Or you will else, you rogue.\\
\lovewitspeaks \vinphantom{Or you will else, you rogue.} O, is it so? Then you converse with spirits!---\\
Come, sir. No more of your tricks, good Jeremy.\\
The truth, the shortest way.\\
\facespeaks \vinphantom{The truth, the shortest way.} Dismiss this rabble, sir.---\\
\direct*{ASIDE.}
What shall I do? I am catch'd.\\
\lovewitspeaks \vinphantom{What shall I do? I am catch'd.} Good neighbours,\\
I thank you all. You may depart.\\
\direct*{EXEUNT NEIGHBOURS.}
---Come, sir,\\
You know that I am an indulgent master;\\
And therefore conceal nothing. What's your medicine,\\
To draw so many several sorts of wild fowl?\\
\facespeaks Sir, you were wont to affect mirth and wit---\\
But here's no place to talk on't in the street.\\
Give me but leave to make the best of my fortune,\\
And only pardon me the abuse of your house:\\
It's all I beg. I'll help you to a widow,\\
In recompence, that you shall give me thanks for,\\
Will make you seven years younger, and a rich one.\\
'Tis but your putting on a Spanish cloak:\\
I have her within. You need not fear the house;\\
It was not visited.\\
\lovewitspeaks \vinphantom{It was not visited.} But by me, who came\\
Sooner than you expected.\\
\facespeaks \vinphantom{Sooner than you expected.} It is true, sir.\\
'Pray you forgive me.\\
\lovewitspeaks \vinphantom{'Pray you forgive me.} Well: let's see your widow.\\
\direct*{EXEUNT.}
\scene

\stagedir{A ROOM IN THE SAME. ENTER SUBTLE, LEADING IN DAPPER, WITH HIS EYES BOUND AS BEFORE.}
\subtlespeaks How! you have eaten your gag?\\
\dapperspeaks \vinphantom{How! you have eaten your gag?} Yes faith, it crumbled\\
Away in my mouth.\\
\subtlespeaks \vinphantom{Away in my mouth.} You have spoil'd all then.\\
\dapperspeaks No!\\
I hope my aunt of Fairy will forgive me.\\
\subtlespeaks Your aunt's a gracious lady; but in troth\\
You were to blame.\\
\dapperspeaks \vinphantom{You were to blame.} The fume did overcome me,\\
And I did do't to stay my stomach. 'Pray you\\
So satisfy her grace.\\
\direct*{ENTER FACE, IN HIS UNIFORM.}
Here comes the captain.\\
\facespeaks \vinphantom{Here comes the captain.} How now! is his mouth down?\\
\subtlespeaks Ay, he has spoken!\\
\facespeaks \vinphantom{Ay, he has spoken!} A pox, I heard him, and you too.\\
---He's undone then.---\\
I have been fain to say, the house is haunted\\
With spirits, to keep churl back.\\
\subtlespeaks \vinphantom{With spirits, to keep churl back.} And hast thou done it?\\
\facespeaks Sure, for this night.\\
\subtlespeaks \vinphantom{Sure, for this night.} Why, then triumph and sing\\
Of Face so famous, the precious king\\
Of present wits.\\
\facespeaks \vinphantom{Of present wits.} Did you not hear the coil\\
About the door?\\
\subtlespeaks \vinphantom{About the door?} Yes, and I dwindled with it.\\
\facespeaks Show him his aunt, and let him be dispatch'd:\\
I'll send her to you.\\
\direct*{EXIT FACE.}
\subtlespeaks \vinphantom{I'll send her to you.} Well, sir, your aunt her grace\\
Will give you audience presently, on my suit,\\
And the captain's word that you did not eat your gag\\
In any contempt of her highness.\\
\direct*{UNBINDS HIS EYES.}
\dapperspeaks \vinphantom{In any contempt of her highness.} Not I, in troth, sir.\\
\direct*{ENTER DOL, LIKE THE QUEEN OF FAIRY.}
\subtlespeaks Here she is come. Down o' your knees and wriggle:\\
She has a stately presence.\\
\direct*{DAPPER KNEELS, AND SHUFFLES TOWARDS HER.}
Good! Yet nearer,\\
And bid, God save you!\\
\dapperspeaks \vinphantom{And bid, God save you!} Madam!\\
\subtlespeaks \vinphantom{And bid, God save you! Madam!} And your aunt.\\
\dapperspeaks And my most gracious aunt, God save your grace.\\
\dolspeaks Nephew, we thought to have been angry with you;\\
But that sweet face of yours hath turn'd the tide,\\
And made it flow with joy, that ebb'd of love.\\
Arise, and touch our velvet gown.\\
\subtlespeaks \vinphantom{Arise, and touch our velvet gown.} The skirts,\\
And kiss 'em. So!\\
\dolspeaks \vinphantom{And kiss 'em. So!} Let me now stroak that head.\\
"Much, nephew, shalt thou win, much shalt thou spend,\\
Much shalt thou give away, much shalt thou lend."\\
\subtlespeaks \direct*{ASIDE} Ay, much! indeed.---\\
Why do you not thank her grace?\\
\dapperspeaks \vinphantom{Why do you not thank her grace?} I cannot speak for joy.\\
\subtlespeaks See, the kind wretch!\\
Your grace's kinsman right.\\
\dolspeaks \vinphantom{Your grace's kinsman right.} Give me the bird.\\
Here is your fly in a purse, about your neck, cousin;\\
Wear it, and feed it about this day sev'n-night,\\
On your right wrist---\\
\subtlespeaks \vinphantom{On your right wrist---} Open a vein with a pin,\\
And let it suck but once a week; till then,\\
You must not look on't.\\
\dolspeaks \vinphantom{You must not look on't.} No: and kinsman,\\
Bear yourself worthy of the blood you come on.\\
\subtlespeaks Her grace would have you eat no more Woolsack pies,\\
Nor Dagger frumety.\\
\dolspeaks \vinphantom{Nor Dagger frumety.} Nor break his fast\\
In Heaven and Hell.\\
\subtlespeaks \vinphantom{In Heaven and Hell.} She's with you every where!\\
Nor play with costarmongers, at mum-chance, tray-trip,\\
God make you rich; (when as your aunt has done it);\\
But keep\\
The gallant'st company, and the best games---\\
\dapperspeaks Yes, sir.\\
\subtlespeaks \vinphantom{Yes, sir.} Gleek and primero; and what you get, be true to us.\\
\dapperspeaks By this hand, I will.\\
\subtlespeaks \vinphantom{By this hand, I will.} You may bring's a thousand pound\\
Before to-morrow night, if but three thousand\\
Be stirring, an you will.\\
\dapperspeaks \vinphantom{Be stirring, an you will.} I swear I will then.\\
\subtlespeaks Your fly will learn you all games.\\
\facespeaks \direct*{WITHIN} Have you done there?\\
\subtlespeaks \vinphantom{Your fly will learn you all games.} Your grace will command him no more duties?\\
\dolspeaks No:\\
But come, and see me often. I may chance\\
To leave him three or four hundred chests of treasure,\\
And some twelve thousand acres of fairy land,\\
If he game well and comely with good gamesters.\\
\subtlespeaks There's a kind aunt! kiss her departing part.---\\
But you must sell your forty mark a year, now.\\
\dapperspeaks Ay, sir, I mean.\\
\subtlespeaks \vinphantom{Ay, sir, I mean.} Or, give't away; pox on't!\\
\dapperspeaks I'll give't mine aunt. I'll go and fetch the writings.\\
\direct*{EXIT.}
\subtlespeaks 'Tis well---away!\\
\direct*{RE-ENTER FACE.}
\facespeaks \vinphantom{'Tis well---away!} Where's Subtle?\\
\subtlespeaks \vinphantom{'Tis well---away! Where's Subtle?} Here: what news?\\
\facespeaks Drugger is at the door, go take his suit,\\
And bid him fetch a parson, presently;\\
Say, he shall marry the widow. Thou shalt spend\\
A hundred pound by the service!\\
\direct*{EXIT SUBTLE.}
Now, queen Dol,\\
Have you pack'd up all?\\
\dolspeaks \vinphantom{Have you pack'd up all?} Yes.\\
\facespeaks \vinphantom{Have you pack'd up all? Yes.} And how do you like\\
The lady Pliant?\\
\dolspeaks \vinphantom{The lady Pliant?} A good dull innocent.\\
\direct*{RE-ENTER SUBTLE.}
\subtlespeaks Here's your Hieronimo's cloak and hat.\\
\facespeaks Give me them.\\
\subtlespeaks \vinphantom{Give me them.} And the ruff too?\\
\facespeaks \vinphantom{Give me them. And the ruff too?} Yes; I'll come to you presently.\\
\direct*{EXIT.}
\subtlespeaks Now he is gone about his project, Dol,\\
I told you of, for the widow.\\
\dolspeaks \vinphantom{I told you of, for the widow.} 'Tis direct\\
Against our articles.\\
\subtlespeaks \vinphantom{Against our articles.} Well, we will fit him, wench.\\
Hast thou gull'd her of her jewels or her bracelets?\\
\dolspeaks No; but I will do't.\\
\subtlespeaks \vinphantom{No; but I will do't.} Soon at night, my Dolly,\\
When we are shipp'd, and all our goods aboard,\\
Eastward for Ratcliff, we will turn our course\\
To Brainford, westward, if thou sayst the word,\\
And take our leaves of this o'er-weening rascal,\\
This peremptory Face.\\
\dolspeaks \vinphantom{This peremptory Face.} Content, I'm weary of him.\\
\subtlespeaks Thou'st cause, when the slave will run a wiving, Dol,\\
Against the instrument that was drawn between us.\\
\dolspeaks I'll pluck his bird as bare as I can.\\
\subtlespeaks \vinphantom{I'll pluck his bird as bare as I can.} Yes, tell her,\\
She must by any means address some present\\
To the cunning man, make him amends for wronging\\
His art with her suspicion; send a ring,\\
Or chain of pearl; she will be tortured else\\
Extremely in her sleep, say, and have strange things\\
Come to her. Wilt thou?\\
\dolspeaks \vinphantom{Come to her. Wilt thou?} Yes.\\
\subtlespeaks \vinphantom{Come to her. Wilt thou? Yes.} My fine flitter-mouse,\\
My bird o' the night! we'll tickle it at the Pigeons,\\
When we have all, and may unlock the trunks,\\
And say, this's mine, and thine; and thine, and mine.\\
\direct*{THEY KISS.}
\direct*{RE-ENTER FACE.}
\facespeaks What now! a billing?\\
\subtlespeaks \vinphantom{What now! a billing?} Yes, a little exalted\\
In the good passage of our stock-affairs.\\
\facespeaks Drugger has brought his parson; take him in, Subtle,\\
And send Nab back again to wash his face.\\
\subtlespeaks I will: and shave himself?\\
\direct*{EXIT.}
\facespeaks \vinphantom{I will: and shave himself?} If you can get him.\\
\dolspeaks You are hot upon it, Face, whate'er it is!\\
\facespeaks A trick that Dol shall spend ten pound a month by.\\
\direct*{RE-ENTER SUBTLE.}
Is he gone?\\
\subtlespeaks \vinphantom{Is he gone?} The chaplain waits you in the hall, sir.\\
\facespeaks I'll go bestow him.\\
\direct*{EXIT.}
\dolspeaks \vinphantom{I'll go bestow him.} He'll now marry her, instantly.\\
\subtlespeaks He cannot yet, he is not ready. Dear Dol,\\
Cozen her of all thou canst. To deceive him\\
Is no deceit, but justice, that would break\\
Such an inextricable tie as ours was.\\
\dolspeaks Let me alone to fit him.\\
\direct*{RE-ENTER FACE.}
\facespeaks \vinphantom{Let me alone to fit him.} Come, my venturers,\\
You have pack'd up all? where be the trunks? bring forth.\\
\subtlespeaks Here.\\
\facespeaks \vinphantom{Here.} Let us see them. Where's the money?\\
\subtlespeaks \vinphantom{Here. Let us see them. Where's the money?} Here,\\
In this.\\
\facespeaks \vinphantom{In this.} Mammon's ten pound; eight score before:\\
The brethren's money, this. Drugger's and Dapper's.\\
What paper's that?\\
\dolspeaks \vinphantom{What paper's that?} The jewel of the waiting maid's,\\
That stole it from her lady, to know certain---\\
\facespeaks If she should have precedence of her mistress?\\
\dolspeaks Yes.\\
\facespeaks \vinphantom{Yes.} What box is that?\\
\subtlespeaks \vinphantom{Yes. What box is that?} The fish-wives' rings, I think,\\
And the ale-wives' single money. Is't not, Dol?\\
\dolspeaks Yes; and the whistle that the sailor's wife\\
Brought you to know an her husband were with Ward.\\
\facespeaks We'll wet it to-morrow; and our silver-beakers\\
And tavern cups. Where be the French petticoats,\\
And girdles and hangers?\\
\subtlespeaks \vinphantom{And girdles and hangers?} Here, in the trunk,\\
And the bolts of lawn.\\
\facespeaks \vinphantom{And the bolts of lawn.} Is Drugger's damask there,\\
And the tobacco?\\
\subtlespeaks \vinphantom{And the tobacco?} Yes.\\
\facespeaks \vinphantom{And the tobacco? Yes.} Give me the keys.\\
\dolspeaks Why you the keys?\\
\subtlespeaks \vinphantom{Why you the keys?} No matter, Dol; because\\
We shall not open them before he comes.\\
\facespeaks 'Tis true, you shall not open them, indeed;\\
Nor have them forth, do you see? Not forth, Dol.\\
\dolspeaks No!\\
\facespeaks \vinphantom{No!} No, my smock rampant. The right is, my master\\
Knows all, has pardon'd me, and he will keep them;\\
Doctor, 'tis true---you look---for all your figures:\\
I sent for him, indeed. Wherefore, good partners,\\
Both he and she be satisfied; for here\\
Determines the indenture tripartite\\
'Twixt Subtle, Dol, and Face. All I can do\\
Is to help you over the wall, o' the back-side,\\
Or lend you a sheet to save your velvet gown, Dol.\\
Here will be officers presently, bethink you\\
Of some course suddenly to 'scape the dock:\\
For thither you will come else.\\
\direct*{LOUD KNOCKING.}
Hark you, thunder.\\
\subtlespeaks \vinphantom{Hark you, thunder.} You are a precious fiend!\\
\officeronespeaks \direct*{WITHOUT} Open the door.\\
\facespeaks Dol, I am sorry for thee i'faith; but hear'st thou?\\
It shall go hard but I will place thee somewhere:\\
Thou shalt have my letter to mistress Amo---\\
\dolspeaks Hang you!\\
\facespeaks \vinphantom{Hang you!} Or madam Caesarean.\\
\dolspeaks \vinphantom{Hang you! Or madam Caesarean.} Pox upon you, rogue,\\
Would I had but time to beat thee!\\
\facespeaks \vinphantom{Would I had but time to beat thee!} Subtle,\\
Let's know where you set up next; I will send you\\
A customer now and then, for old acquaintance:\\
What new course have you?\\
\subtlespeaks \vinphantom{What new course have you?} Rogue, I'll hang myself;\\
That I may walk a greater devil than thou,\\
And haunt thee in the flock-bed and the buttery.\\
\direct*{EXEUNT.}
\scene

\stagedir{AN OUTER ROOM IN THE SAME. ENTER LOVEWIT IN THE SPANISH DRESS, WITH THE PARSON. LOUD KNOCKING AT THE DOOR.}
\lovewitspeaks What do you mean, my masters?\\
\mammonspeaks \direct*{WITHOUT} Open your door,\\
Cheaters, bawds, conjurers.\\
\officeronespeaks \direct*{WITHOUT} Or we will break it open.\\
\lovewitspeaks \vinphantom{Cheaters, bawds, conjurers.} What warrant have you?\\
\officeronespeaks \direct*{WITHOUT} Warrant enough, sir, doubt not,\\
If you'll not open it.\\
\lovewitspeaks \vinphantom{If you'll not open it.} Is there an officer, there?\\
\officeronespeaks \direct*{WITHOUT} Yes, two or three for failing.\\
\lovewitspeaks Have but patience,\\
And I will open it straight.\\
\direct*{ENTER FACE, AS BUTLER.}
\facespeaks \vinphantom{And I will open it straight.} Sir, have you done?\\
Is it a marriage? perfect?\\
\lovewitspeaks \vinphantom{Is it a marriage? perfect?} Yes, my brain.\\
\facespeaks Off with your ruff and cloak then; be yourself, sir.\\
\surlyspeaks \direct*{WITHOUT} Down with the door.\\
\kastrilspeaks \direct*{WITHOUT} 'Slight, ding it open.\\
\lovewitspeaks \direct*{OPENING THE DOOR} Hold,\\
Hold, gentlemen, what means this violence?\\
\direct*{MAMMON, SURLY, KASTRIL, ANANIAS, TRIBULATION, AND OFFICERS, RUSH IN.}
\mammonspeaks Where is this collier?\\
\surlyspeaks \vinphantom{Where is this collier?} And my captain Face?\\
\mammonspeaks These day owls.\\
\surlyspeaks \vinphantom{These day owls.} That are birding in men's purses.\\
\mammonspeaks Madam suppository.\\
\kastrilspeaks \vinphantom{Madam suppository.} Doxy, my suster.\\
\ananiasspeaks Locusts\\
Of the foul pit.\\
\tribulationspeaks \vinphantom{Of the foul pit.} Profane as Bel and the dragon.\\
\ananiasspeaks Worse than the grasshoppers, or the lice of Egypt.\\
\lovewitspeaks Good gentlemen, hear me. Are you officers,\\
And cannot stay this violence?\\
\officeronespeaks \vinphantom{And cannot stay this violence?} Keep the peace.\\
\lovewitspeaks Gentlemen, what is the matter? whom do you seek?\\
\mammonspeaks The chemical cozener.\\
\surlyspeaks \vinphantom{The chemical cozener.} And the captain pander.\\
\kastrilspeaks The nun my suster.\\
\mammonspeaks \vinphantom{The nun my suster.} Madam Rabbi.\\
\ananiasspeaks \vinphantom{The nun my suster. Madam Rabbi.} Scorpions,\\
And caterpillars.\\
\lovewitspeaks \vinphantom{And caterpillars.} Fewer at once, I pray you.\\
\officertwospeaks One after another, gentlemen, I charge you,\\
By virtue of my staff.\\
\ananiasspeaks \vinphantom{By virtue of my staff.} They are the vessels\\
Of pride, lust, and the cart.\\
\lovewitspeaks \vinphantom{Of pride, lust, and the cart.} Good zeal, lie still\\
A little while.\\
\tribulationspeaks \vinphantom{A little while.} Peace, deacon Ananias.\\
\lovewitspeaks The house is mine here, and the doors are open;\\
If there be any such persons as you seek for,\\
Use your authority, search on o' God's name.\\
I am but newly come to town, and finding\\
This tumult 'bout my door, to tell you true,\\
It somewhat mazed me; till my man, here, fearing\\
My more displeasure, told me he had done\\
Somewhat an insolent part, let out my house\\
(Belike, presuming on my known aversion\\
From any air o' the town while there was sickness,)\\
To a doctor and a captain: who, what they are\\
Or where they be, he knows not.\\
\mammonspeaks \vinphantom{Or where they be, he knows not.} Are they gone?\\
\lovewitspeaks You may go in and search, sir.\\
\direct*{MAMMON, ANA., AND TRIB. GO IN.}
Here, I find\\
The empty walls worse than I left them, smoak'd,\\
A few crack'd pots, and glasses, and a furnace:\\
The ceiling fill'd with poesies of the candle,\\
And madam with a dildo writ o' the walls:\\
Only one gentlewoman, I met here,\\
That is within, that said she was a widow---\\
\kastrilspeaks Ay, that's my suster; I'll go thump her. Where is she?\\
\direct*{GOES IN.}
\lovewitspeaks And should have married a Spanish count, but he,\\
When he came to't, neglected her so grossly,\\
That I, a widower, am gone through with her.\\
\surlyspeaks How! have I lost her then?\\
\lovewitspeaks \vinphantom{How! have I lost her then?} Were you the don, sir?\\
Good faith, now, she does blame you extremely, and says\\
You swore, and told her you had taken the pains\\
To dye your beard, and umber o'er your face,\\
Borrowed a suit, and ruff, all for her love;\\
And then did nothing. What an oversight,\\
And want of putting forward, sir, was this!\\
Well fare an old harquebuzier, yet,\\
Could prime his powder, and give fire, and hit,\\
All in a twinkling!\\
\direct*{RE-ENTER MAMMON.}
\mammonspeaks \vinphantom{All in a twinkling!} The whole nest are fled!\\
\lovewitspeaks What sort of birds were they?\\
\mammonspeaks \vinphantom{What sort of birds were they?} A kind of choughs,\\
Or thievish daws, sir, that have pick'd my purse\\
Of eight score and ten pounds within these five weeks,\\
Beside my first materials; and my goods,\\
That lie in the cellar, which I am glad they have left,\\
I may have home yet.\\
\lovewitspeaks \vinphantom{I may have home yet.} Think you so, sir?\\
\mammonspeaks \vinphantom{I may have home yet. Think you so, sir?} Ay.\\
\lovewitspeaks By order of law, sir, but not otherwise.\\
\mammonspeaks Not mine own stuff!\\
\lovewitspeaks \vinphantom{Not mine own stuff!} Sir, I can take no knowledge\\
That they are yours, but by public means.\\
If you can bring certificate that you were gull'd of them,\\
Or any formal writ out of a court,\\
That you did cozen your self, I will not hold them.\\
\mammonspeaks I'll rather lose them.\\
\lovewitspeaks \vinphantom{I'll rather lose them.} That you shall not, sir,\\
By me, in troth: upon these terms, they are yours.\\
What! should they have been, sir, turn'd into gold, all?\\
\mammonspeaks No,\\
I cannot tell---It may be they should.---What then?\\
\lovewitspeaks What a great loss in hope have you sustain'd!\\
\mammonspeaks Not I, the commonwealth has.\\
\facespeaks \vinphantom{Not I, the commonwealth has.} Ay, he would have built\\
The city new; and made a ditch about it\\
Of silver, should have run with cream from Hogsden;\\
That every Sunday, in Moorfields, the younkers,\\
And tits and tom-boys should have fed on, gratis.\\
\mammonspeaks I will go mount a turnip-cart, and preach\\
The end of the world, within these two months. Surly,\\
What! in a dream?\\
\surlyspeaks \vinphantom{What! in a dream?} Must I needs cheat myself,\\
With that same foolish vice of honesty!\\
Come, let us go and hearken out the rogues:\\
That Face I'll mark for mine, if e'er I meet him.\\
\facespeaks If I can hear of him, sir, I'll bring you word,\\
Unto your lodging; for in troth, they were strangers\\
To me, I thought them honest as my self, sir.\\
\direct*{EXEUNT MAM. AND SUR.}
\direct*{RE-ENTER ANANIAS AND TRIBULATION.}
\tribulationspeaks 'Tis well, the saints shall not lose all yet. Go,\\
And get some carts---\\
\lovewitspeaks \vinphantom{And get some carts---} For what, my zealous friends?\\
\ananiasspeaks To bear away the portion of the righteous\\
Out of this den of thieves.\\
\lovewitspeaks \vinphantom{Out of this den of thieves.} What is that portion?\\
\ananiasspeaks The goods sometimes the orphan's, that the brethren\\
Bought with their silver pence.\\
\lovewitspeaks \vinphantom{Bought with their silver pence.} What, those in the cellar,\\
The knight sir Mammon claims?\\
\ananiasspeaks \vinphantom{The knight sir Mammon claims?} I do defy\\
The wicked Mammon, so do all the brethren,\\
Thou profane man! I ask thee with what conscience\\
Thou canst advance that idol against us,\\
That have the seal? were not the shillings number'd,\\
That made the pounds; were not the pounds told out,\\
Upon the second day of the fourth week,\\
In the eighth month, upon the table dormant,\\
The year of the last patience of the saints,\\
Six hundred and ten?\\
\lovewitspeaks \vinphantom{Six hundred and ten?} Mine earnest vehement botcher,\\
And deacon also, I cannot dispute with you:\\
But if you get you not away the sooner,\\
I shall confute you with a cudgel.\\
\ananiasspeaks Sir!\\
\tribulationspeaks \vinphantom{Sir!} Be patient, Ananias.\\
\ananiasspeaks \vinphantom{Sir! Be patient, Ananias.} I am strong,\\
And will stand up, well girt, against an host\\
That threaten Gad in exile.\\
\lovewitspeaks \vinphantom{That threaten Gad in exile.} I shall send you\\
To Amsterdam, to your cellar.\\
\ananiasspeaks \vinphantom{To Amsterdam, to your cellar.} I will pray there,\\
Against thy house: may dogs defile thy walls,\\
And wasps and hornets breed beneath thy roof,\\
This seat of falsehood, and this cave of cozenage!\\
\direct*{EXEUNT ANA. AND TRIB.}
\direct*{ENTER DRUGGER.}
\lovewitspeaks Another too?\\
\druggerspeaks \vinphantom{Another too?} Not I, sir, I am no brother.\\
\lovewitspeaks \direct*{BEATS HIM} Away, you Harry Nicholas! do you talk?\\
\direct*{EXIT DRUG.}
\facespeaks No, this was Abel Drugger. Good sir, go,\\
\direct*{TO THE PARSON.}
And satisfy him; tell him all is done:\\
He staid too long a washing of his face.\\
The doctor, he shall hear of him at West-chester;\\
And of the captain, tell him, at Yarmouth, or\\
Some good port-town else, lying for a wind.\\
\direct*{EXIT PARSON.}
If you can get off the angry child, now, sir---\\
\direct*{ENTER KASTRIL, DRAGGING IN HIS SISTER.}
\kastrilspeaks Come on, you ewe, you have match'd most sweetly,\\
have you not?\\
Did not I say, I would never have you tupp'd\\
But by a dubb'd boy, to make you a lady-tom?\\
'Slight, you are a mammet! O, I could touse you, now.\\
Death, mun' you marry, with a pox!\\
\lovewitspeaks \vinphantom{Death, mun' you marry, with a pox!} You lie, boy;\\
As sound as you; and I'm aforehand with you.\\
\kastrilspeaks Anon!\\
\lovewitspeaks \vinphantom{Anon!} Come, will you quarrel? I will feize you, sirrah;\\
Why do you not buckle to your tools?\\
\kastrilspeaks \vinphantom{Why do you not buckle to your tools?} Od's light,\\
This is a fine old boy as e'er I saw!\\
\lovewitspeaks What, do you change your copy now? proceed;\\
Here stands my dove: stoop at her, if you dare.\\
\kastrilspeaks 'Slight, I must love him! I cannot choose, i'faith,\\
An I should be hang'd for't! Suster, I protest,\\
I honour thee for this match.\\
\lovewitspeaks \vinphantom{I honour thee for this match.} O, do you so, sir?\\
\kastrilspeaks Yes, an thou canst take tobacco and drink, old boy,\\
I'll give her five hundred pound more to her marriage,\\
Than her own state.\\
\lovewitspeaks \vinphantom{Than her own state.} Fill a pipe full, Jeremy.\\
\facespeaks Yes; but go in and take it, sir.\\
\lovewitspeaks \vinphantom{Yes; but go in and take it, sir.} We will---\\
I will be ruled by thee in any thing, Jeremy.\\
\kastrilspeaks 'Slight, thou art not hide-bound, thou art a jovy boy!\\
Come, let us in, I pray thee, and take our whiffs.\\
\lovewitspeaks Whiff in with your sister, brother boy.\\
\direct*{EXEUNT KAS. AND DAME P.}
That master\\
That had received such happiness by a servant,\\
In such a widow, and with so much wealth,\\
Were very ungrateful, if he would not be\\
A little indulgent to that servant's wit,\\
And help his fortune, though with some small strain\\
Of his own candour.\\
\direct*{ADVANCING.}
---"Therefore, gentlemen,\\
And kind spectators, if I have outstript\\
An old man's gravity, or strict canon, think\\
What a young wife and a good brain may do;\\
Stretch age's truth sometimes, and crack it too.\\
Speak for thy self, knave."\\
\facespeaks \vinphantom{Speak for thy self, knave."} "So I will, sir."\\
\direct*{ADVANCING TO THE FRONT OF THE STAGE.}
"Gentlemen,\\
My part a little fell in this last scene,\\
Yet 'twas decorum. And though I am clean\\
Got off from Subtle, Surly, Mammon, Dol,\\
Hot Ananias, Dapper, Drugger, all\\
With whom I traded: yet I put my self\\
On you, that are my country: and this pelf\\
Which I have got, if you do quit me, rests\\
To feast you often, and invite new guests."\\
\direct*{EXEUNT.}

\end{drama*}
\end{document}
